### 2010-02-18 StatusReport.tex
Added status report!

% StatusReport.tex
% Chrisna Aing and Sarah Halls

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage[hmargin=2cm, vmargin=2cm]{geometry}
\begin{document}

    Since both our standard model and our CAR model have been working relatively
    well on our simulated data, we moved onto running them on the actual data.

    The standard model ran without convergence issues on the 2003 data for all
    rivers, but it did not converge on any of the 2004 data.  We simplified the
    model by eliminating E in an attempt to improve convergence, which worked
    for the 402m Whitewater River and 402m Zumbro River data.  Although none of
    the Mississippi River's 2004 data converged, we are still including
    estimates for the 402m Mississippi River data.

    The CAR model did not converge for any of the data.  Although the standard
    model and the CAR model performed similarly in testing when both models
    converged, the standard model converged much more often than the CAR model
    did.  We think this is because the CAR model attempts to estimate too many
    parameters.  There is a different estimate of \(\theta\) for every site,
    which adds a considerable amount of complexity to the model.  We also tried
    to simplify this model by eliminating E, but this did not improve
    convergence.  Therefore, we are not going to include any estimates of
    parameters from the CAR model in this report.

    We decided not to run the models on the alternating plots. In our earlier
    analysis, we discovered that the main difference between alternating plots
    and all plots was that the alternating plots were more independent. When
    testing the standard model on spatially dependent simulated data, we found
    that the estimates were fairly accurate. The CAR model estimates were also
    fairly accurate on this simulated data, when it converged. Therefore, we
    decided to use the data that includes all plots because we feel that having
    the additional information from all plots is more important to the model
    than independence is.

    All of our estimates are below and are accompanied with some thoughts about
    their convergence.  We are wondering if these estimates seem reasonable,
    based upon what you know about the three rivers. \\

    % Mississippi River!
    \begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
        \hline
            & 402m & 804m & 1608m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.01954 & 0.04808 & 0.06885 \\
            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.05216 & 0.07771 & 0.13650 \\
            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.02668 & 0.06140 & 0.10230 \\
            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.65240 & 0.67340 & 0.74960 \\
            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.67770 & 0.77580 & 0.82540 \\
            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.53670 & 0.54220 & 0.65330 \\
            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.59380 & 0.69370 & 0.75820 \\
            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.15280 & 0.20730 & 0.27690 \\
            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
        \hline
    \end{tabular}
    \\

    This data converged well, for the most part.  At the 1608m plot size, E,
    \(\psi\), and \(\theta\) did not converge well. \\

    \begin{tabular}{|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Mississippi River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.3129 \\
            & (0.2174, 0.8107) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.7613 \\
            & (0.6609, 0.8833) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.8511 \\
            & (0.2426, 0.9973) \\
        \hline
    \end{tabular}
    \\

    This data did not converge.  We do not think that these estimates are
    accurate, but they are included for the sake of completeness. \\

    % Whitewater River!
    \begin{tabular}{|l|l|l|}
        \hline
        \multicolumn{3}{|l|}{\textbf{Whitewater River, 2003}} \\
        \hline
            & 402m & 804m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.05608 & 0.0555 \\
            & (0.02474, 0.09554) & (0.002859, 0.1390) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.16570 & 0.1931 \\
            & (0.10130, 0.23510) & (0.065910, 0.3384) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.15350 & 0.2241 \\
            & (0.09758, 0.21660) & (0.110000, 0.3623) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.66990 & 0.6985 \\
            & (0.56080, 0.79040) & (0.573100, 0.8366) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.92040 & 0.9877 \\
            & (0.84270, 0.97590) & (0.945100, 1.0000) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.75910 & 0.8575 \\
            & (0.65060, 0.85800) & (0.745400, 0.9600) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.87590 & 0.9126 \\
            & (0.72910, 0.98880) & (0.746900, 0.9955) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.24100 & 0.3779 \\
            & (0.17890, 0.30900) & (0.286500, 0.4721) \\
        \hline
    \end{tabular}
    \\

    While the 402m data converged well, E, \(\psi\), and \(\theta\) did not
    converge well for the 804m data. \\

    \begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Whitewater River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.8237 \\
            & (0.5323, 0.9804) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.6066 \\
            & (0.4246, 0.8382) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.2068 \\
            & (0.1211, 0.3550) \\
        \hline
    \end{tabular}
    \\

    This data converged well.  However, we feel that the absence of E from the
    model makes the estimates much less dependable. \\

    % Zumbro River!
    \begin{tabular}{|l|l|l|}
        \hline
        \multicolumn{3}{|l|}{\textbf{Zumbro River, 2003}} \\
        \hline
            & 402m & 804m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.02922 & 0.007066 \\
            & (0.006786, 0.05994) & (0.0000, 0.03403) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.03388 & 0.008312 \\
            & (0.008584, 0.06880) & (0.0000, 0.04188) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.02228 & 0.012960 \\
            & (0.002430, 0.05385) & (0.0000, 0.05642) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.81890 & 0.837100 \\
            & (0.749500, 0.88540) & (0.7615, 0.90700) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.81770 & 0.852500 \\
            & (0.740100, 0.88670) & (0.7597, 0.92230) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.73120 & 0.805400 \\
            & (0.625400, 0.82440) & (0.6885, 0.89780) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.66550 & 0.651600 \\
            & (0.549100, 0.78150) & (0.5098, 0.78210) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.29950 & 0.395800 \\
            & (0.236400, 0.36550) & (0.3100, 0.48520) \\
        \hline
    \end{tabular}
    \\

    Again, the 402m data converged well.  With respect to the 804m data, while
    \(\psi\) and \(\theta\) converged really well, none of the other parameters
    did. \\

    \begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{2}{|l|}{\textbf{Zumbro River, 2004}} \\
        \hline
            & 402m \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.9166 \\
            & (0.7739, 0.9918) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.4773 \\
            & (0.3586, 0.6114) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.4232 \\
            & (0.3073, 0.5520) \\
        \hline
    \end{tabular}
    \\

    This data converged well.  However, as is the case with all of the 2004
    data, we feel that the absence of E from the model makes the estimates much
    less dependable. \\

\end{document}

### 2010-02-18 StatusReport.tex
Added more detail about CAR model without E.

@@
     which adds a considerable amount of complexity to the model.  We also tried
-    to simplify this model by eliminating E, but this did not improve
-    convergence.  Therefore, we are not going to include any estimates of
+    to simplify this model by eliminating E because this was somewhat successful
+    with the standard model.  This improved convergence with the 402m data, but
+    gave us \(\psi\) estimates greater than 0.9 for all rivers, which does not
+    seem reasonable.  Therefore, we are not going to include any estimates of
     parameters from the CAR model in this report.

### 2010-02-18 StatusReport.tex
Added all of Bob's changes.

@@
 \usepackage{multirow}
-\usepackage[hmargin=2cm, vmargin=2cm]{geometry}
+\usepackage[hmargin=3cm, vmargin=3cm]{geometry}
 \begin{document}
+
+    \noindent{\textbf{Status Report for Otter Estimation Project \hfill Feb. 18,
+    2010}
+
+    \noindent{From: Chrisna Aing, Sarah Halls, Kiva Oken, Carleton College
+    Senior Comps group} \\
+
+    Recall that our "standard model," contains the following parameters:
+    \(\psi\) (occupancy rate), \(\theta\) (probability of laying down a track,
+    given occupancy), \(p\) (the detection probability for each observer; so \(1
+    - p\) is the false negative rate), and \(E\) (the false positive rate for
+    each observer).  We have also worked with a CAR model using the WinBUGS CAR
+    installation.
 
@@
     rivers, but it did not converge on any of the 2004 data.  We simplified the
-    model by eliminating E in an attempt to improve convergence, which worked
-    for the 402m Whitewater River and 402m Zumbro River data.  Although none of
-    the Mississippi River's 2004 data converged, we are still including
-    estimates for the 402m Mississippi River data.
+    model by eliminating E (assuming no false positives) in an attempt to
+    improve convergence, which worked for the 402m Whitewater River and 402m 
+    Zumbro River data.  Although none of the Mississippi River's 2004 data 
+    converged, we are still including estimates for the 402m Mississippi River 
+    data.
 
@@
     analysis, we discovered that the main difference between alternating plots
-    and all plots was that the alternating plots were more independent. When
-    testing the standard model on spatially dependent simulated data, we found
-    that the estimates were fairly accurate. The CAR model estimates were also
-    fairly accurate on this simulated data, when it converged. Therefore, we
-    decided to use the data that includes all plots because we feel that having
-    the additional information from all plots is more important to the model
-    than independence is.
+    and all plots was that the alternating plots showed more independence in the
+    data.  When testing the standard model on spatially dependent simulated
+    data, we found that the estimates were fairly accurate.  The CAR model
+    estimates were also fairly accurate on this simulated data, when it
+    converged.  Therefore, we decided to use the data that includes all plots
+    because we feel that having the additional information from all plots is
+    more important to the model than independence is.
 

### 2010-02-21 kivaSections.tex

kkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkk

### 2010-02-21 ChrisnaSections.tex
Screw all of you.

--- 2010-02-18 StatusReport.tex
@@
-% StatusReport.tex
-% Chrisna Aing and Sarah Halls
-
-\documentclass[12pt]{article}
+\documentclass{article}
 \usepackage{amsmath}
-\usepackage{multirow}
-\usepackage[hmargin=2cm, vmargin=2cm]{geometry}
 \begin{document}
-
-    Since both our standard model and our CAR model have been working relatively
-    well on our simulated data, we moved onto running them on the actual data.
-
-    The standard model ran without convergence issues on the 2003 data for all
-    rivers, but it did not converge on any of the 2004 data.  We simplified the
-    model by eliminating E in an attempt to improve convergence, which worked
-    for the 402m Whitewater River and 402m Zumbro River data.  Although none of
-    the Mississippi River's 2004 data converged, we are still including
-    estimates for the 402m Mississippi River data.
-
-    The CAR model did not converge for any of the data.  Although the standard
-    model and the CAR model performed similarly in testing when both models
-    converged, the standard model converged much more often than the CAR model
-    did.  We think this is because the CAR model attempts to estimate too many
-    parameters.  There is a different estimate of \(\theta\) for every site,
-    which adds a considerable amount of complexity to the model.  We also tried
-    to simplify this model by eliminating E, but this did not improve
-    convergence.  Therefore, we are not going to include any estimates of
-    parameters from the CAR model in this report.
-
-    We decided not to run the models on the alternating plots. In our earlier
-    analysis, we discovered that the main difference between alternating plots
-    and all plots was that the alternating plots were more independent. When
-    testing the standard model on spatially dependent simulated data, we found
-    that the estimates were fairly accurate. The CAR model estimates were also
-    fairly accurate on this simulated data, when it converged. Therefore, we
-    decided to use the data that includes all plots because we feel that having
-    the additional information from all plots is more important to the model
-    than independence is.
-
-    All of our estimates are below and are accompanied with some thoughts about
-    their convergence.  We are wondering if these estimates seem reasonable,
-    based upon what you know about the three rivers. \\
-
-    % Mississippi River!
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
-        \hline
-            & 402m & 804m & 1608m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.01954 & 0.04808 & 0.06885 \\
-            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.05216 & 0.07771 & 0.13650 \\
-            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.02668 & 0.06140 & 0.10230 \\
-            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.65240 & 0.67340 & 0.74960 \\
-            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.67770 & 0.77580 & 0.82540 \\
-            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.53670 & 0.54220 & 0.65330 \\
-            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.59380 & 0.69370 & 0.75820 \\
-            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.15280 & 0.20730 & 0.27690 \\
-            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
-        \hline
-    \end{tabular}
-    \\
-
-    This data converged well, for the most part.  At the 1608m plot size, E,
-    \(\psi\), and \(\theta\) did not converge well. \\
-
-    \begin{tabular}{|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Mississippi River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.3129 \\
-            & (0.2174, 0.8107) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.7613 \\
-            & (0.6609, 0.8833) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.8511 \\
-            & (0.2426, 0.9973) \\
-        \hline
-    \end{tabular}
-    \\
-
-    This data did not converge.  We do not think that these estimates are
-    accurate, but they are included for the sake of completeness. \\
-
-    % Whitewater River!
-    \begin{tabular}{|l|l|l|}
-        \hline
-        \multicolumn{3}{|l|}{\textbf{Whitewater River, 2003}} \\
-        \hline
-            & 402m & 804m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.05608 & 0.0555 \\
-            & (0.02474, 0.09554) & (0.002859, 0.1390) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.16570 & 0.1931 \\
-            & (0.10130, 0.23510) & (0.065910, 0.3384) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.15350 & 0.2241 \\
-            & (0.09758, 0.21660) & (0.110000, 0.3623) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.66990 & 0.6985 \\
-            & (0.56080, 0.79040) & (0.573100, 0.8366) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.92040 & 0.9877 \\
-            & (0.84270, 0.97590) & (0.945100, 1.0000) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.75910 & 0.8575 \\
-            & (0.65060, 0.85800) & (0.745400, 0.9600) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.87590 & 0.9126 \\
-            & (0.72910, 0.98880) & (0.746900, 0.9955) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.24100 & 0.3779 \\
-            & (0.17890, 0.30900) & (0.286500, 0.4721) \\
-        \hline
-    \end{tabular}
-    \\
-
-    While the 402m data converged well, E, \(\psi\), and \(\theta\) did not
-    converge well for the 804m data. \\
-
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Whitewater River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.8237 \\
-            & (0.5323, 0.9804) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.6066 \\
-            & (0.4246, 0.8382) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.2068 \\
-            & (0.1211, 0.3550) \\
-        \hline
-    \end{tabular}
-    \\
-
-    This data converged well.  However, we feel that the absence of E from the
-    model makes the estimates much less dependable. \\
-
-    % Zumbro River!
-    \begin{tabular}{|l|l|l|}
-        \hline
-        \multicolumn{3}{|l|}{\textbf{Zumbro River, 2003}} \\
-        \hline
-            & 402m & 804m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.02922 & 0.007066 \\
-            & (0.006786, 0.05994) & (0.0000, 0.03403) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.03388 & 0.008312 \\
-            & (0.008584, 0.06880) & (0.0000, 0.04188) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.02228 & 0.012960 \\
-            & (0.002430, 0.05385) & (0.0000, 0.05642) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.81890 & 0.837100 \\
-            & (0.749500, 0.88540) & (0.7615, 0.90700) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.81770 & 0.852500 \\
-            & (0.740100, 0.88670) & (0.7597, 0.92230) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.73120 & 0.805400 \\
-            & (0.625400, 0.82440) & (0.6885, 0.89780) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.66550 & 0.651600 \\
-            & (0.549100, 0.78150) & (0.5098, 0.78210) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.29950 & 0.395800 \\
-            & (0.236400, 0.36550) & (0.3100, 0.48520) \\
-        \hline
-    \end{tabular}
-    \\
-
-    Again, the 402m data converged well.  With respect to the 804m data, while
-    \(\psi\) and \(\theta\) converged really well, none of the other parameters
-    did. \\
-
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Zumbro River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.9166 \\
-            & (0.7739, 0.9918) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.4773 \\
-            & (0.3586, 0.6114) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.4232 \\
-            & (0.3073, 0.5520) \\
-        \hline
-    \end{tabular}
-    \\
-
-    This data converged well.  However, as is the case with all of the 2004
-    data, we feel that the absence of E from the model makes the estimates much
-    less dependable. \\
-
 \end{document}

### 2010-02-21 sarahDraft.tex

\documentclass{article}
\usepackage{amsmath}
\begin{document}

\section{Overview of Occupancy Modeling}
	\subsection{What is Occupancy Rate?}
		"The probability that a randomly selected site or sampling unit in an area 
			of interest is occupied by a species" [MacKenzie2006] (pg. 11)
	\subsection{Different Ways of Sampling Data}
		Measuring rate of occupancy vs. population \\
		Looking at tracks vs. listening to stuff or acutally looking for the animal
	\subsection{Assumptions}
		Independence of detection between sites \\
		No false positives \\
		Presence and detection probabilities constant across sites and surveys \\
		Closed population
	\subsection{Types of Occupancy Models}
		Maximum Likelihood \\
		any other models? \\
		Hierarchical Bayesian?  is that the name of what we used?

\section{Actual data}
Stuff by Kiva.
	\subsection{Testing the Independence Assumption}
		We tried to see if the assumption of independence is viable and whether or
			not there were differences between data with different plot sizes and 
			differences between data containing plots and data containing every 
			other plot.
	\begin{enumerate}
    	\item Using Markov Chains to find r. \\
          * r = potential long-term probability of detection, based upon the 
			short-term conditional probabilities p and q acquired from our data 
			(infinite limit as n approaches infinity of Pn) \\
          * p = probability that a 0 follows a 0, q = probability that a 1 follows
			 a 1  \\
          * If 0->0 and 1->0 (the values of a column, basically) equaled each 
			other, plots would have been independent?  \\
          * Ultimately not useful to us because it does not take into account 
			whether or not a site is occupied, so it doesn't apply to the 
			hierarchical model that we are using. \\
   		\item Using s, where s = (\# of 0-1's) - (\# of 1-1's) for a given flight.\\
          * Decided to use the z-score of s as a measure of independence.  \\
          * Therefore, needed the expected value and the standard deviation of s. \\
          * So, we needed the expected value of each term and the variance, which 
			required finding the variance of each term and the covariance between 
			the two terms. \\
          * Expected value is straightforward and we have all of the equations. \\
          * Variance is fun and we also have all of the relevant equations. \\
          * Calculated z-score for every flight and box-plotted them for each year
			and plot size and pitted data with every other plot against data with
			all plots. \\
          * Nothing important with respect to plot size. Alternating plots show 
			more independence than all plots. Look at report for more. \\
          * CAR model started working once we took s values of simulated data into
			account by increasing dependence.  \\
	\end{enumerate}

\end{document}

### 2010-02-21 sarahDraft.tex
Put what I missed from the outline in my draft

@@
 
+\section{Theoretical Stuff} 
+	\subsection{Bayesian Statistics}
+	   	Bayes' Theorem \\
+    	Useful for hierarchical modeling: look at section 3.3 in book. \\
+    	Can use Markov Chain Monte Carlo (MCMC) to calculate posterior 
+			distribution- bimodal issues.
+	\subsection{BRugs/R}
+    	All of our statistical analysis was performed in R.  \\
+    	All of our modeling was performed using BRugs, which allowed us to use 
+			WinBUGS commands and stuff in R.
+
 \end{document}

### 2010-02-22 ChrisnaSections.tex
I think this is the beginning of a beautiful friendship.

--- 2010-02-18 StatusReport.tex
@@
-% StatusReport.tex
-% Chrisna Aing and Sarah Halls
-
-\documentclass[12pt]{article}
+\documentclass{article}
 \usepackage{amsmath}
 \usepackage{multirow}
-\usepackage[hmargin=2cm, vmargin=2cm]{geometry}
+\usepackage[dvipsnames, table]{xcolor}
+
 \begin{document}
 
-    Since both our standard model and our CAR model have been working relatively
-    well on our simulated data, we moved onto running them on the actual data.
+%%% *STANDARD HIERARCHICAL MODEL*:
+\section{Standard Hierarchical Model}
 
-    The standard model ran without convergence issues on the 2003 data for all
-    rivers, but it did not converge on any of the 2004 data.  We simplified the
-    model by eliminating E in an attempt to improve convergence, which worked
-    for the 402m Whitewater River and 402m Zumbro River data.  Although none of
-    the Mississippi River's 2004 data converged, we are still including
-    estimates for the 402m Mississippi River data.
+    %% MODEL PARAMETERS:
+    \subsection{Model Parameters}
 
-    The CAR model did not converge for any of the data.  Although the standard
-    model and the CAR model performed similarly in testing when both models
-    converged, the standard model converged much more often than the CAR model
-    did.  We think this is because the CAR model attempts to estimate too many
-    parameters.  There is a different estimate of \(\theta\) for every site,
-    which adds a considerable amount of complexity to the model.  We also tried
-    to simplify this model by eliminating E because this was somewhat successful
-    with the standard model.  This improved convergence with the 402m data, but
-    gave us \(\psi\) estimates greater than 0.9 for all rivers, which does not
-    seem reasonable.  Therefore, we are not going to include any estimates of
-    parameters from the CAR model in this report.
+    %% SIMULATED DATA, PT. 1:
+    \subsection{Simulated Data}
 
-    We decided not to run the models on the alternating plots. In our earlier
-    analysis, we discovered that the main difference between alternating plots
-    and all plots was that the alternating plots were more independent. When
-    testing the standard model on spatially dependent simulated data, we found
-    that the estimates were fairly accurate. The CAR model estimates were also
-    fairly accurate on this simulated data, when it converged. Therefore, we
-    decided to use the data that includes all plots because we feel that having
-    the additional information from all plots is more important to the model
-    than independence is.
+    \noindent{\begin{tabular}{|l|l|l|l|}
+        \hline
+        \multicolumn{4}{|l|}{\textbf{The Model vs. Simulated Data}} \\
+        \hline
+            \(\psi\) & \(\theta\) & Convergence & 95\% CI (\(\psi\)) \\
+        \hline
+        \multirow{5}{*}{0.20}
+            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.112, 0.305) \\
+            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.142, 0.341) \\
+            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.098, 0.284) \\
+            & \cellcolor{Green}0.80 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.097, 0.281) \\
+            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.117, 0.302) \\
+        \hline
+        \multirow{5}{*}{0.40}
+            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.388, 0.721) \\
+            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.270, 0.494) \\
+            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.287, 0.512) \\
+            & \cellcolor{Red}0.80 & \cellcolor{Red}Bad &
+              \cellcolor{Red}(0.277, 0.507) \\
+            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.279, 0.503) \\
+        \hline
+        \multirow{5}{*}{0.60}
+            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
+              \cellcolor{Yellow}(0.312, 0.581) \\
+            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.400, 0.630) \\
+            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.462, 0.687) \\
+            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.502, 0.727) \\
+            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
+              \cellcolor{Red}(0.318, 0.549) \\
+        \hline
+        \multirow{5}{*}{0.80}
+            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.578, 0.838) \\
+            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.736, 0.915) \\
+            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.587, 0.829) \\
+            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.599, 0.811) \\
+            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
+              \cellcolor{Red}(0.612, 0.830) \\
+        \hline
+        \multirow{5}{*}{1.00}
+            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
+              \cellcolor{Yellow}(0.895, 0.999) \\
+            & \cellcolor{Yellow}0.40 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.927, 0.999) \\
+            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.949, 1.000) \\
+            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Good &
+              \cellcolor{Yellow}(0.802, 0.954) \\
+            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
+              \cellcolor{Red}(0.800, 0.955) \\
+        \hline
+    \end{tabular}}
 
-    All of our estimates are below and are accompanied with some thoughts about
-    their convergence.  We are wondering if these estimates seem reasonable,
-    based upon what you know about the three rivers. \\
+%%% *DISCUSSION*:
+\section{Discussion}
+
+    %% GETTING THE ACTUAL DATA TO CONVERGE:
+    \subsection{Obtaining Convergence on the Actual Data}
+    Recall that our standard hierarchical model contains the following
+    parameters: \(\psi\) (the rate of occupancy), \(\theta\) (the probability
+    of laying down a track given occupancy), \(p\) (the probability of
+    detecting a track given that it is present, which is unique to each
+    observer and makes \(1-p\) the false negative rate of detecting a track),
+    and \(E\) (the false positive rate of detecting a track, which is unique to
+    each observer).
+
+    Once both the standard hierarchical model and the addition of a CAR prior
+    distribution on \(\theta\) to the standard model exhibited convergence on
+    our simulated data, we moved onto running both models on the data given to
+    us by the Minnesota DNR.  Recall that data was gathered during the winters
+    of 2003 and 2004 along the Mississippi, Whitewater, and Zumbro Rivers and
+    aggregated with ArcGIS into vectors containing ones and zeroes representing
+    the detection and non-detection of a track, respectively, within plots of
+    length 402 meters, 804 meters, and 1608 meters.  The covariates recorded
+    were: the observer, the date, and the number of days since the last snow
+    event.  In general, the sparseness of the data gathered in 2004 caused us
+    to simplify the model in order to obtain results.  Unfortunately, the
+    simplification results in estimates of \(\psi\) that do not seem accurate
+    and are not directly comparable to estimates of \(\psi\) on 2003's data.
+
+    The standard model converged reliably on all data gathered in 2003, but it
+    did not converge on any of the data gathered in 2004.  Considering that the
+    amount of data gathered in 2004 pales in comparison to the amount of data
+    gathered in 2003, this result was not surprising.  The vast majority of
+    2004's data consists of just one observer taking observations once after
+    each snow event, which causes \(p\) and \(E\) in particular to not converge,
+    regardless of plot size.
+
+    To obtain convergence on 2004's relatively sparse data, we removed \(E\),
+    which adds to the model the assumption that there are no false positives
+    and, consequently, simplifies it greatly.  Although the model still did not
+    converge on data with plot sizes of 804m and 1608m, both the Whitewater and
+    Zumbro Rivers' data at the 402m plot size converged.  Although none of the
+    Mississippi River's 2004 data converged, we still include estimates of
+    \(p\), \(\psi\), and \(\theta\) at the 402m plot size to accompany the
+    estimates corresponding to the Whitewater and Zumbro Rivers.  For the most
+    part, however, we feel that the absence of \(E\) from the model makes the
+    the estimates of \(\psi\) much less dependable.
+
+    The addition of a CAR prior distribution on \(\theta\) to the standard model
+    caused the model to fail to converge on all of the data.  Although the
+    standard model and the CAR model performed similarly on simulated data
+    whenever both models converged, the standard model converged much more often
+    than did the CAR model.  This is most likely due to the fact that the CAR
+    model attempts to estimate too many parameters.  Whereas the standard model
+    estimates between eight and twelve parameters on the actual data (\(\psi\),
+    \(\theta\), the four parameters of the beta distributions corresponding to
+    \(p\) and \(E\), \(p\) for each observer, and \(E\) for each observer), the
+    CAR model estimates \(\theta\) for each site.  Because there are at least
+    thirty-five sites in each data set (regardless of plot size), the CAR prior
+    distribution adds a considerable amount of complexity to the model.
+
+    We also tried to simplify this model by eliminating \(E\) because this was
+    somewhat successful with respect to convergence with the standard model.
+    We first attempted to fit this model to data from 2003 at the 402m plot
+    size.  Although the model successfully converged on all of this data, its
+    estimates of \(\psi\), all of which were greater than 0.9, did not seem
+    reasonable, especially in comparison to the standard model's estimates when
+    given the same data.  As a result, we did not attempt to fit this model to
+    the rest of the data and do not include this model's estimates within our
+    results.
+
+    %% RESULTS:
+    \subsection{Results}
+    Below, in tables, lie our estimates for all of the Minnesota DNR's data from
+    2003 and 2004.  Below each table is detailed commentary about the ability of
+    the standard hierarchical model to convergence to the corresponding data. \\
 
     % Mississippi River!
-    \begin{tabular}{|l|l|l|l|}
+    \noindent{\begin{tabular}{|l|l|l|l|}
         \hline
@@
         \hline
-    \end{tabular}
+    \end{tabular}}
     \\
 
-    This data converged well, for the most part.  At the 1608m plot size, E,
-    \(\psi\), and \(\theta\) did not converge well. \\
+    This data converged well at the 402m and 804m plot sizes.  At the 1608m plot
+    size, \(E\), \(\psi\), and \(\theta\) did not converge well. \\
 
-    \begin{tabular}{|l|l|}
+    \noindent{\begin{tabular}{|l|l|}
         \hline
@@
         \hline
-    \end{tabular}
+    \end{tabular}}
     \\
@@
     % Whitewater River!
-    \begin{tabular}{|l|l|l|}
+    \noindent{\begin{tabular}{|l|l|l|}
         \hline
@@
         \hline
-    \end{tabular}
+    \end{tabular}}
     \\
 
-    While the 402m data converged well, E, \(\psi\), and \(\theta\) did not
+    While the 402m data converged well, \(E\), \(\psi\), and \(\theta\) did not
     converge well for the 804m data. \\
 
-    \begin{tabular}{|l|l|l|l|}
+    \noindent{\begin{tabular}{|l|l|l|l|}
         \hline
@@
         \hline
-    \end{tabular}
+    \end{tabular}}
     \\
 
-    This data converged well.  However, we feel that the absence of E from the
-    model makes the estimates much less dependable. \\
+    This data converged well.  However, we feel that the absence of \(E\) from
+    the model makes the estimates much less dependable.  The estimates are also
+    much less precise than those from 2003 -- the confidence intervals of
+    \(\psi\) and of \(p[Danny]\) are twice as large as those from 2003. \\
 
     % Zumbro River!
-    \begin{tabular}{|l|l|l|}
+    \noindent{\begin{tabular}{|l|l|l|}
         \hline
@@
     Again, the 402m data converged well.  With respect to the 804m data, while
-    \(\psi\) and \(\theta\) converged really well, none of the other parameters
-    did. \\
+    \(\psi\) and \(\theta\) converged well, none of the other parameters did. \\
 
-    \begin{tabular}{|l|l|l|l|}
+    \noindent{\begin{tabular}{|l|l|l|l|}
         \hline
@@
     This data converged well.  However, as is the case with all of the 2004
-    data, we feel that the absence of E from the model makes the estimates much
-    less dependable. \\
+    data, we feel that the absence of \(E\) from the model makes the estimates
+    much less dependable.
+
+    %%
+    \subsection{The Effect of Plot Size}
+    %%
+    \subsection{All Plots vs. Every Other Plot}
+    %%
+    \subsection{How Important Are Our Covariates?}
+    %%
+    \subsection{Best Practices for Measuring Otter Occupancy}
 

### 2010-02-22 sarahDraft.tex

@@
 Stuff by Kiva.
-	\subsection{Testing the Independence Assumption}
-		We tried to see if the assumption of independence is viable and whether or
+	\subsection{Markov Chains and Testing the Independence Assumption}
+		First we considered using Markov Chains to come up with a long term 
+probability of detection, which we called \(r\).  We let \(p\) be the number of 
+\(0\)'s followed by a \(1\) as a fraction of all successive pairs that start with 
+a \(0\).  We let \(q\) be the number of \(1\)'s followed by a \(1\) as a fraction 
+of all successive pairs that start with a \(1\).
+
+		We wanted to find out if the assumption of independence across sites held
+true for our data from the DNR.  We also wanted to see what sampling techniques 
+led to the greatest independence between plots.  The different techniques that we
+compared were using different plot sizes, and using data that includes all plots
+or data that uses only every other plot. \\
+		(We tried to see if the assumption of independence is viable and whether or
 			not there were differences between data with different plot sizes and 
 			differences between data containing plots and data containing every 
-			other plot.
+			other plot.) \\
+		If the observations at different sites are independent, then the 
+probability that there is a \(1\) at site \(i\)  is independent of site \(i-1\).  
+We let \(p\) be the number of \(0\)'s followed by a \(1\) as a fraction of all 
+successive pairs that start with a \(0\).  We let \(1-q\) be the number of \(1\)'s
+followed by a \(1\) as a fraction of all successive pairs that start with a \(1\).
+The difference of these two proportions \((=p-(1-q)\) is a measure of independence,
+where \(p-(1-q)\) being approximately \(0\) under the null hypothesis of 
+independence.  
 	\begin{enumerate}
@@
 
-\section{Theoretical Stuff} 
+\section{Theoretical Stuff} - CHRISNA??
 	\subsection{Bayesian Statistics}

### 2010-02-22 sarahDraft.tex
More writing

@@
 	\subsection{Markov Chains and Testing the Independence Assumption}
+
 		First we considered using Markov Chains to come up with a long term 
 probability of detection, which we called \(r\).  We let \(p\) be the number of 
-\(0\)'s followed by a \(1\) as a fraction of all successive pairs that start with 
+\(0\)'s followed by a \(0\) as a fraction of all successive pairs that start with 
 a \(0\).  We let \(q\) be the number of \(1\)'s followed by a \(1\) as a fraction 
-of all successive pairs that start with a \(1\).
+of all successive pairs that start with a \(1\).  These probabilities can be
+represented by the matrix \(M = \begin{bmatrix} p & 1-p \\ 1-q & q \end{bmatrix}\).
+The long term probability of detection can be found from this by finding 
+\(\lim_{n \to \inf} M^n\), which returns \( \begin{bmatrix} 1-r & r \\ 1-r & r 
+\end{bmatrix}\), where \(r\) is the probability of a \(1\), which is the long term 
+probability of detection.
+
+	We calculated \(p\), \(q\), and \(r\) for every flight in our data, using 
+\(n = 500\).  We then realized that \(r\) does not answer the questions we are 
+asking.  It does not take into account whether or not a site is occupied, or 
+whether or not a track was laid down on that site, or any of the ideas behind the
+hierarchical model.  Thesse are the parameters that we eventually want to find 
+out something about, and the value of \(r\) cannot help us get there.  The ideas 
+behind using Markov Chains did end up being useful, however.
+
+	Although we rejected the idea of using \(r\) as a way to analyze the data, we
+realized that \(p\) and \(q\) could still be used to test the viability of the 
+independence assumption.  If there is independece between sites, the difference 
+of the proportions \(p\) and \(1-q\) should be approximately \(0\).  Looking at 
+boxplots of \(p - (1-q)\), we see that data from flights that record alternating 
+sites may be more independent than data from flights that record all sites, which
+suggests a possible benefit to using the alternating sampling method.  The
+boxplots for the Whitewater and Zumbro Rivers looked similar to the Mississippi:
+*INSERT BOXPLOT FROM STATUS REPORT HERE
+We found no obvious trend indicating one site size to be more independent than 
+the others.
+
+	While this is an interesting observation, we would like to be able to 
+quantify it better by finding the distribution of our statistic.
 
@@
 or data that uses only every other plot. \\
-		(We tried to see if the assumption of independence is viable and whether or
-			not there were differences between data with different plot sizes and 
-			differences between data containing plots and data containing every 
-			other plot.) \\
+
 		If the observations at different sites are independent, then the 

### 2010-02-23 sarahDraft.tex
More writing...

@@
 \usepackage{amsmath}
+\usepackage{fullpage}
 \begin{document}
@@
 of all successive pairs that start with a \(1\).  These probabilities can be
-represented by the matrix \(M = \begin{bmatrix} p & 1-p \\ 1-q & q \end{bmatrix}\).
-The long term probability of detection can be found from this by finding 
+represented by the matrix \(M = \begin{bmatrix} p & 1-p \\ 1-q & q 
+\end{bmatrix}\). The long term probability of detection can be found from this by finding 
 \(\lim_{n \to \inf} M^n\), which returns \( \begin{bmatrix} 1-r & r \\ 1-r & r 
-\end{bmatrix}\), where \(r\) is the probability of a \(1\), which is the long term 
-probability of detection.
+\end{bmatrix}\), where \(r\) is the probability of a \(1\), which is the long 
+term probability of detection.
 
@@
 independence assumption.  If there is independece between sites, the difference 
-of the proportions \(p\) and \(1-q\) should be approximately \(0\).  Looking at 
-boxplots of \(p - (1-q)\), we see that data from flights that record alternating 
-sites may be more independent than data from flights that record all sites, which
-suggests a possible benefit to using the alternating sampling method.  The
-boxplots for the Whitewater and Zumbro Rivers looked similar to the Mississippi:
-*INSERT BOXPLOT FROM STATUS REPORT HERE
+of the proportions \(p\) and \(1-q\) should be approximately \(0\).  This would 
+mean that the probability that a \(0\) follows a \(0\) is about the same as the 
+probability that a \(1\) follows a \(0\), meaning that it does not depend on what
+the observation is at the previous site.  Looking at boxplots of \(p - (1-q)\), 
+we see that data from flights that record alternating sites may be more 
+independent than data from flights that record all sites, which suggests a 
+possible benefit to using the alternating sampling method.  The boxplots for the 
+Whitewater and Zumbro Rivers looked similar to the Mississippi:  \\
+*INSERT BOXPLOT FROM STATUS REPORT HERE  \\
 We found no obvious trend indicating one site size to be more independent than 
@@
 	While this is an interesting observation, we would like to be able to 
-quantify it better by finding the distribution of our statistic.
+quantify it better by finding the distribution of our statistic.  This would 
+allow us to make statements about how independent the data is, and if the 
+difference in independence between different types of data is significant.   
+This statistic seems hard to get, but we considered a modification.  Let \(S\) be
+the number of successive \(0\),\(1\)'s minus the number of successive \(1\),\(1\)'s
+per flight.  Under the null hypothesis that each site is an independent Bernoulli
+trial with probability \(p\) we were able to compute the mean and standard 
+deviation of \(S\).
 
-		We wanted to find out if the assumption of independence across sites held
-true for our data from the DNR.  We also wanted to see what sampling techniques 
-led to the greatest independence between plots.  The different techniques that we
-compared were using different plot sizes, and using data that includes all plots
-or data that uses only every other plot. \\
+	The mean ...
 
-		If the observations at different sites are independent, then the 
-probability that there is a \(1\) at site \(i\)  is independent of site \(i-1\).  
-We let \(p\) be the number of \(0\)'s followed by a \(1\) as a fraction of all 
-successive pairs that start with a \(0\).  We let \(1-q\) be the number of \(1\)'s
-followed by a \(1\) as a fraction of all successive pairs that start with a \(1\).
-The difference of these two proportions \((=p-(1-q)\) is a measure of independence,
-where \(p-(1-q)\) being approximately \(0\) under the null hypothesis of 
-independence.  
-	\begin{enumerate}
-    	\item Using Markov Chains to find r. \\
-          * r = potential long-term probability of detection, based upon the 
-			short-term conditional probabilities p and q acquired from our data 
-			(infinite limit as n approaches infinity of Pn) \\
-          * p = probability that a 0 follows a 0, q = probability that a 1 follows
-			 a 1  \\
-          * If 0->0 and 1->0 (the values of a column, basically) equaled each 
-			other, plots would have been independent?  \\
-          * Ultimately not useful to us because it does not take into account 
-			whether or not a site is occupied, so it doesn't apply to the 
-			hierarchical model that we are using. \\
-   		\item Using s, where s = (\# of 0-1's) - (\# of 1-1's) for a given flight.\\
+	The standard deviation...
+
+	Simulations of \(S\) show that the distribution of \(S\) is roughly normal.  
+So using our calculations for the mean and the standard deviation, we computed
+a Z-score for each flight, where we use a plug-in estimate for \(p\) to be the 
+proportion of \(1\)'s per flight.  We said that Z-scores with an absolute value 
+greater than \(2\) suggest that the assumption of independence between sites is 
+violated for that flight.  We found that, in general, alternating sites are more
+"independent" than all sites as measured by this statistic.  However, alternating
+sites do not guarentee independence.  For instance, six of eleven flights along 
+the Zumbro River during 2003 still have significant Z-scores with alternating
+sites.  There were no noticable trends for this statistic with respect to 
+observer, days since snow, or snow event.
+
+*FIGURES OR TABLES??? WHICH ONES?
+
+	We did find interesting trends with regard to sampling method.  We think that 
+the lack of independence results from strings of ones where an otter left a track
+across multiple sites.  When alternating sites are used, the strings are cut in 
+half and the occurrence of \(1\),\(1\)'s is lower while the occurence of \(0\),
+\(1\)'s stays about the same.  It is noteworthy that virtually all the violations
+of independence are in the direction of more \(1\),\(1\)'s than would be expected.
+This reinforces our thoughts about successive \(1\)'s representing the same otter.
+We will discuss later whether this violation of independence actually matters in 
+our model, and how this might affect the estimation of actual occupancy rates.
+
+   		Using s, where s = (\# of 0-1's) - (\# of 1-1's) for a given flight.\\
           * Decided to use the z-score of s as a measure of independence.  \\
@@
 			more independence than all plots. Look at report for more. \\
-          * CAR model started working once we took s values of simulated data into
-			account by increasing dependence.  \\
-	\end{enumerate}
 

### 2010-02-23 kivaSections.tex
why am i writing ths?

@@
-kkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkk
+%kkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkk
+\documentclass[12pt]{article}
+\usepackage{graphicx,amssymb}
+
+\begin{document}
+\section{Questions of investigation}
+The North American river otter (\textit{Lontra canadensis}) is endemic to all of Minnesota; however, populations declined precipitously in the 19th and 20th centuries due to trapping and habitat degradation.
+In recent years their populations have significantly rebounded, especially in southeastern Minnesota where hunters would like to be able to trap otters again.
+In order for that to be possible governmental organizations must be able to monitor the population health.
+In addition, otter presence can indicate high quality aquatic habitat.
+Due to these reasons, there is interest in monitoring changes in otter abundance over time.
+Unfortunately, otters are not easy to observe.
+Thus, this study investigated the potential of using wintertime aerial surveys of otter tracks on rivers to make inferences about the abundance of otters and how it changes over time.
+We used the concepts from occupancy theory to develop models to estimate abundance.
+
+\subsection{Data collection}
+During the winters of 2003 and 2004 people flew over the Whitewater, Zumbrow, and Mississippi Rivers.
+When they saw what looked like an otter track, they pushed a button recording the GPS coordinates.
+They pushed the button every five seconds until the track ended.
+Flights occurred one, two, or three days after a snow event.
+In some cases, multiple observers flew over the same river on the same day.
+Surveys were conducted during multiple snow events throughout both winters.
+The rivers were then divided into 400 and 800 meter plots and the presence (1) or absence (0) of a waypoint in a plot on a given flight was determined.
+The Mississippi River was also divided into 1600 meter plots.
+
+\subsection{Issues related to the data}
+One major issue with respect to the data is that if a track occurred near the boundary between two plots, a person on one flight may have recorded the track in one site whereas someone on a later flight during that same snow event may have seen the same track but recorded it in the neighboring plot.
+So while the data indicate that the observers on the two flights disagreed in two plots, they saw the same track.
+Increasing the plot size should decrease the magnitude of this effect.
+
+In addition, there is concern that the probability of an otter laying down a track in one site is not independent of the probability that an otter lays down a track in neighboring sites, violating a principle assumption of occupancy theory.
+That is, a single track can extend across multiple plots.
+This leads to dependence among the plots.
+It was important to quantify this dependence and find an optimum way to handle it.
+
+\subsection{Descriptive statistics}
+
+In 2003, three different observers flew over the rivers.
+Assuming that no tracks are laid down between individuals’ flights on the same day, all three people should have seen the same tracks.
+While there is some error in classifying which site the observer saw the actual track in, each observer should at least be seeing the same, or nearly the same, number of sites with tracks.
+However, we did not find that to be the case.
+If we look only at days that all three people observed, there appear to be major differences among the proportion of sites in which the scientists detected a track (Figure x).
+The difference in estimated na\"ive occupancy among the observers indicates that the scientists did not always correctly identify the tracks.
+
+Correlation among the flight sequences is low.
+For example, average correlation between flights on the Mississippi in 2003 is 0.176 for 400 meter plots.
+Some of this is due to the difficulties in determining to which site each track belongs.
+However, we have also found quite a bit of variation in the data not necessarily due to these classification issues.
+This is difficult to quantify, as an algorithm for finding classification errors would be complicated.
+Correlation between flights is a bit higher for flights on the same day or flights during the same snow event, but still not as high as one might hope.
+The proportion of sites with tracks detected generally increases with days since snow, likely because the otters have a longer time to lay down a track as time after the snow event passes.
+
+\subsection{False positive rates}
+The lack of consistency among the flights is strong evidence that the data contain false positives where scientists think they see a track when it does not exist.
+Not accounting for such a process can lead to major biases; as the number of flights increases, the occupancy rate estimate will approach one (Royle and Link 2006).
+We used the observer as a covariate (by means of indicator variables) for both the probability of correctly identifying a track that exists ($p$) and the probability of identifying a track that does not exist ($E$).
+Including false positives introduces a duality whereby $(p = x, E = y, \psi = z)$ is as good a solution as $(p = y, E = x, \psi = 1-z)$.
+One possible way to address this problem is to assume $p\gg0.5$ and $E\ll0.5$.
+
+Figure x caption:
+The box plots show how the proportion of sites in which an observer saw tracks (“occupancy”) differs among the three observers. Only those dates in which all three observers flew are included.
+
+\end{document}

### 2010-02-23 ChrisnaSections.tex

@@
     Recall that our standard hierarchical model contains the following
-    parameters: \(\psi\) (the rate of occupancy), \(\theta\) (the probability
-    of laying down a track given occupancy), \(p\) (the probability of
-    detecting a track given that it is present, which is unique to each
-    observer and makes \(1-p\) the false negative rate of detecting a track),
-    and \(E\) (the false positive rate of detecting a track, which is unique to
-    each observer).
+    parameters: \(\psi\) (the rate of occupancy), \(\theta\) (the probability of
+    laying down a track given occupancy), \(p\) (the probability of detecting a
+    track given that it is present, which is unique to each observer and makes
+    \(1-p\) the false negative rate of detecting a track), and \(E\) (the false
+    positive rate of detecting a track, which is unique to each observer).
 
@@
     were: the observer, the date, and the number of days since the last snow
-    event.  In general, the sparseness of the data gathered in 2004 caused us
-    to simplify the model in order to obtain results.  Unfortunately, the
+    event.  In general, the sparseness of the data gathered in 2004 caused us to
+    simplify the model in order to obtain results.  Unfortunately, the
     simplification results in estimates of \(\psi\) that do not seem accurate
@@
     We also tried to simplify this model by eliminating \(E\) because this was
-    somewhat successful with respect to convergence with the standard model.
-    We first attempted to fit this model to data from 2003 at the 402m plot
-    size.  Although the model successfully converged on all of this data, its
-    estimates of \(\psi\), all of which were greater than 0.9, did not seem
-    reasonable, especially in comparison to the standard model's estimates when
-    given the same data.  As a result, we did not attempt to fit this model to
-    the rest of the data and do not include this model's estimates within our
-    results.
+    somewhat successful with respect to convergence with the standard model.  We
+    first attempted to fit this model to data from 2003 at the 402m plot size.
+    Although the model successfully converged on all of this data, its estimates
+    of \(\psi\), all of which were greater than 0.9, did not seem reasonable,
+    especially in comparison to the standard model's estimates when given the
+    same data.  As a result, we did not attempt to fit this model to the rest of
+    the data and do not include this model's estimates within our results.
 
@@
     2003 and 2004.  Below each table is detailed commentary about the ability of
-    the standard hierarchical model to convergence to the corresponding data. \\
+    the standard hierarchical model to convergence to the corresponding data.
+
+    With respect to the model's ability to converge on the actual data, the most
+    important factors are plot size, the number of flights recorded, and the
+    number of estimated parameters.  In general, both plot size and the number
+    of flights recorded must be sufficiently small and large, respectively, for
+    the standard model with \(E\) to converge.
+
+    2003's data worked relatively well with our model.  The presence of numerous
+    observations that occur in the second and third days since the most recent
+    snow event allows one of the more crucial assumptions of our hierarchical
+    model -- that otter tracks laid on each day after a snow event are visible
+    throughout the entire observation period and therefore, that it is more
+    likely for an observer to detect a track on later days -- to be thoroughly
+    tested.
+
+    2004's data, which suffered from the lack of multiple observers and
+    observations later than one day after each snow event, does not provide
+    enough data with respect to time (measured by the number of days since the
+    last snow event) for the parameters \(p\) and \(E\) to converge, regardless
+    of the number of Markov Chain Monte Carlo (MCMC) iterations performed.
 
     % Mississippi River!
-    \noindent{\begin{tabular}{|l|l|l|l|}
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|}
         \hline
@@
         \hline
-    \end{tabular}}
-    \\
+    \end{tabular}
+    \end{center}
 
     This data converged well at the 402m and 804m plot sizes.  At the 1608m plot
-    size, \(E\), \(\psi\), and \(\theta\) did not converge well. \\
+    size, \(E\), \(\psi\), and \(\theta\) did not converge well.
 
-    \noindent{\begin{tabular}{|l|l|}
+    \begin{center}
+    \begin{tabular}{|l|l|}
         \hline
@@
         \hline
-    \end{tabular}}
-    \\
+    \end{tabular}
+    \end{center}
 
     This data did not converge.  We do not think that these estimates are
-    accurate, but they are included for the sake of completeness. \\
+    accurate, but they are included for the sake of completeness.
 
     % Whitewater River!
-    \noindent{\begin{tabular}{|l|l|l|}
+    \begin{center}
+    \begin{tabular}{|l|l|l|}
         \hline
@@
         \hline
-    \end{tabular}}
-    \\
+    \end{tabular}
+    \end{center}
 
     While the 402m data converged well, \(E\), \(\psi\), and \(\theta\) did not
-    converge well for the 804m data. \\
+    converge well for the 804m data.
 
-    \noindent{\begin{tabular}{|l|l|l|l|}
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|}
         \hline
@@
         \hline
-    \end{tabular}}
-    \\
+    \end{tabular}
+    \end{center}
 
@@
     much less precise than those from 2003 -- the confidence intervals of
-    \(\psi\) and of \(p[Danny]\) are twice as large as those from 2003. \\
+    \(\psi\) and of \(p[Danny]\) are twice as large as those from 2003.
 
     % Zumbro River!
-    \noindent{\begin{tabular}{|l|l|l|}
+    \begin{center}
+    \begin{tabular}{|l|l|l|}
         \hline
@@
         \hline
-    \end{tabular}}
-    \\
+    \end{tabular}
+    \end{center}
 
     Again, the 402m data converged well.  With respect to the 804m data, while
-    \(\psi\) and \(\theta\) converged well, none of the other parameters did. \\
+    \(\psi\) and \(\theta\) converged well, none of the other parameters did.
 
-    \noindent{\begin{tabular}{|l|l|l|l|}
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|}
         \hline

### 2010-02-23 kivaSections.tex

@@
 \section{Questions of investigation}
-The North American river otter (\textit{Lontra canadensis}) is endemic to all of Minnesota; however, populations declined precipitously in the 19th and 20th centuries due to trapping and habitat degradation.
-In recent years their populations have significantly rebounded, especially in southeastern Minnesota where hunters would like to be able to trap otters again.
-In order for that to be possible governmental organizations must be able to monitor the population health.
-In addition, otter presence can indicate high quality aquatic habitat.
-Due to these reasons, there is interest in monitoring changes in otter abundance over time.
-Unfortunately, otters are not easy to observe.
-Thus, this study investigated the potential of using wintertime aerial surveys of otter tracks on rivers to make inferences about the abundance of otters and how it changes over time.
-We used the concepts from occupancy theory to develop models to estimate abundance.
+The North American river otter (\textit{Lontra canadensis}) is endemic to all of 
+Minnesota; however, populations declined precipitously in the 19th and 20th 
+centuries due to trapping and habitat degradation. In recent years their 
+populations have significantly rebounded, especially in southeastern Minnesota 
+where hunters would like to be able to trap otters again. In order for that to 
+be possible governmental organizations must be able to monitor the population 
+health. In addition, otter presence can indicate high quality aquatic habitat. 
+Due to these reasons, there is interest in monitoring changes in otter abundance 
+over time. Unfortunately, otters are not easy to observe. Thus, this study 
+investigated the potential of using wintertime aerial surveys of otter tracks on 
+rivers to make inferences about the abundance of otters and how it changes over 
+time. We used the concepts from occupancy theory to develop models to estimate 
+abundance.
 
 \subsection{Data collection}
-During the winters of 2003 and 2004 people flew over the Whitewater, Zumbrow, and Mississippi Rivers.
-When they saw what looked like an otter track, they pushed a button recording the GPS coordinates.
-They pushed the button every five seconds until the track ended.
-Flights occurred one, two, or three days after a snow event.
-In some cases, multiple observers flew over the same river on the same day.
-Surveys were conducted during multiple snow events throughout both winters.
-The rivers were then divided into 400 and 800 meter plots and the presence (1) or absence (0) of a waypoint in a plot on a given flight was determined.
-The Mississippi River was also divided into 1600 meter plots.
+During the winters of 2003 and 2004 people flew over the Whitewater, Zumbrow, 
+and Mississippi Rivers. When they saw what looked like an otter track, they 
+pushed a button recording the GPS coordinates. They pushed the button every five 
+seconds until the track ended. Flights occurred one, two, or three days after a 
+snow event. In some cases, multiple observers flew over the same river on the 
+same day. Surveys were conducted during multiple snow events throughout both 
+winters. The rivers were then divided into 400 and 800 meter plots and the 
+presence (1) or absence (0) of a waypoint in a plot on a given flight was 
+determined. The Mississippi River was also divided into 1600 meter plots.
 
 \subsection{Issues related to the data}
-One major issue with respect to the data is that if a track occurred near the boundary between two plots, a person on one flight may have recorded the track in one site whereas someone on a later flight during that same snow event may have seen the same track but recorded it in the neighboring plot.
-So while the data indicate that the observers on the two flights disagreed in two plots, they saw the same track.
-Increasing the plot size should decrease the magnitude of this effect.
+One major issue with respect to the data is that if a track occurred near the 
+boundary between two plots, a person on one flight may have recorded the track 
+in one site whereas someone on a later flight during that same snow event may 
+have seen the same track but recorded it in the neighboring plot. So while the 
+data indicate that the observers on the two flights disagreed in two plots, they 
+saw the same track. Increasing the plot size should decrease the magnitude of 
+this effect.
 
-In addition, there is concern that the probability of an otter laying down a track in one site is not independent of the probability that an otter lays down a track in neighboring sites, violating a principle assumption of occupancy theory.
-That is, a single track can extend across multiple plots.
-This leads to dependence among the plots.
-It was important to quantify this dependence and find an optimum way to handle it.
+In addition, there is concern that the probability of an otter laying down a 
+track in one site is not independent of the probability that an otter lays down 
+a track in neighboring sites, violating a principle assumption of occupancy 
+theory. That is, a single track can extend across multiple plots. This leads to 
+dependence among the plots. It was important to quantify this dependence and 
+find an optimum way to handle it.
 
@@
 
-In 2003, three different observers flew over the rivers.
-Assuming that no tracks are laid down between individuals’ flights on the same day, all three people should have seen the same tracks.
-While there is some error in classifying which site the observer saw the actual track in, each observer should at least be seeing the same, or nearly the same, number of sites with tracks.
-However, we did not find that to be the case.
-If we look only at days that all three people observed, there appear to be major differences among the proportion of sites in which the scientists detected a track (Figure x).
-The difference in estimated na\"ive occupancy among the observers indicates that the scientists did not always correctly identify the tracks.
+In 2003, three different observers flew over the rivers. Assuming that no tracks 
+are laid down between individuals’ flights on the same day, all three people 
+should have seen the same tracks. While there is some error in classifying which 
+site the observer saw the actual track in, each observer should at least be 
+seeing the same, or nearly the same, number of sites with tracks. However, we 
+did not find that to be the case. If we look only at days that all three people 
+observed, there appear to be major differences among the proportion of sites in 
+which the scientists detected a track (Figure x). The difference in estimated na
+\"ive occupancy among the observers indicates that the scientists did not always 
+correctly identify the tracks.
 
-Correlation among the flight sequences is low.
-For example, average correlation between flights on the Mississippi in 2003 is 0.176 for 400 meter plots.
-Some of this is due to the difficulties in determining to which site each track belongs.
-However, we have also found quite a bit of variation in the data not necessarily due to these classification issues.
-This is difficult to quantify, as an algorithm for finding classification errors would be complicated.
-Correlation between flights is a bit higher for flights on the same day or flights during the same snow event, but still not as high as one might hope.
-The proportion of sites with tracks detected generally increases with days since snow, likely because the otters have a longer time to lay down a track as time after the snow event passes.
+Correlation among the flight sequences is low. For example, average correlation 
+between flights on the Mississippi in 2003 is 0.176 for 400 meter plots. Some of 
+this is due to the difficulties in determining to which site each track belongs. 
+However, we have also found quite a bit of variation in the data not necessarily 
+due to these classification issues. This is difficult to quantify, as an 
+algorithm for finding classification errors would be complicated. Correlation 
+between flights is a bit higher for flights on the same day or flights during 
+the same snow event, but still not as high as one might hope. The proportion of 
+sites with tracks detected generally increases with days since snow, likely 
+because the otters have a longer time to lay down a track as time after the snow 
+event passes.
 
 \subsection{False positive rates}
-The lack of consistency among the flights is strong evidence that the data contain false positives where scientists think they see a track when it does not exist.
-Not accounting for such a process can lead to major biases; as the number of flights increases, the occupancy rate estimate will approach one (Royle and Link 2006).
-We used the observer as a covariate (by means of indicator variables) for both the probability of correctly identifying a track that exists ($p$) and the probability of identifying a track that does not exist ($E$).
-Including false positives introduces a duality whereby $(p = x, E = y, \psi = z)$ is as good a solution as $(p = y, E = x, \psi = 1-z)$.
-One possible way to address this problem is to assume $p\gg0.5$ and $E\ll0.5$.
+The lack of consistency among the flights is strong evidence that the data 
+contain false positives where scientists think they see a track when it does not 
+exist. Not accounting for such a process can lead to major biases; as the number 
+of flights increases, the occupancy rate estimate will approach one (Royle and 
+Link 2006). We introduced a false detection process and used the observer as a 
+covariate (by means of indicator variables) for both the probability of 
+correctly identifying a track that exists ($p$) and the probability of 
+identifying a track that does not exist ($E$). Including false positives 
+introduces a duality whereby $\begin{array}{ccc}(p = x, & E = y, & \psi = z)\end
+{array}$ is as good a solution as $\begin{array}{ccc}(p = y, & E = x, & \psi = 
+1-z)\end{array}$. To address this problem we assumed $p\gg0.5$ and $E\ll0.5$, an 
+approach recommended by Royle and Link (2006).
 
 Figure x caption:
-The box plots show how the proportion of sites in which an observer saw tracks (“occupancy”) differs among the three observers. Only those dates in which all three observers flew are included.
+The box plots show how the proportion of sites in which an observer saw tracks 
+(“occupancy”) differs among the three observers. Only those dates in which all 
+three observers flew are included.
+
+\section{Spatial correlation of the data}
+We found that there is significant correlation between the probability of 
+observing a track in neighboring sites, but this violates of a central 
+assumption of occupancy theory. This issue needed to be addressed. One approach 
+was to make the prior distribution of $\theta$, the probability that an otter 
+lays down a track, spatially correlated using a conditional autoregressive
+(CAR) model.
 

### 2010-02-24 ChrisnaSections.tex
Sorta done with results; starting covariates talk.

@@
     2003 and 2004.  Below each table is detailed commentary about the ability of
-    the standard hierarchical model to convergence to the corresponding data.
+    the standard hierarchical model to converge, given the corresponding data.
 
@@
 
-    2003's data worked relatively well with our model.  The presence of numerous
+    2003's data worked relatively well with the model.  The presence of numerous
     observations that occur in the second and third days since the most recent
@@
     likely for an observer to detect a track on later days -- to be thoroughly
-    tested.
+    tested.  Most of the \(\psi\) and \(\theta\) estimates of 2003's data fall
+    into ranges where our model is at its best, both with respect to convergence
+    and to accurately estimating \(\psi\).
 
@@
     of the number of Markov Chain Monte Carlo (MCMC) iterations performed.
+
+    % Wow, this stuff SUCKS.
+    As plot size increases, every estimated parameter tends to increase as well.
+    Although the vast majority of these increases, even at the 1608m plot size,
+    lie well within the 95\% confidence interval of the accompanying estimate at
+    the 402m plot size, we believe that this trend makes sense.  Only one track
+    detection within a plot is necessary to assign it a 1.  Therefore, once 402m
+    plots are aggregated into 1608m plots, only one of the four 402m plots needs
+    to be assigned a 1 in order for the entire 1608m plot to receive a 1.  The
+    amount of correlation between 1's determines the increase in the proportion
+    of 1's to the number of plots -- as the amount of correlation increases, the
+    increase in the proportion decreases.
+
+    We believe that this increase leads to higher \(\psi\) and \(\theta\)
+    estimates because the potential for multiple otters to lay tracks within the
+    same plot increases.  In southeastern Minnesota, most otters tolerate the
+    presence of other otters in the areas they use most frequently, suggesting
+    that they are neither solitary nor territorial \cite{Gorman2006}.  This is
+    consistent with the noticeable but modest increase in \(\theta\) across plot
+    % I have no idea how to explain this... :(.
+    sizes for all three rivers.  However, estimates of \(\psi\) along the
+    Whitewater and Zumbro Rivers remain constant, despite large increases in
+    \(\theta\).
+
+    % More detail, maybe?
+    Similarly, the increases in estimates of \(p\) and \(E\) seem to result from
+    the general increase of tracks laid per plot as the plot size grows.  As
+    plot size grows, the influence of a 1 increases, regardless of whether an
+    otter track or a track from a different animal was observed.
+
+    Lastly, we decided to use the data from every plot even though we believe
+    that, as a result, our data violates the assumption of independence between
+    sites.  However, the model remains effective with spatially correlated data.
+    Since the actual data is relatively sparse as it is, we consider the removal
+    of half of it in order to preserve independence to be more harmful to the
+    model than supplying it with correlated data.  Therefore, all estimates come
+    from all of the data that we were given.
 
@@
 
-    %%
-    \subsection{The Effect of Plot Size}
-    %%
-    \subsection{All Plots vs. Every Other Plot}
-    %%
-    \subsection{How Important Are Our Covariates?}
-    %%
+    %% COVARIATES:
+    \subsection{Covariates}
+    To determine the relative importance of our covariates (the observer, the
+    date, and the number of days since the most recent snow event), we ran the
+    standard hierarchical model with \(E\) on simulated data with values of
+    \(\psi\) and \(\theta\) similar to our estimates of \(\psi\) and \(\theta\)
+    on 2003's data with varying numbers of observers, snow events, and
+    observations after each snow event.
+
     \subsection{Best Practices for Measuring Otter Occupancy}
 
+\bibliographystyle{alpha}
+\bibliography{Bibliography}
 \end{document}

### 2010-02-24 sarahDraft.tex

@@
 \usepackage{fullpage}
+\usepackage[hmargin=3cm, vmargin=3cm]{geometry}
 \begin{document}
@@
 \section{Overview of Occupancy Modeling}
+
 	\subsection{What is Occupancy Rate?}
-		"The probability that a randomly selected site or sampling unit in an area 
-			of interest is occupied by a species" [MacKenzie2006] (pg. 11)
+
+		Occupancy is "The probability that a randomly selected site or sampling 
+unit in an area of interest is occupied by a species" [MacKenzie2006] This is 
+not the same as the abundance, or population size, of a species.  Instead of 
+measuring the number of animals present in a species, occupancy measures the 
+proportion of land that is used as the habitat of at least one member of the 
+species. 
+	is either the probability or the proportion 25-  note importance of the 
+distinction although they can be considered the same thing in most cases
+
+		*What is it used for?
+
 	\subsection{Different Ways of Sampling Data}
-		Measuring rate of occupancy vs. population \\
-		Looking at tracks vs. listening to stuff or acutally looking for the animal
+		Most common way of sampling:  visit sites and spend time within each one 
+looking for either individuals of the species or for evidence the species is 
+present- "presence-absence survey"(26).  can use visual observation of animals,
+capture animals, ovserve tracks, detect vocalizations, camera traps, sound 
+recording.  leads to naive occupancy rate.  PROBLEM: false negatives.
+
+		*Measuring rate of occupancy vs. population \\
+		*Looking at tracks vs. listening to stuff or acutally looking for the 
+			animal
+
 	\subsection{Assumptions}
-		Independence of detection between sites \\
-		No false positives \\
-		Presence and detection probabilities constant across sites and surveys \\
-		Closed population
+		*Independence of detection between sites \\
+		*No false positives \\
+		*Presence and detection probabilities constant across sites and surveys \\
+		*Closed population
+
 	\subsection{Types of Occupancy Models}
-		Maximum Likelihood \\
-		any other models? \\
-		Hierarchical Bayesian?  is that the name of what we used?
+		We only considered single-season, single-species models because this is 
+the data we had, but any of these models can be extended to take into account
+multiple seasons or multiple species.  
+		Naive Occupancy pg 11- this is when you just divide the number of sites
+			that are indicated as occupied by the total number of sites.  it is
+			generally accepted that this is too low because not all animals are 
+			detected.
+		Hierarchical Models:
+		Maximum Likelihood \\  Give the simple model, then say it can be 
+			modified in many different ways.  give some examples possibly 
+			without going into detail.
+		Any other models? \\
+		Bayesian?  is that the name of what we used?  should save most of this 
+			for theoretical section?
 
@@
 Stuff by Kiva.
+
 	\subsection{Markov Chains and Testing the Independence Assumption}
@@
 probability of detection, which we called \(r\).  We let \(p\) be the number of 
-\(0\)'s followed by a \(0\) as a fraction of all successive pairs that start with 
-a \(0\).  We let \(q\) be the number of \(1\)'s followed by a \(1\) as a fraction 
-of all successive pairs that start with a \(1\).  These probabilities can be
-represented by the matrix \(M = \begin{bmatrix} p & 1-p \\ 1-q & q 
-\end{bmatrix}\). The long term probability of detection can be found from this by finding 
-\(\lim_{n \to \inf} M^n\), which returns \( \begin{bmatrix} 1-r & r \\ 1-r & r 
-\end{bmatrix}\), where \(r\) is the probability of a \(1\), which is the long 
-term probability of detection.
+\(0\)'s followed by a \(0\) as a fraction of all successive pairs that start 
+with a \(0\).  We let \(q\) be the number of \(1\)'s followed by a \(1\) as a 
+fraction of all successive pairs that start with a \(1\).  These probabilities 
+can be represented by the matrix \(M = \begin{bmatrix} p & 1-p \\ 1-q & q 
+\end{bmatrix}\). The long term probability of detection can be found from this 
+by finding \(\lim_{n \to \inf} M^n\), which returns \( \begin{bmatrix} 1-r & r 
+\\ 1-r & r\end{bmatrix}\), where \(r\) is the probability of a \(1\), which is 
+the long term probability of detection.
 
@@
 asking.  It does not take into account whether or not a site is occupied, or 
-whether or not a track was laid down on that site, or any of the ideas behind the
-hierarchical model.  Thesse are the parameters that we eventually want to find 
-out something about, and the value of \(r\) cannot help us get there.  The ideas 
-behind using Markov Chains did end up being useful, however.
+whether or not a track was laid down on that site, or any of the ideas behind    
+the hierarchical model.  These are the parameters that we eventually want to 
+find out something about, and the value of \(r\) cannot help us get there.  The 
+ideas behind using Markov Chains did end up being useful, however.
 
-	Although we rejected the idea of using \(r\) as a way to analyze the data, we
-realized that \(p\) and \(q\) could still be used to test the viability of the 
-independence assumption.  If there is independece between sites, the difference 
-of the proportions \(p\) and \(1-q\) should be approximately \(0\).  This would 
-mean that the probability that a \(0\) follows a \(0\) is about the same as the 
-probability that a \(1\) follows a \(0\), meaning that it does not depend on what
-the observation is at the previous site.  Looking at boxplots of \(p - (1-q)\), 
-we see that data from flights that record alternating sites may be more 
-independent than data from flights that record all sites, which suggests a 
-possible benefit to using the alternating sampling method.  The boxplots for the 
-Whitewater and Zumbro Rivers looked similar to the Mississippi:  \\
+	Although we rejected the idea of using \(r\) as a way to analyze the data, 
+we realized that \(p\) and \(q\) could still be used to test the viability of 
+the independence assumption.  If there is independece between sites, the 
+difference of the proportions \(p\) and \(1-q\) should be approximately \(0\). 
+This would mean that the probability that a \(0\) follows a \(0\) is about the 
+same as the probability that a \(1\) follows a \(0\), meaning that it does not 
+depend on what the observation is at the previous site.  Looking at boxplots of 
+\(p - (1-q)\), we see that data from flights that record alternating sites may 
+be more independent than data from flights that record all sites, which suggests
+a possible benefit to using the alternating sampling method.  The boxplots for 
+the Whitewater and Zumbro Rivers looked similar to the Mississippi:  \\
 *INSERT BOXPLOT FROM STATUS REPORT HERE  \\
@@
 difference in independence between different types of data is significant.   
-This statistic seems hard to get, but we considered a modification.  Let \(S\) be
-the number of successive \(0\),\(1\)'s minus the number of successive \(1\),\(1\)'s
-per flight.  Under the null hypothesis that each site is an independent Bernoulli
-trial with probability \(p\) we were able to compute the mean and standard 
-deviation of \(S\).
+This statistic seems hard to get, but we considered a modification.  Let \(S\) 
+be the number of successive \(0\),\(1\)'s minus the number of successive \(1\),
+\(1\)'s per flight.  Under the null hypothesis that each site is an independent 
+Bernoulli trial with probability \(p\) we were able to compute the mean and 
+standard deviation of \(S\), which we used to calculate a Z-score.
+	In order to compute these values, we needed to find the expected value and
+variance of each term in \(S\).  We let \(A = \text{\# of 0-1's}\), 
+\(B = \text{\# of 1-1's}\) and \(p = P(``1")\text{ in the sequence}\).
 
 	The mean ...
+ 	\begin{align*}
+        E[A] 
+            &= \sum_{i=1}^{n-1} P(X_i = 0, X_{i+1} = 1) \\
+            &= \sum_{i=1}^{n-1} (1-p)p \\
+            &= (n-1)(1-p)p
+    \end{align*}
+    \begin{align*}
+        E[B]
+            &= \sum_{i=1}^{n-1} P(X_i = 1, X_{i+1} = 1) \\
+            &= \sum_{i=1}^{n-1} p*p \\
+            &= (n-1)p^2
+    \end{align*}
+	\begin{align*}
+        E[\hat{s}] 
+            &= E[A] - E[B] \\
+            &= (n-1)\bigg((1-p)p - p^2\bigg)
+    \end{align*}
 
 	The standard deviation...
+	\begin{align*}
+        Var[A] 
+            &= \sum_{i=1}^{n-1} Var[A] + 2\sum_{i=1}^{n-2} Cov[I_i = (0, 1),
+               I_{i+1} = (1, 1)] \\
+            &= (n-1)P((0, 1))(1-P((0, 1))) + 2\sum_{i=1}^{n-2}\bigg(E[I_i=(0, 
+               1), I_{i+1} = (0, 1)] - E[I_i = (0, 1)]E[I_{i+1} = (0, 1)]\bigg)
+               \\
+            &= (n-1)(1-p)p(1-(1-p)p) + 2\sum_{i=1}^{n-2}\bigg(0 - (1-p)p(1-p)p
+               \bigg) \\
+            &= (1-p)p(-1+n+5p-3np-5p^2+3np^2)
+    \end{align*}
+    \begin{align*}
+        Var[B] 
+            &= \sum_{i=1}^{n-1} Var[B] + 2\sum_{i=1}^{n-2} Cov[I_i = (1, 1),
+               I_{i+1} = (1, 1)] \\
+            &= (n-1)P((1, 1))(1-P((1, 1))) + 2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,   
+               1), I_{i+1} = (1, 1)] - E[I_i = (1, 1)]E[I_{i+1} = (1, 1)]\bigg) 
+               \\
+            &= (n-1)p^2(1-p^2) + 2(n-2)(p^3-p^4) \\
+            &= (1-p)p^2(-1+n-5p+3np)
+    \end{align*}
+	\begin{align*}
+		2Cov[A, B]  
+			&= 2 \sum_{i=1}^{n-1} \sum_{j=1}^{n-1} Cov[A, B]\\
+			&= 2 \bigg(\sum_{i=1}^{n-2} Cov[AB] + \sum_{i=1}^{n-2} Cov[BA] +
+				\\ \sum_{i=1}^{n-1} Cov[X_i = 0 \text{ or } 1, X_{i+1} = 1]\bigg) \\
+			&= 2 \bigg(\sum_{i=1}^{n-2} E[AB] - E[A]E[B] + 
+				\sum_{i=1}^{n-2} E[BA] - E[B]E[A] + \\
+				\sum_{i=1}^{n-1} E[X_i = 0 \text{ or } 1, X_{i+1} = 1] - E[B]E[A] \\
+			&= 2(n-2)(p-1)^2p^2 - 2(n-2)p^3(1-p) - 2(n-1)p^3(1-p)			
+	\end{align*}	
+    \begin{align*}
+        Var[\hat{s}] 
+            &= Var[A-B] \\
+            &= Var[A] + Var[B] - 2Cov[A, B] \\
+            &= (1-p)p(-1+n+8p-4np-20p^2+12np^2)
+    \end{align*}
 
@@
 sites.  There were no noticable trends for this statistic with respect to 
-observer, days since snow, or snow event.
+observer, days since snow, snow event, or plot size.
 
@@
 
-	We did find interesting trends with regard to sampling method.  We think that 
-the lack of independence results from strings of ones where an otter left a track
-across multiple sites.  When alternating sites are used, the strings are cut in 
-half and the occurrence of \(1\),\(1\)'s is lower while the occurence of \(0\),
-\(1\)'s stays about the same.  It is noteworthy that virtually all the violations
-of independence are in the direction of more \(1\),\(1\)'s than would be expected.
-This reinforces our thoughts about successive \(1\)'s representing the same otter.
-We will discuss later whether this violation of independence actually matters in 
-our model, and how this might affect the estimation of actual occupancy rates.
+	We did find interesting trends with regard to sampling method.  We think 
+that the lack of independence results from strings of ones where an otter left a
+track across multiple sites.  When alternating sites are used, the strings are 
+cut in half and the occurrence of \(1\),\(1\)'s is lower while the occurence of 
+\(0\),\(1\)'s stays about the same.  It is noteworthy that virtually all the 
+violations of independence are in the direction of more \(1\),\(1\)'s than would
+be expected.  This reinforces our thoughts about successive \(1\)'s representing
+the same otter.  We will discuss later whether this violation of independence 
+actually matters in our model, and how this might affect the estimation of 
+actual occupancy rates.
 
-   		Using s, where s = (\# of 0-1's) - (\# of 1-1's) for a given flight.\\
-          * Decided to use the z-score of s as a measure of independence.  \\
-          * Therefore, needed the expected value and the standard deviation of s. \\
-          * So, we needed the expected value of each term and the variance, which 
-			required finding the variance of each term and the covariance between 
-			the two terms. \\
-          * Expected value is straightforward and we have all of the equations. \\
-          * Variance is fun and we also have all of the relevant equations. \\
-          * Calculated z-score for every flight and box-plotted them for each year
-			and plot size and pitted data with every other plot against data with
-			all plots. \\
-          * Nothing important with respect to plot size. Alternating plots show 
-			more independence than all plots. Look at report for more. \\
+\section{Theoretical Stuff}
 
-\section{Theoretical Stuff} - CHRISNA??
 	\subsection{Bayesian Statistics}
@@
 			distribution- bimodal issues.
-	\subsection{BRugs/R}
-    	All of our statistical analysis was performed in R.  \\
-    	All of our modeling was performed using BRugs, which allowed us to use 
-			WinBUGS commands and stuff in R.
+
+	\subsection{BRugs/R} - Chrisna
 

### 2010-02-24 ChrisnaSections.tex
Finished covariates; onto best practices!

@@
     \subsection{Covariates}
-    To determine the relative importance of our covariates (the observer, the
-    date, and the number of days since the most recent snow event), we ran the
+    To determine the relative importance of our covariates (observer,
+    date, and number of days since the most recent snow event), we ran the
     standard hierarchical model with \(E\) on simulated data with values of
@@
     on 2003's data with varying numbers of observers, snow events, and
-    observations after each snow event.
+    observations after each snow event.  We assume that observations after each
+    snow event will occur on consecutive days, starting from the first day after
+    the snow event.  This information helped us to find the most efficient ways
+    (with respect to number of flights) to gather aerial occupancy data given
+    one, two, and three observers while still obtaining useful estimates from
+    the model.
 
-    \subsection{Best Practices for Measuring Otter Occupancy}
+    The number of observers proved to be the least important of the three
+    covariates.  We were able to obtain accurate estimates regardless of the
+    number of observers, as long as there were sufficient numbers of snow events
+    and observations after each snow event.  For one observer, at least three
+    snow events and three observations on consecutive days after each snow event
+    are required for the model to converge and give good estimates.  For two or
+    three observers, two events with three observations on consecutive days
+    after each snow event is sufficient.  Therefore, roughly nine, twelve, and
+    eighteen flights are required to obtain good estimates for one, two, and
+    three observers, respectively.  Surprisingly, the most cost-effective method
+    (as measured by the number of flights required) only requires one observer.
+
+    The number of snow events proved to be the second-most important of the
+    three covariates.  Although two snow events are required, only with one
+    observer does the addition of a third snow event noticeably affect the
+    model's estimates.
+
+    The number of observations after each snow event proved to be the most
+    important of the three covariates.  Because the model relies upon the
+    assumption that the probability of seeing a track increases each day because
+    tracks made on previous days within the same snow event can still be seen,
+    this is not surprising.  Two observations after each snow event on
+    consecutive days are required.  Additional observations after each event
+    help the model converge more quickly, but the estimates become only slightly
+    more accurate.
+
+    Additional observations are more helpful than additional snow events at
+    first, although both are much more helpful than the addition of an observer.
+    Once the numbers of snow events and observations after each snow event both
+    surpass three, they improve the model's estimates at roughly equal rates.
+    Note, however, that increasing the number of snow events is likely to be far
+    less challenging than increasing the number of observations after each snow
+    event at that point.
+
+    \subsection{Best Practices for Measuring Otter Populations}
 

### 2010-02-24 kivaSections.tex
I'm confused
wtf

@@
-%kkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkkk
+pppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppp
 \documentclass[12pt]{article}
@@
 observed, there appear to be major differences among the proportion of sites in 
-which the scientists detected a track (Figure x). The difference in estimated na
-\"ive occupancy among the observers indicates that the scientists did not always 
-correctly identify the tracks.
+which the scientists detected a track (Figure x). The difference in estimated 
+na\"ive occupancy among the observers indicates that the scientists did not 
+always correctly identify the tracks.
 
@@
 
-Figure x caption:
-The box plots show how the proportion of sites in which an observer saw tracks 
-(“occupancy”) differs among the three observers. Only those dates in which all 
-three observers flew are included.
-
 \section{Spatial correlation of the data}
@@
 
+Figure x caption:
+The box plots show how the proportion of sites in which an observer saw tracks 
+(``occupancy") differs among the three observers. Only those dates in which all 
+three observers flew are included.
+
 \end{document}

### 2010-02-24 kivaSections.tex
wtf

@@
-pppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppppp
 \documentclass[12pt]{article}
@@
 In 2003, three different observers flew over the rivers. Assuming that no tracks 
-are laid down between individuals’ flights on the same day, all three people 
+are laid down between individuals' flights on the same day, all three people 
 should have seen the same tracks. While there is some error in classifying which 
@@
 The box plots show how the proportion of sites in which an observer saw tracks 
-(``occupancy") differs among the three observers. Only those dates in which all 
+(``occupancy'') differs among the three observers. Only those dates in which all 
 three observers flew are included.

### 2010-02-24 ChrisnaSections.tex
Discussion: done!  Hierarchical models: uh oh.

@@
 \usepackage{amsmath}
+\usepackage[hmargin=3cm, vmargin=3cm]{geometry}
 \usepackage{multirow}
@@
 
-%%% *STANDARD HIERARCHICAL MODEL*:
 \section{Standard Hierarchical Model}
-
-    %% MODEL PARAMETERS:
     \subsection{Model Parameters}
-
-    %% SIMULATED DATA, PT. 1:
     \subsection{Simulated Data}
-
     \noindent{\begin{tabular}{|l|l|l|l|}
@@
 
-%%% *DISCUSSION*:
 \section{Discussion}
-
-    %% GETTING THE ACTUAL DATA TO CONVERGE:
     \subsection{Obtaining Convergence on the Actual Data}
@@
 
-    %% RESULTS:
     \subsection{Results}
@@
 
-    %% COVARIATES:
     \subsection{Covariates}
@@
 
+    \subsection{Future Work}
     Additional observations are more helpful than additional snow events at
@@
     less challenging than increasing the number of observations after each snow
-    event at that point.
+    event at that point, especially in southeastern Minnesota.
 
-    \subsection{Best Practices for Measuring Otter Populations}
+    In summary, for future data gathering trips, we strongly suggest using one
+    observer, as long as they are relatively confident that they can correctly
+    identify an otter track and distinguish an otter track from the tracks of
+    other animals in the immediate area, since the model most accurately
+    estimates \(p\) and \(E\) when their true values are high and low,
+    respectively.  With one observer, three snow events and three or four snow
+    days' worth of flights per snow event would be sufficient.  If the same
+    observer flies twice (or more often) in one day, perhaps they could act as
+    a substitute for an additional observer.  This is because it is probable
+    that, with one flight's worth of experience observing the given river, the
+    observer's \(p\) and \(E\) would change due to their newfound familiarity
+    with the river.
 

### 2010-02-24 sarahDraft.tex
Types of Occupancy Models includes MLE

@@
 
-		Occupancy is "The probability that a randomly selected site or sampling 
-unit in an area of interest is occupied by a species" [MacKenzie2006] This is 
-not the same as the abundance, or population size, of a species.  Instead of 
-measuring the number of animals present in a species, occupancy measures the 
-proportion of land that is used as the habitat of at least one member of the 
-species. 
-	is either the probability or the proportion 25-  note importance of the 
-distinction although they can be considered the same thing in most cases
+		Occupancy has two similar definitions.  It can be "the probability that a
+randomly selected site or sampling unit in an area of interest is occupied by a 
+species" [MacKenzie2006].  It also can be used to describe "the proportion of 
+area, patches, or sample units that is occupied" [MacKenzie2006].  Occupancy is 
+not the same as the abundance, or population size, of a species.  The number of
+members of a species is not important, only whether or not at least one member of
+the species is present in each site.
 
-		*What is it used for?
+	While there is a difference between these two definitions of occupancy, in 
+general the probability is unknown and the observed proportion is used as an 
+estimate of the underlying probability.  Because of this, these two definitions
+of occupancy are oftentimes used interchangably.
+
+	The main reasons that people are interested in measuring occupancy are science
+or conservation and management [MacKenzie2006].  This statistic can give insight 
+into the way a system works.  It also can be used to monitor animals habitats 
+and discover trends over time in animal populations.
 
 	\subsection{Different Ways of Sampling Data}
-		Most common way of sampling:  visit sites and spend time within each one 
-looking for either individuals of the species or for evidence the species is 
-present- "presence-absence survey"(26).  can use visual observation of animals,
-capture animals, ovserve tracks, detect vocalizations, camera traps, sound 
-recording.  leads to naive occupancy rate.  PROBLEM: false negatives.
 
-		*Measuring rate of occupancy vs. population \\
-		*Looking at tracks vs. listening to stuff or acutally looking for the 
-			animal
+		In order to determine occupancy rate, the data is often collected in the
+ form of a presence-absence survey.  This means that observers visit sites and 
+spend time within each one looking for the animals themselves, tracks, 
+vocalizations, or some other indicator that the animal is present.  In some cases, 
+camera traps and sound recording are used instead of human observers.
+
+	The main problem with these methods of sampling data is that it is possible 
+for an animal to occupy a site, but not be detected by the observer.  This means
+that the site is recorded as absent, but there really is an animal there.  This 
+is called a false negative.  Marking an animal present when there really is no
+animal is also a possible problem, but is thought to be much less likely and is
+often ignored.  We will discuss this more below.
 
 	\subsection{Assumptions}
-		*Independence of detection between sites \\
-		*No false positives \\
-		*Presence and detection probabilities constant across sites and surveys \\
-		*Closed population
+
+	There are some assumptions that are made in order to create models of 
+occupancy.  Without these assumptions, the models become much more complicated.
+One common assumption is no false positives, which means if a site is recorded as
+present, this cannot be incorrect.  This is an assumption that we attempted to 
+get rid of in our models.  A second common assumption is that presence and 
+detection probabilties are constant across sites and surveys.  A third common 
+assumption is that there is independence of detection between sites.  A fourth
+common assumption is that the surveys are being conducted on a closed population.
+This means that no animals leave the area or join the area during the time of the
+surveys.
+
+	Obviously, not all of these assumptions are always true.  In some cases, there
+are established ways to deal with getting rid of an assumption in the models, 
+and in others there are no common ways to get rid of the assumption.  Either way, 
+it adds a level of complication to the model, and therefore we will only talk 
+about dealing with the assumption of no false positives, and independence of 
+detection across sites.
 
 	\subsection{Types of Occupancy Models}
-		We only considered single-season, single-species models because this is 
-the data we had, but any of these models can be extended to take into account
-multiple seasons or multiple species.  
-		Naive Occupancy pg 11- this is when you just divide the number of sites
-			that are indicated as occupied by the total number of sites.  it is
-			generally accepted that this is too low because not all animals are 
-			detected.
-		Hierarchical Models:
-		Maximum Likelihood \\  Give the simple model, then say it can be 
-			modified in many different ways.  give some examples possibly 
-			without going into detail.
-		Any other models? \\
-		Bayesian?  is that the name of what we used?  should save most of this 
-			for theoretical section?
+
+		All of the models that we looked at were single-species, single-season
+models because this is what fit our data, but any of these models can be modified
+to take into account multiple seasons or multiple species.
+
+	The most simple way to measure occupancy is to simply divide the number 
+of sites where an animal was found to be present by the total number of sites.  
+This is called naive occupancy [MacKenzie2006].  It is generally accepted that 
+this estimate is too low because animals are not detected at all sites where they
+are present, which is the issue of false negatives.  The other occupancy models
+attempt to do better than this simple estimate.
+
+	One way to create more accurate models is to use a Hierarchical model to 
+think about the parameters that affect occupancy.
+
+	A more complicated model that can be used as a Hierarchical model is Maximum
+Likelihood Estimation (MLE).  This is done by finding the probability 
+distribution of the observed data as a function of the parameters.  This 
+distribution is then said to be the function of the parameters conditional on the
+data, and becomes the likelihood function.  Then, the values of the parameters 
+that maximize the likelihood function are found, which means that we find the 
+parameters that make the data most likely given the underlying model.  
+[MacKenzie2006]
+
+	Another way to do this is using a Bayesian model, which we will discuss in 
+more detail below.
 

### 2010-02-24 sarahDraft.tex

@@
 \usepackage[hmargin=3cm, vmargin=3cm]{geometry}
+\usepackage[all]{xy}
 \begin{document}
@@
 		All of the models that we looked at were single-species, single-season
-models because this is what fit our data, but any of these models can be modified
-to take into account multiple seasons or multiple species.
+models because this is what fit our data, but these models can be modified to
+be used for multiple seasons or multiple species.
 
@@
 this estimate is too low because animals are not detected at all sites where they
-are present, which is the issue of false negatives.  The other occupancy models
+are present, which is the issue of false negatives.  Other occupancy models
 attempt to do better than this simple estimate.
 
-	One way to create more accurate models is to use a Hierarchical model to 
-think about the parameters that affect occupancy.
+	One way to create more accurate models is to use a Hierarchical model as a 
+way to think about the parameters that affect occupancy.  This allows us to take
+into account that when an animal is not detected, it could either not be there, 
+or it could be there and just was not detected.  This could either be due to
+a false negative, where the observer did not see the indicator, or it could be 
+due to the fact that the animal does occupy the site but there was no indication
+of the animal at the site- for example, no track, no vocalization, etc.  \\
+Here is a visualization of how the Hierarchical model works:
+	\begin{displaymath}
+		\xymatrix{          	   &                                  		 &                               		&\text{Site} \ar[dl] \ar[dr] & \\
+				            	   &                                  		 & \text{Occupied} \ar[dl] \ar[d]		&                            & \text{Unoccupied} \ar[d] \\
+				            	   & \text{Presence Indicated} \ar[dl] \ar[d]& \text{Presence Not Indicated} \ar[d]&                            & \text{Not Detected} \\
+				   \text{Detected} & \text{Not Detected}              		 & \text{Not Detected}                  &                            &   }
+	\end{displaymath}
 
-	A more complicated model that can be used as a Hierarchical model is Maximum
-Likelihood Estimation (MLE).  This is done by finding the probability 
-distribution of the observed data as a function of the parameters.  This 
-distribution is then said to be the function of the parameters conditional on the
-data, and becomes the likelihood function.  Then, the values of the parameters 
-that maximize the likelihood function are found, which means that we find the 
-parameters that make the data most likely given the underlying model.  
-[MacKenzie2006]
-
-	Another way to do this is using a Bayesian model, which we will discuss in 
-more detail below.
+From this diagram, you can see that we need separate parameters for the 
+probability that a site is occupied, the probability that presence is indicated
+given that a site is occupied, and the probability of detection given that 
+presence is indicated and the site is occupied.  We call these \(\psi\),
+\(\theta\), and \(p\), respectively.
 
@@
 	\subsection{Bayesian Statistics}
-	   	Bayes' Theorem \\
+
+	   	Bayes' Theorem: \(P(\theta|X) = P(X|\theta)P(\theta) / P(X)\) \\
+
+	Usually, the mean of the posterior distribution is then used as a point 
+estimate for the parameter.
+
+Therefore, the posterior distribution \( \propto \) the likelihood times the prior.
     	Useful for hierarchical modeling: look at section 3.3 in book. \\

### 2010-02-24 sarahDraft.tex
Added Chrisna's fixed equations

@@
 	The mean ...
- 	\begin{align*}
-        E[A] 
-            &= \sum_{i=1}^{n-1} P(X_i = 0, X_{i+1} = 1) \\
-            &= \sum_{i=1}^{n-1} (1-p)p \\
-            &= (n-1)(1-p)p
-    \end{align*}
-    \begin{align*}
-        E[B]
-            &= \sum_{i=1}^{n-1} P(X_i = 1, X_{i+1} = 1) \\
-            &= \sum_{i=1}^{n-1} p*p \\
-            &= (n-1)p^2
-    \end{align*}
-	\begin{align*}
-        E[\hat{s}] 
-            &= E[A] - E[B] \\
-            &= (n-1)\bigg((1-p)p - p^2\bigg)
-    \end{align*}
+ 	  % E[A]
+    \begin{equation*}
+        \begin{split}
+        E[A]& =\sum_{i=1}^{n-1}P(X_i=0,X_{i+1}=1) \\
+            & =\sum_{i=1}^{n-1}(1-p)p \\
+            & =(n-1)(1-p)p \\
+        \end{split}
+    \end{equation*}
+
+    % E[B]
+    \begin{equation*}
+        \begin{split}
+        E[B]& =\sum_{i=1}^{n-1}P(X_i=1,X_{i+1}=1) \\
+            & =\sum_{i=1}^{n-1}p*p \\
+            & =(n-1)p^2
+        \end{split}
+    \end{equation*}
+
+  % E[s-hat]
+    \begin{equation*}
+        \begin{split}
+        E[\hat{s}]& =E[A]-E[B] \\
+                  & =(n-1)\left((1-p)p-p^2\right)
+        \end{split}
+    \end{equation*}
 
 	The standard deviation...
-	\begin{align*}
-        Var[A] 
-            &= \sum_{i=1}^{n-1} Var[A] + 2\sum_{i=1}^{n-2} Cov[I_i = (0, 1),
-               I_{i+1} = (1, 1)] \\
-            &= (n-1)P((0, 1))(1-P((0, 1))) + 2\sum_{i=1}^{n-2}\bigg(E[I_i=(0, 
-               1), I_{i+1} = (0, 1)] - E[I_i = (0, 1)]E[I_{i+1} = (0, 1)]\bigg)
-               \\
-            &= (n-1)(1-p)p(1-(1-p)p) + 2\sum_{i=1}^{n-2}\bigg(0 - (1-p)p(1-p)p
+    % Var[A]
+    \begin{equation*}
+        \begin{split}
+        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[I_i=(0,1),I_{i+1}=
+                 (1,1)] \\
+              & =(n-1)P((0,1))(1-P((0,1))) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(0,1),I_{i+1}=(0,1)]-E[I_i=
+                (0,1)]E[I_{i+1}=(0,1)]\bigg) \\
+              & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
                 \bigg) \\
-            &= (1-p)p(-1+n+5p-3np-5p^2+3np^2)
-    \end{align*}
-    \begin{align*}
-        Var[B] 
-            &= \sum_{i=1}^{n-1} Var[B] + 2\sum_{i=1}^{n-2} Cov[I_i = (1, 1),
-               I_{i+1} = (1, 1)] \\
-            &= (n-1)P((1, 1))(1-P((1, 1))) + 2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,   
-               1), I_{i+1} = (1, 1)] - E[I_i = (1, 1)]E[I_{i+1} = (1, 1)]\bigg) 
-               \\
-            &= (n-1)p^2(1-p^2) + 2(n-2)(p^3-p^4) \\
-            &= (1-p)p^2(-1+n-5p+3np)
-    \end{align*}
-	\begin{align*}
-		2Cov[A, B]  
-			&= 2 \sum_{i=1}^{n-1} \sum_{j=1}^{n-1} Cov[A, B]\\
-			&= 2 \bigg(\sum_{i=1}^{n-2} Cov[AB] + \sum_{i=1}^{n-2} Cov[BA] +
-				\\ \sum_{i=1}^{n-1} Cov[X_i = 0 \text{ or } 1, X_{i+1} = 1]\bigg) \\
-			&= 2 \bigg(\sum_{i=1}^{n-2} E[AB] - E[A]E[B] + 
-				\sum_{i=1}^{n-2} E[BA] - E[B]E[A] + \\
-				\sum_{i=1}^{n-1} E[X_i = 0 \text{ or } 1, X_{i+1} = 1] - E[B]E[A] \\
-			&= 2(n-2)(p-1)^2p^2 - 2(n-2)p^3(1-p) - 2(n-1)p^3(1-p)			
-	\end{align*}	
-    \begin{align*}
-        Var[\hat{s}] 
-            &= Var[A-B] \\
-            &= Var[A] + Var[B] - 2Cov[A, B] \\
-            &= (1-p)p(-1+n+8p-4np-20p^2+12np^2)
-    \end{align*}
+              & =(1-p)p(-1+n+5p-3np-5p^2+3np^2)
+        \end{split}
+    \end{equation*}
+
+    % Var[B]
+    \begin{equation*}
+        \begin{split}
+        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[I_i=(1,1),I_{i+1}=
+                (1,1)] \\
+              & =(n-1)P((1, 1))(1-P((1, 1))) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,1),I_{i+1}=(1,1)]-E[I_i=
+                (1,1)]E[I_{i+1}=(1,1)]\bigg) \\
+              & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
+              & =(1-p)p^2(-1+n-5p+3np)
+        \end{split}
+    \end{equation*}
+
+    % Cov[A, B]
+    \begin{equation*}
+        \begin{split}
+        Cov[A,B]& =\sum_{i=1}^{n-1}\sum_{j=1}^{n-1}Cov[A,B] \\
+                & =\sum_{i=1}^{n-2}Cov[A,B]+\sum_{i=1}^{n-2}Cov[B,A] \\
+                & \quad +\sum_{i=1}^{n-1}Cov[X_i=0\text{ or }1,X_{i+1}=1] \\
+                & =\sum_{i=1}^{n-2}E[AB]-E[A]E[B]+\sum_{i=1}^{n-2}E[BA]-
+                  E[B]E[A] \\
+                & \quad +\sum_{i=1}^{n-1}E[X_i=0\text{ or }1,X_{i+1}=1]-
+                  E[B]E[A] \\
+                & =(n-2)(p-1)^2p^2-(n-2)p^3(1-p)-(n-1)p^3(1-p)
+        \end{split}
+    \end{equation*}
+
+    % Var[s-hat]
+    \begin{equation*}
+        \begin{split}
+        Var[\hat{s}]& =Var[A-B] \\
+                    & =Var[A]+Var[B]-2Cov[A,B] \\
+                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
+        \end{split}
+    \end{equation*}
 

### 2010-02-24 sarahDraft.tex
Draft

@@
 
-		Occupancy has two similar definitions.  It can be "the probability that a
+		Occupancy has two different definitions.  It can mean "the probability that a
 randomly selected site or sampling unit in an area of interest is occupied by a 
@@
 not the same as the abundance, or population size, of a species.  The number of
-members of a species is not important, only whether or not at least one member of
-the species is present in each site.
+members of a species is not important for either definition of occupancy.
 
@@
 estimate of the underlying probability.  Because of this, these two definitions
-of occupancy are oftentimes used interchangably.
+of occupancy are oftentimes used interchangeably.
 
@@
 into the way a system works.  It also can be used to monitor animals habitats 
-and discover trends over time in animal populations.
+and discover trends over time in groups of animals.
 
@@
  form of a presence-absence survey.  This means that observers visit sites and 
-spend time within each one looking for the animals themselves, tracks, 
-vocalizations, or some other indicator that the animal is present.  In some cases, 
-camera traps and sound recording are used instead of human observers.
+look for the animals themselves, tracks, vocalizations, or some other indicator 
+that the animal is present.  In some cases, camera traps and sound recording are 
+used instead of human observers. [MacKenzie 2006]
 
@@
 for an animal to occupy a site, but not be detected by the observer.  This means
-that the site is recorded as absent, but there really is an animal there.  This 
-is called a false negative.  Marking an animal present when there really is no
-animal is also a possible problem, but is thought to be much less likely and is
-often ignored.  We will discuss this more below.
+that the site is recorded as unoccupied, but there really is an animal there.  
+This is called a false negative.  Marking an animal present when there really is
+no animal is also a possible problem, but is thought to be much less likely and 
+is often ignored.  We will discuss this issue more below.
 
@@
 
-	There are some assumptions that are made in order to create models of 
-occupancy.  Without these assumptions, the models become much more complicated.
-One common assumption is no false positives, which means if a site is recorded as
-present, this cannot be incorrect.  This is an assumption that we attempted to 
-get rid of in our models.  A second common assumption is that presence and 
-detection probabilties are constant across sites and surveys.  A third common 
-assumption is that there is independence of detection between sites.  A fourth
-common assumption is that the surveys are being conducted on a closed population.
-This means that no animals leave the area or join the area during the time of the
-surveys.
+	There are some assumptions that are made in most models of occupancy.  [MacKenzie 2006].
+Without these assumptions, the models become much more complicated.  One 
+common assumption is no false positives, which means if a site is recorded as
+occupied we say it must actually be occupied.  A second common assumption is that
+presence and detection probabilities are constant across sites and surveys.  This 
+simplifies the estimation of each parameter.  A third common assumption is that 
+there is independence of detection between sites.  A fourth common assumption is 
+that the surveys are being conducted on a closed population.  This means that no 
+animals leave the area or join the area during the time of the surveys.
 
-	Obviously, not all of these assumptions are always true.  In some cases, there
-are established ways to deal with getting rid of an assumption in the models, 
-and in others there are no common ways to get rid of the assumption.  Either way, 
-it adds a level of complication to the model, and therefore we will only talk 
-about dealing with the assumption of no false positives, and independence of 
-detection across sites.
+	Obviously, not all of these assumptions are always true, but they are often 
+necessary in order to get a model that is simple enough to use.  There often are
+possible ways to get rid of these assumptions, but it complicates the model.  In
+this paper, we will only talk about attempting to get rid of the assumption of no
+false positives, and the assumption of independence of parameters across sites.
 
@@
 
-		All of the models that we looked at were single-species, single-season
-models because this is what fit our data, but these models can be modified to
-be used for multiple seasons or multiple species.
+		All of the models discussed in this paper are single-species, single-season
+models.  It is possible to modify all of these models to be used for multiple-seasons
+or multiple-species, but it would make them more complicated.
 
@@
 this estimate is too low because animals are not detected at all sites where they
-are present, which is the issue of false negatives.  Other occupancy models
-attempt to do better than this simple estimate.
+are present, which is the issue of false negatives mentioned above.  Other 
+occupancy models attempt to do better than this simple estimate.
 
@@
 or it could be there and just was not detected.  This could either be due to
-a false negative, where the observer did not see the indicator, or it could be 
-due to the fact that the animal does occupy the site but there was no indication
-of the animal at the site- for example, no track, no vocalization, etc.  \\
+a false negative, where the observer did not see the indication- for example they 
+did not see the animal, track, etc.- or it could be due to the fact that the 
+animal does occupy the site but there was no indication of the animal at the 
+site- for example, no track, no vocalization, etc.  \\
 Here is a visualization of how the Hierarchical model works:
@@
 
-From this diagram, you can see that we need separate parameters for the 
+From this diagram, it is apparent that we need separate parameters for the 
 probability that a site is occupied, the probability that presence is indicated
 given that a site is occupied, and the probability of detection given that 
-presence is indicated and the site is occupied.  We call these \(\psi\),
+presence is indicated and the site is occupied.  We will call these \(\psi\),
 \(\theta\), and \(p\), respectively.
@@
 	We calculated \(p\), \(q\), and \(r\) for every flight in our data, using 
-\(n = 500\).  We then realized that \(r\) does not answer the questions we are 
-asking.  It does not take into account whether or not a site is occupied, or 
+\(n = 500\).  We then realized that \(r\) is actually not so useful in analyzing
+our data.  It does not take into account whether or not a site is occupied, or 
 whether or not a track was laid down on that site, or any of the ideas behind    
@@
 find out something about, and the value of \(r\) cannot help us get there.  The 
-ideas behind using Markov Chains did end up being useful, however.
+idea to use Markov Chains was still useful, however.
 
 	Although we rejected the idea of using \(r\) as a way to analyze the data, 
-we realized that \(p\) and \(q\) could still be used to test the viability of 
-the independence assumption.  If there is independece between sites, the 
+we realized that \(p\) and \(q\) could be used to test the viability of 
+the independence assumption.  If there is independence between sites, the 
 difference of the proportions \(p\) and \(1-q\) should be approximately \(0\). 
@@
 the Whitewater and Zumbro Rivers looked similar to the Mississippi:  \\
-*INSERT BOXPLOT FROM STATUS REPORT HERE  \\
+*INSERT BOXPLOT FROM STATUS REPORT HERE OF p-(1-q) Mississippi \\
 We found no obvious trend indicating one site size to be more independent than 
@@
 
-	The mean ...
+In order to find \(E[\hat{s}]\), we first needed to find \(E[A]\) and \(E[B]\).
  	  % E[A]
@@
 
-	The standard deviation...
+The standard deviation was a little more difficult.  
     % Var[A]
@@
 "independent" than all sites as measured by this statistic.  However, alternating
-sites do not guarentee independence.  For instance, six of eleven flights along 
+sites do not guarantee independence.  For instance, six of eleven flights along 
 the Zumbro River during 2003 still have significant Z-scores with alternating
-sites.  There were no noticable trends for this statistic with respect to 
+sites.  There were no noticeable trends for this statistic with respect to 
 observer, days since snow, snow event, or plot size.
 
-*FIGURES OR TABLES??? WHICH ONES?
+*FIGURES OR TABLES??? WHICH ONES? from status report compilation
 
-	We did find interesting trends with regard to sampling method.  We think 
-that the lack of independence results from strings of ones where an otter left a
-track across multiple sites.  When alternating sites are used, the strings are 
-cut in half and the occurrence of \(1\),\(1\)'s is lower while the occurence of 
-\(0\),\(1\)'s stays about the same.  It is noteworthy that virtually all the 
-violations of independence are in the direction of more \(1\),\(1\)'s than would
-be expected.  This reinforces our thoughts about successive \(1\)'s representing
-the same otter.  We will discuss later whether this violation of independence 
-actually matters in our model, and how this might affect the estimation of 
-actual occupancy rates.
+	We think that the lack of independence results from strings of ones where an 
+otter left a track across multiple sites.  When alternating sites are used, the 
+strings are cut in half and the occurrence of \(1\),\(1\)'s is lower while the 
+occurrence of \(0\),\(1\)'s stays about the same.  It is noteworthy that 
+virtually all the violations of independence are in the direction of more 
+\(1\),\(1\)'s than would be expected.  This reinforces our thoughts about 
+successive \(1\)'s representing the same otter.  We will discuss later whether 
+this violation of independence actually matters in our model, and how this might
+affect the estimation of actual occupancy rates.
 
@@
 
+	This states that the posterior distribution \( \propto \) the likelihood 
+\(\times\) the prior.
 	Usually, the mean of the posterior distribution is then used as a point 
@@
 
-Therefore, the posterior distribution \( \propto \) the likelihood times the prior.
-    	Useful for hierarchical modeling: look at section 3.3 in book. \\
-    	Can use Markov Chain Monte Carlo (MCMC) to calculate posterior 
-			distribution- bimodal issues.
-
-	\subsection{BRugs/R} - Chrisna
+	We can use Bayesian statistics in the Hierarchical model as long as we set a
+prior for every parameter that we include.  We can then use Markov Chain Monte 
+Carlo (MCMC) to find a posterior distributions for every parameter.  We run into 
+a problem of duality with the probability of correct detection \(p\), and the 
+probability of false detection \(E\).  These two variables have bimodal posterior
+distributions, because \(p = x, E = y, \psi = z\) is the same as \(p = y, E = x, 
+\psi = 1 - z\).  Since we know that is reasonable for \(p\) to be greater than \(E\),
+we solved this problem by restricting the distribution of \(p\) to the range
+\(0.6 \text{ to } 1\), and \(E\) to \(0 \text{ to } 0.4\).
 

### 2010-02-24 kivaSections.tex

@@
 \documentclass[12pt]{article}
-\usepackage{graphicx,amssymb}
+\usepackage{graphicx,amssymb,amsmath}
 
@@
 \subsection{Descriptive statistics}
-
 In 2003, three different observers flew over the rivers. Assuming that no tracks 
@@
 identifying a track that does not exist ($E$). Including false positives 
-introduces a duality whereby $\begin{array}{ccc}(p = x, & E = y, & \psi = z)\end
-{array}$ is as good a solution as $\begin{array}{ccc}(p = y, & E = x, & \psi = 
-1-z)\end{array}$. To address this problem we assumed $p\gg0.5$ and $E\ll0.5$, an 
-approach recommended by Royle and Link (2006).
+introduces a duality whereby 
+$\begin{array}{ccc}(p = x, & E = y, & \psi = z)\end{array}$ 
+is as good a solution as 
+$\begin{array}{ccc}(p = y, & E = x, & \psi = 1-z)\end{array}$. 
+To address this problem we assumed $p\gg0.5$ and $E\ll0.5$, an approach 
+recommended by Royle and Link (2006).
 
 \section{Spatial correlation of the data}
-We found that there is significant correlation between the probability of 
-observing a track in neighboring sites, but this violates of a central 
-assumption of occupancy theory. This issue needed to be addressed. One approach 
-was to make the prior distribution of $\theta$, the probability that an otter 
-lays down a track, spatially correlated using a conditional autoregressive
-(CAR) model.
+One approach to dealing with the significant correlation in the probability of 
+observing a track in neighboring sites was to use a conditional autogregressive 
+(CAR) prior distribution for $\theta$, the probability that an otter lays down a 
+track in a site on a given day.
+
+\subsection{Conditional autoregressive (CAR) models}
+Let $\theta_s$ be the probability of a track being laid down in site $s$ on a 
+given day. Then the Gaussian CAR model can be defined using the following 
+conditional distributions:
+\begin{equation}\theta_s|\theta_i,\tau_s^2 \sim N(\mu_s+\sum_{i\in N_s}c_{si}
+(y_i-\mu_i),\tau_s^2)\end{equation}
+where $N_s$ is the set of neighbors of area $s$ (in this case, the two bordering 
+sites), $E(\theta_s)=\mu_s$, $\tau_s^2$ is the conditional variance, $c_{si}$ is 
+the weight defining how strong the correlation is between sites $s$ and $i$, and 
+$c_{ii}=0$ for all sites $i$ (Arab et al., year?). Essentially, this model means 
+that if site $s$'s neighbor has a particularly high or low track-laying 
+probability, that will increase or decrease the track-laying probability for 
+site $s$. This makes it more likely to have strings of tracks that cross over 
+multiple sites.
+
+\subsection{Our CAR model}\label{car model}
+For our model, both neighbors are weighted equally. Since $\tau_s$ for any site 
+$s$ is unknown, we followed convention and used the same value of $\tau$ for all 
+sites and placed a prior distribution on it. We chose a $\gamma(0.5,0.0005)$ 
+prior distribution for $(\tau^2)^{-1}$, a common choice (GeoBUGS manual). In our 
+model we define $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$ where $\alpha_0$ 
+has an improper uniform distribution and $\alpha_1$ has a Gaussian CAR 
+distribution. We again used a uniform prior distribution on $\psi$ and the same 
+beta prior distributions on $p$ and $E$. In this case, we limited the range of 
+$p$ to the interval $(0.7,1)$ and the range of $E$ to $(0,0.3)$. We increased 
+the burn-in time to 12,000 iterations. With this model, instead of a single 
+estimate for $\theta$ there are now $S$ estimates for $\theta$ where $S$ is the 
+total number of sites. While ideally we would estimate an overall value for $
+\theta$, it seems more important for the purposes of this investigation to 
+estimate $\psi$.
+
+\subsection{Simulating spatially correlated data}
+Because we modeled spatial correlation in $\theta$, we began by simulating 
+spatial correlation in the track-laying process. To do this, we added an 
+additional variable to the simulated data, $\alpha$. Under the new simulation 
+method, $\theta$ is the probability of an otter laying down a track in site $i$ 
+given site $i$ is occupied and site $i-1$ does \textit{not} have a track. If 
+site $i$ is occupied and site $i-1$ does have a track, then the probability of 
+an otter laying down a track in site $i$ is equal to $\theta+\alpha$. This means 
+if there is a track in one site, there is more likely to be a track in the next site. However, when we ran data simulated using this new method on the model 
+described in \ref{car model}, the model almost never converged. We tested 
+independence of the data using the $\hat{s}$ statistic we developed (section %
+\ref{shat}
+) and found that the simulated data displayed much more independence than the 
+actual data. We tried simulating spatial correlation only in $\psi$ with similar 
+results.
+
+To increase the correlation of the tracks we simulated data with spatial 
+correlation in both $\theta$ and $\psi$. Under this simulation method we again 
+had the variables $\theta$ and $\alpha$, but we also had analogous variables $
+\psi$ and $\beta$. In this case, $\psi$ is the probability an otter occupies 
+site $i$ given that site $i-1$ is not occupied and $\psi+\beta$ is the 
+probability that an otter occupies site $i$ given that site $i-1$ is occupied. 
+With this increased spatial correlation we were able to more closely match the $
+\hat{s}$ $z$-scores found in the actual data. In addition, the CAR model 
+converged more often with data simulated using this method, but it still did not 
+converge consistently.
+
+When we ran the data with spatial correlation in the occupancy process on the 
+CAR model, we could no longer compare the model estimate for $\psi$ to the 
+simulated value of $\psi$ because they had different meanings. To get a sense of 
+the overall occupancy rate in the simulated data we counted the number of sites 
+that were occupied (known because the data are simulated) and divided that by 
+the total number of sites. This proportion is the value we compared to the $\psi
+$ estimate from both hierarchical models.
 

### 2010-02-24 Paper.tex
First draft with (almost) everything combined.

--- 2010-02-24 ChrisnaSections.tex
@@
-\documentclass{article}
+% Paper.tex
+% Chrisna Aing, Sarah Halls, Kiva Oken
+% OTTER COMPS!
+
+\documentclass[12pt]{article}
 \usepackage{amsmath}
-\usepackage[hmargin=3cm, vmargin=3cm]{geometry}
+\usepackage{amssymb}
+\usepackage{graphicx}
 \usepackage{multirow}
 \usepackage[dvipsnames, table]{xcolor}
+\usepackage[all]{xy}
 
@@
 
+% This should be a lot prettier.
+\title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
+Population in Southeastern Minnesota}
+\date{}
+\author{Chrisna Aing, Sarah Halls, Kiva Oken}
+\maketitle
+
+\section{Overview of Occupancy Modeling}
+
+    \subsection{What is Occupancy Rate?}
+    Occupancy has two different definitions. It can mean ``the probability that
+    a randomly selected site or sampling unit in an area of interest is occupied
+    by a species'' \cite{MacKenzie2006}. It also can be used to describe ``the
+    proportion of area, patches, or sample units that is occupied''
+    \cite{MacKenzie2006}. Occupancy is not the same as the abundance, or
+    population size, of a species. The number of members of a species is not
+    important for either definition of occupancy.
+
+    While there is a difference between these two definitions of occupancy, in
+    general the probability is unknown and the observed proportion is used as an
+    estimate of the underlying probability. Because of this, these two
+    definitions of occupancy are oftentimes used interchangeably.
+
+    The main reasons that people are interested in measuring occupancy are
+    science or conservation and management \cite{MacKenzie2006}. This statistic
+    can give insight into the way a system works. It also can be used to monitor
+    animals habitats and discover trends over time in groups of animals.
+
+    \subsection{Different Ways of Sampling Data}
+    In order to determine occupancy rate, the data is often collected in the
+    form of a presence-absence survey. This means that observers visit sites and
+    look for the animals themselves, tracks, vocalizations, or some other
+    indicator that the animal is present. In some cases, camera traps and sound
+    recording are used instead of human observers \cite{MacKenzie2006}.
+
+    The main problem with these methods of sampling data is that it is possible
+    for an animal to occupy a site, but not be detected by the observer. This
+    means that the site is recorded as unoccupied, but there really is an animal
+    there. This is called a false negative. Marking an animal present when there
+    really is no animal is also a possible problem, but is thought to be much
+    less likely and is often ignored. We will discuss this issue more below.
+
+    \subsection{Assumptions}
+    There are some assumptions that are made in most models of occupancy.
+    \cite{MacKenzie2006}. Without these assumptions, the models become much more
+    complicated. One common assumption is no false positives, which means if a
+    site is recorded as occupied we say it must actually be occupied. A second
+    common assumption is that presence and detection probabilities are constant
+    across sites and surveys. This simplifies the estimation of each parameter.
+    A third common assumption is that there is independence of detection between
+    sites. A fourth common assumption is that the surveys are being conducted on
+    a closed population. This means that no animals leave the area or join the
+    area during the time of the surveys.
+
+    Obviously, not all of these assumptions are always true, but they are often
+    necessary in order to get a model that is simple enough to use. There often
+    are possible ways to get rid of these assumptions, but it complicates the
+    model. In this paper, we will only talk about attempting to get rid of the
+    assumption of no false positives, and the assumption of independence of
+    parameters across sites.
+
+    \subsection{Types of Occupancy Models}
+    All of the models discussed in this paper are single-species, single-season
+    models. It is possible to modify all of these models to be used for multiple
+    seasons or multiple species, but it would make them more complicated.
+
+    The most simple way to measure occupancy is to simply divide the number of
+    sites where an animal was found to be present by the total number of sites.
+    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
+    accepted that this estimate is too low because animals are not detected at
+    all sites where they are present, which is the issue of false negatives
+    mentioned above. Other occupancy models attempt to do better than this
+    simple estimate.
+
+    One way to create more accurate models is to use a Hierarchical model as a
+    way to think about the parameters that affect occupancy. This allows us to
+    take into account that when an animal is not detected, it could either not
+    be there, or it could be there and just was not detected. This could either
+    be due to a false negative, where the observer did not see the indication --
+    for example they did not see the animal, track, etc. -- or it could be due
+    to the fact that the animal does occupy the site but there was no indication
+    of the animal at the site -- for example, no track, no vocalization, etc.
+
+    Here is a visualization of how the Hierarchical model works:
+    % Either the margins need to be increased or we need a prettier diagram :(.
+    \begin{displaymath}
+        \xymatrix{
+        & & & \text{Site}\ar[dl]\ar[dr] & \\
+        & & \text{Occupied}\ar[dl]\ar[d] & & \text{Unoccupied}\ar[d] \\
+        & \text{Presence Indicated}\ar[dl]\ar[d] & 
+          \text{Presence Not Indicated}\ar[d] & & \text{Not Detected} \\
+        \text{Detected} & \text{Not Detected} & \text{Not Detected} & &}
+    \end{displaymath}
+
+    From this diagram, it is apparent that we need separate parameters for the
+    probability that a site is occupied, the probability that presence is
+    indicated given that a site is occupied, and the probability of detection
+    given that presence is indicated and the site is occupied. We will call
+    these \(\psi\), \(\theta\), and \(p\), respectively.
+
+\section{Questions of investigation}
+The North American river otter (\textit{Lontra canadensis}) is endemic to all of
+Minnesota; however, populations declined precipitously in the 19th and 20th
+centuries due to trapping and habitat degradation. In recent years their
+populations have significantly rebounded, especially in southeastern Minnesota
+where hunters would like to be able to trap otters again. In order for that to
+be possible governmental organizations must be able to monitor the population
+health. In addition, otter presence can indicate high quality aquatic habitat.
+Due to these reasons, there is interest in monitoring changes in otter abundance
+over time. Unfortunately, otters are not easy to observe. Thus, this study
+investigated the potential of using wintertime aerial surveys of otter tracks on
+rivers to make inferences about the abundance of otters and how it changes over
+time. We used the concepts from occupancy theory to develop models to estimate
+abundance.
+
+    \subsection{Data collection}
+    During the winters of 2003 and 2004 people flew over the Whitewater, Zumbro,
+    and Mississippi Rivers. When they saw what looked like an otter track, they
+    pushed a button recording the GPS coordinates. They pushed the button every
+    five seconds until the track ended. Flights occurred one, two, or three days
+    after a snow event. In some cases, multiple observers flew over the same
+    river on the same day. Surveys were conducted during multiple snow events
+    throughout both winters. The rivers were then divided into 400 and 800 meter
+    plots and the presence (1) or absence (0) of a waypoint in a plot on a given
+    flight was determined. The Mississippi River was also divided into 1600
+    meter plots.
+
+    \subsection{Issues related to the data}
+    One major issue with respect to the data is that if a track occurred near
+    the boundary between two plots, a person on one flight may have recorded the
+    track in one site whereas someone on a later flight during that same snow
+    event may have seen the same track but recorded it in the neighboring plot.
+    So while the data indicate that the observers on the two flights disagreed
+    in two plots, they saw the same track. Increasing the plot size should
+    decrease the magnitude of this effect.
+
+    In addition, there is concern that the probability of an otter laying down a
+    track in one site is not independent of the probability that an otter lays
+    down a track in neighboring sites, violating a principle assumption of
+    occupancy theory. That is, a single track can extend across multiple plots.
+    This leads to dependence among the plots. It was important to quantify this
+    dependence and find an optimum way to handle it.
+
+    \subsection{Descriptive statistics}
+    In 2003, three different observers flew over the rivers. Assuming that no
+    tracks are laid down between individuals' flights on the same day, all three
+    people should have seen the same tracks. While there is some error in
+    classifying which site the observer saw the actual track in, each observer
+    should at least be seeing the same, or nearly the same, number of sites with
+    tracks. However, we did not find that to be the case. If we look only at
+    days that all three people observed, there appear to be major differences
+    among the proportion of sites in which the scientists detected a track
+    (Figure x). The difference in estimated na\"ive occupancy among the
+    observers indicates that the scientists did not always correctly identify
+    the tracks.
+
+    Correlation among the flight sequences is low. For example, average
+    correlation between flights on the Mississippi in 2003 is 0.176 for 400
+    meter plots. Some of this is due to the difficulties in determining to which
+    site each track belongs. However, we have also found quite a bit of
+    variation in the data not necessarily due to these classification issues.
+    This is difficult to quantify, as an algorithm for finding classification
+    errors would be complicated. Correlation between flights is a bit higher for
+    flights on the same day or flights during the same snow event, but still not
+    as high as one might hope. The proportion of sites with tracks detected
+    generally increases with days since snow, likely because the otters have a
+    longer time to lay down a track as time after the snow event passes.
+
+    \subsection{False positive rates}
+    The lack of consistency among the flights is strong evidence that the data
+    contain false positives where scientists think they see a track when it does
+    not exist. Not accounting for such a process can lead to major biases; as
+    the number of flights increases, the occupancy rate estimate will approach
+    one (Royle and Link 2006). We introduced a false detection process and used
+    the observer as a covariate (by means of indicator variables) for both the
+    probability of correctly identifying a track that exists ($p$) and the
+    probability of identifying a track that does not exist ($E$). Including
+    false positives introduces a duality whereby $(p=x,E=y,\psi=z)$ is as good a
+    solution as $(p=y,E=x,\psi=1-z)$. To address this problem we assumed
+    $p\gg0.5$ and $E\ll0.5$, an approach recommended by Royle and Link (2006).
+
+    \subsection{Markov Chains and Testing the Independence Assumption}
+    \label{shat}
+    First we considered using Markov Chains to come up with a long term
+    probability of detection, which we called \(r\). We let \(p\) be the number
+    of \(0\)'s followed by a \(0\) as a fraction of all successive pairs that
+    start with a \(0\). We let \(q\) be the number of \(1\)'s followed by a
+    \(1\) as a fraction of all successive pairs that start with a \(1\). These
+    probabilities can be represented by the matrix:
+    \[M=\begin{bmatrix}
+        p & 1-p \\
+        1-q & q
+    \end{bmatrix}\text{.}\]
+
+    The long term probability of detection can be found from this by finding
+    \(\lim_{n \to \inf}M^n\), which returns
+    \[\begin{bmatrix}
+        1-r & r \\
+        1-r & r
+    \end{bmatrix}\text{,}\]
+    where \(r\) is the probability of a \(1\), which is the long term
+    probability of detection.
+
+    We calculated \(p\), \(q\), and \(r\) for every flight in our data, using
+    \(n = 500\). We then realized that \(r\) is actually not so useful in
+    analyzing our data. It does not take into account whether or not a site is
+    occupied, or whether or not a track was laid down on that site, or any of
+    the ideas behind the hierarchical model. These are the parameters that we
+    eventually want to find out something about, and the value of \(r\) cannot
+    help us get there. The idea to use Markov Chains was still useful, however.
+
+    Although we rejected the idea of using \(r\) as a way to analyze the data,
+    we realized that \(p\) and \(q\) could be used to test the viability of the
+    independence assumption. If there is independence between sites, the
+    difference of the proportions \(p\) and \(1-q\) should be approximately
+    \(0\). This would mean that the probability that a \(0\) follows a \(0\) is
+    about the same as the probability that a \(1\) follows a \(0\), meaning that
+    it does not depend on what the observation is at the previous site. Looking
+    at boxplots of \(p-(1-q)\), we see that data from flights that record
+    alternating sites may be more independent than data from flights that record
+    all sites, which suggests a possible benefit to using the alternating
+    sampling method. The boxplots for the Whitewater and Zumbro Rivers looked
+    similar to the Mississippi:
+
+    % INSERT BOXPLOT FROM STATUS REPORT HERE OF p-(1-q) Mississippi
+    We found no obvious trend indicating one site size to be more independent
+    than the others.
+
+    While this is an interesting observation, we would like to be able to
+    quantify it better by finding the distribution of our statistic. This would
+    allow us to make statements about how independent the data is, and if the
+    difference in independence between different types of data is significant.
+    This statistic seems hard to get, but we considered a modification. Let
+    \(S\) be the number of successive \(0\), \(1\)'s minus the number of
+    successive \(1\), \(1\)'s per flight. Under the null hypothesis that each
+    site is an independent Bernoulli trial with probability \(p\) we were able
+    to compute the mean and standard deviation of \(S\), which we used to
+    calculate a Z-score. In order to compute these values, we needed to find the
+    expected value and variance of each term in \(S\). We let \(A = \text{\# of
+    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).
+
+    In order to find \(E[\hat{s}]\), we first needed to find \(E[A]\) and
+    \(E[B]\).
+
+    % E[A]
+    \begin{equation*}
+        \begin{split}
+        E[A]& =\sum_{i=1}^{n-1}P(X_i=0,X_{i+1}=1) \\
+            & =\sum_{i=1}^{n-1}(1-p)p \\
+            & =(n-1)(1-p)p \\
+        \end{split}
+    \end{equation*}
+
+    % E[B]
+    \begin{equation*}
+        \begin{split}
+        E[B]& =\sum_{i=1}^{n-1}P(X_i=1,X_{i+1}=1) \\
+            & =\sum_{i=1}^{n-1}p*p \\
+            & =(n-1)p^2
+        \end{split}
+    \end{equation*}
+
+    % E[s-hat]
+    \begin{equation*}
+        \begin{split}
+        E[\hat{s}]& =E[A]-E[B] \\
+                  & =(n-1)\left((1-p)p-p^2\right)
+        \end{split}
+    \end{equation*}
+
+    The standard deviation was a little more difficult.
+
+    % Var[A]
+    \begin{equation*}
+        \begin{split}
+        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[I_i=(0,1),I_{i+1}=
+                 (1,1)] \\
+              & =(n-1)P((0,1))(1-P((0,1))) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(0,1),I_{i+1}=(0,1)]-E[I_i=
+                (0,1)]E[I_{i+1}=(0,1)]\bigg) \\
+              & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
+                \bigg) \\
+              & =(1-p)p(-1+n+5p-3np-5p^2+3np^2)
+        \end{split}
+    \end{equation*}
+
+    % Var[B]
+    \begin{equation*}
+        \begin{split}
+        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[I_i=(1,1),I_{i+1}=
+                (1,1)] \\
+              & =(n-1)P((1, 1))(1-P((1, 1))) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,1),I_{i+1}=(1,1)]-E[I_i=
+                (1,1)]E[I_{i+1}=(1,1)]\bigg) \\
+              & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
+              & =(1-p)p^2(-1+n-5p+3np)
+        \end{split}
+    \end{equation*}
+
+    % Cov[A, B]
+    \begin{equation*}
+        \begin{split}
+        Cov[A,B]& =\sum_{i=1}^{n-1}\sum_{j=1}^{n-1}Cov[A,B] \\
+                & =\sum_{i=1}^{n-2}Cov[A,B]+\sum_{i=1}^{n-2}Cov[B,A] \\
+                & \quad +\sum_{i=1}^{n-1}Cov[X_i=0\text{ or }1,X_{i+1}=1] \\
+                & =\sum_{i=1}^{n-2}E[AB]-E[A]E[B]+\sum_{i=1}^{n-2}E[BA]-
+                  E[B]E[A] \\
+                & \quad +\sum_{i=1}^{n-1}E[X_i=0\text{ or }1,X_{i+1}=1]-
+                  E[B]E[A] \\
+                & =(n-2)(p-1)^2p^2-(n-2)p^3(1-p)-(n-1)p^3(1-p)
+        \end{split}
+    \end{equation*}
+
+    % Var[s-hat]
+    \begin{equation*}
+        \begin{split}
+        Var[\hat{s}]& =Var[A-B] \\
+                    & =Var[A]+Var[B]-2Cov[A,B] \\
+                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
+        \end{split}
+    \end{equation*}
+
+    Simulations of \(S\) show that the distribution of \(S\) is roughly normal.
+    So using our calculations for the mean and the standard deviation, we
+    computed a Z-score for each flight, where we use a plug-in estimate for
+    \(p\) to be the proportion of \(1\)'s per flight. We said that Z-scores with
+    an absolute value greater than \(2\) suggest that the assumption of
+    independence between sites is violated for that flight. We found that, in
+    general, alternating sites are more "independent" than all sites as measured
+    by this statistic. However, alternating sites do not guarantee independence.
+    For instance, six of eleven flights along the Zumbro River during 2003 still
+    have significant Z-scores with alternating sites. There were no noticeable
+    trends for this statistic with respect to observer, days since snow, snow
+    event, or plot size.
+
+    % FIGURES OR TABLES??? WHICH ONES? from status report compilation
+
+    We think that the lack of independence results from strings of ones where an
+    otter left a track across multiple sites. When alternating sites are used,
+    the strings are cut in half and the occurrence of \(1\), \(1\)'s is lower
+    while the occurrence of \(0\), \(1\)'s stays about the same. It is
+    noteworthy that virtually all the violations of independence are in the
+    direction of more \(1\), \(1\)'s than would be expected. This reinforces our
+    thoughts about successive \(1\)'s representing the same otter. We will
+    discuss later whether this violation of independence actually matters in our
+    model, and how this might affect the estimation of actual occupancy rates.
+
+\section{Theoretical Stuff}
+
+    \subsection{Bayesian Statistics}
+    Bayes' Theorem: \(P(\theta|X) = P(X|\theta)P(\theta) / P(X)\) \\
+
+    This states that the posterior distribution \(\propto\) the likelihood
+    \(\times\) the prior. Usually, the mean of the posterior distribution is
+    then used as a point estimate for the parameter.
+
+    We can use Bayesian statistics in the Hierarchical model as long as we set a
+    prior for every parameter that we include. We can then use Markov Chain
+    Monte Carlo (MCMC) to find a posterior distributions for every parameter. We
+    run into a problem of duality with the probability of correct detection
+    \(p\), and the probability of false detection \(E\). These two variables
+    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is the same
+    as \(p=y,E=x,\psi=1-z\). Since we know that is reasonable for \(p\) to be
+    greater than \(E\), we solved this problem by restricting the distribution
+    of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to \(0\text{ to }0.4\).
+
+    \subsection{WinBUGS/R}
+    To be filled in...
+
 \section{Standard Hierarchical Model}
-    \subsection{Model Parameters}
-    \subsection{Simulated Data}
-    \noindent{\begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{4}{|l|}{\textbf{The Model vs. Simulated Data}} \\
-        \hline
-            \(\psi\) & \(\theta\) & Convergence & 95\% CI (\(\psi\)) \\
-        \hline
-        \multirow{5}{*}{0.20}
-            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.112, 0.305) \\
-            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.142, 0.341) \\
-            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.098, 0.284) \\
-            & \cellcolor{Green}0.80 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.097, 0.281) \\
-            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
-              \cellcolor{Yellow}(0.117, 0.302) \\
-        \hline
-        \multirow{5}{*}{0.40}
-            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Okay &
-              \cellcolor{Yellow}(0.388, 0.721) \\
-            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.270, 0.494) \\
-            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.287, 0.512) \\
-            & \cellcolor{Red}0.80 & \cellcolor{Red}Bad &
-              \cellcolor{Red}(0.277, 0.507) \\
-            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
-              \cellcolor{Yellow}(0.279, 0.503) \\
-        \hline
-        \multirow{5}{*}{0.60}
-            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
-              \cellcolor{Yellow}(0.312, 0.581) \\
-            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.400, 0.630) \\
-            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
-              \cellcolor{Yellow}(0.462, 0.687) \\
-            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
-              \cellcolor{Yellow}(0.502, 0.727) \\
-            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
-              \cellcolor{Red}(0.318, 0.549) \\
-        \hline
-        \multirow{5}{*}{0.80}
-            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.578, 0.838) \\
-            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.736, 0.915) \\
-            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
-              \cellcolor{Yellow}(0.587, 0.829) \\
-            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
-              \cellcolor{Yellow}(0.599, 0.811) \\
-            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
-              \cellcolor{Red}(0.612, 0.830) \\
-        \hline
-        \multirow{5}{*}{1.00}
-            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
-              \cellcolor{Yellow}(0.895, 0.999) \\
-            & \cellcolor{Yellow}0.40 & \cellcolor{Yellow}Okay &
-              \cellcolor{Yellow}(0.927, 0.999) \\
-            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
-              \cellcolor{Green}(0.949, 1.000) \\
-            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Good &
-              \cellcolor{Yellow}(0.802, 0.954) \\
-            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
-              \cellcolor{Red}(0.800, 0.955) \\
-        \hline
-    \end{tabular}}
+Also to be filled in... :(.
+
+\section{Spatial correlation of the data}
+One approach to dealing with the significant correlation in the probability of
+observing a track in neighboring sites was to use a conditional autoregressive
+(CAR) prior distribution for $\theta$, the probability that an otter lays down a
+track in a site on a given day.
+
+    \subsection{Conditional autoregressive (CAR) models}
+    Let $\theta_s$ be the probability of a track being laid down in site $s$ on
+    a given day. Then the Gaussian CAR model can be defined using the following
+    conditional distributions:
+    \begin{equation}
+        \theta_s|\theta_i,\tau_s^2 \sim N(\mu_s+\sum_{i\in N_s}
+        c_{si}(y_i-\mu_i),\tau_s^2)
+    \end{equation}
+    where $N_s$ is the set of neighbors of area $s$ (in this case, the two
+    bordering sites), $E(\theta_s)=\mu_s$, $\tau_s^2$ is the conditional
+    variance, $c_{si}$ is the weight defining how strong the correlation is
+    between sites $s$ and $i$, and $c_{ii}=0$ for all sites $i$ (Arab et al.,
+    year?). Essentially, this model means that if site $s$'s neighbor has a
+    particularly high or low track-laying probability, that will increase or
+    decrease the track-laying probability for site $s$. This makes it more
+    likely to have strings of tracks that cross over multiple sites.
+
+    \subsection{Our CAR model}
+    \label{car model}
+    For our model, both neighbors are weighted equally. Since $\tau_s$ for any
+    site $s$ is unknown, we followed convention and used the same value of
+    $\tau$ for all sites and placed a prior distribution on it. We chose a
+    $\gamma(0.5,0.0005)$ prior distribution for $(\tau^2)^{-1}$, a common choice
+    (GeoBUGS manual). In our model we define $\text{logit}(\theta_s)=\alpha_0+
+    \alpha_{1_s}$ where $\alpha_0$ has an improper uniform distribution and
+    $\alpha_1$ has a Gaussian CAR distribution. We again used a uniform prior
+    distribution on $\psi$ and the same beta prior distributions on $p$ and $E$.
+    In this case, we limited the range of $p$ to the interval $(0.7,1)$ and the
+    range of $E$ to $(0,0.3)$. We increased the burn-in time to 12,000
+    iterations. With this model, instead of a single estimate for $\theta$ there
+    are now $S$ estimates for $\theta$ where $S$ is the total number of sites.
+    While ideally we would estimate an overall value for $\theta$, it seems more
+    important for the purposes of this investigation to estimate $\psi$.
+
+    \subsection{Simulating spatially correlated data}
+    Because we modeled spatial correlation in $\theta$, we began by simulating
+    spatial correlation in the track-laying process. To do this, we added an
+    additional variable to the simulated data, $\alpha$. Under the new
+    simulation method, $\theta$ is the probability of an otter laying down a
+    track in site $i$ given site $i$ is occupied and site $i-1$ does
+    \textit{not} have a track. If site $i$ is occupied and site $i-1$ does have
+    a track, then the probability of an otter laying down a track in site $i$ is
+    equal to $\theta+\alpha$. This means if there is a track in one site, there
+    is more likely to be a track in the next site. However, when we ran data
+    simulated using this new method on the model described in section \ref{car
+    model}, the model almost never converged. We tested independence of the data
+    using the $\hat{s}$ statistic we developed in section \ref{shat} and found
+    that the simulated data displayed much more independence than the actual
+    data. We tried simulating spatial correlation only in $\psi$ with similar
+    results.
+
+    To increase the correlation of the tracks we simulated data with spatial
+    correlation in both $\theta$ and $\psi$. Under this simulation method we
+    again had the variables $\theta$ and $\alpha$, but we also had analogous
+    variables $ \psi$ and $\beta$. In this case, $\psi$ is the probability an
+    otter occupies site $i$ given that site $i-1$ is not occupied and $\psi+
+    \beta$ is the probability that an otter occupies site $i$ given that site
+    $i-1$ is occupied. With this increased spatial correlation we were able to
+    more closely match the $\hat{s}$ $z$-scores found in the actual data. In
+    addition, the CAR model converged more often with data simulated using this
+    method, but it still did not converge consistently.
+
+    When we ran the data with spatial correlation in the occupancy process on
+    the CAR model, we could no longer compare the model estimate for $\psi$ to
+    the simulated value of $\psi$ because they had different meanings. To get a
+    sense of the overall occupancy rate in the simulated data we counted the
+    number of sites that were occupied (known because the data are simulated)
+    and divided that by the total number of sites. This proportion is the value
+    we compared to the $\psi$ estimate from both hierarchical models.
+
+    Figure x caption:
+    The box plots show how the proportion of sites in which an observer saw
+    tracks (``occupancy'') differs among the three observers. Only those dates
+    in which all three observers flew are included.
 
 \section{Discussion}
+
     \subsection{Obtaining Convergence on the Actual Data}
@@
     our simulated data, we moved onto running both models on the data given to
-    us by the Minnesota DNR.  Recall that data was gathered during the winters
-    of 2003 and 2004 along the Mississippi, Whitewater, and Zumbro Rivers and
+    us by the Minnesota DNR. Recall that data was gathered during the winters of
+    2003 and 2004 along the Mississippi, Whitewater, and Zumbro Rivers and
     aggregated with ArcGIS into vectors containing ones and zeroes representing
@@
     important factors are plot size, the number of flights recorded, and the
-    number of estimated parameters.  In general, both plot size and the number
-    of flights recorded must be sufficiently small and large, respectively, for
-    the standard model with \(E\) to converge.
+    number of estimated parameters. In general, both plot size and the number of
+    flights recorded must be sufficiently small and large, respectively, for the
+    standard model with \(E\) to converge.
 
@@
     consistent with the noticeable but modest increase in \(\theta\) across plot
-    % I have no idea how to explain this... :(.
-    sizes for all three rivers.  However, estimates of \(\psi\) along the
-    Whitewater and Zumbro Rivers remain constant, despite large increases in
-    \(\theta\).
+    % I have no idea how to explain this... :(. sizes for all three rivers.
+    However, estimates of \(\psi\) along the Whitewater and Zumbro Rivers remain
+    constant, despite large increases in \(\theta\).
 
@@
     Similarly, the increases in estimates of \(p\) and \(E\) seem to result from
-    the general increase of tracks laid per plot as the plot size grows.  As
-    plot size grows, the influence of a 1 increases, regardless of whether an
-    otter track or a track from a different animal was observed.
+    the general increase of tracks laid per plot as the plot size grows. As plot
+    size grows, the influence of a 1 increases, regardless of whether an otter
+    track or a track from a different animal was observed.
 
@@
 
-    This data converged well.  However, as is the case with all of the 2004
-    data, we feel that the absence of \(E\) from the model makes the estimates
-    much less dependable.
+    This data converged well. However, as is the case with all of the 2004 data,
+    we feel that the absence of \(E\) from the model makes the estimates much
+    less dependable.
 
     \subsection{Covariates}
-    To determine the relative importance of our covariates (observer,
-    date, and number of days since the most recent snow event), we ran the
-    standard hierarchical model with \(E\) on simulated data with values of
-    \(\psi\) and \(\theta\) similar to our estimates of \(\psi\) and \(\theta\)
-    on 2003's data with varying numbers of observers, snow events, and
-    observations after each snow event.  We assume that observations after each
-    snow event will occur on consecutive days, starting from the first day after
-    the snow event.  This information helped us to find the most efficient ways
-    (with respect to number of flights) to gather aerial occupancy data given
-    one, two, and three observers while still obtaining useful estimates from
-    the model.
+    To determine the relative importance of our covariates (observer, date, and
+    number of days since the most recent snow event), we ran the standard
+    hierarchical model with \(E\) on simulated data with values of \(\psi\) and
+    \(\theta\) similar to our estimates of \(\psi\) and \(\theta\) on 2003's
+    data with varying numbers of observers, snow events, and observations after
+    each snow event. We assume that observations after each snow event will
+    occur on consecutive days, starting from the first day after the snow event.
+    This information helped us to find the most efficient ways (with respect to
+    number of flights) to gather aerial occupancy data given one, two, and three
+    observers while still obtaining useful estimates from the model.
 
@@
     this is not surprising. Two observations after each snow event on
-    consecutive days are required.  Additional observations after each event
-    help the model converge more quickly, but the estimates become only slightly
-    more accurate.
+    consecutive days are required. Additional observations after each event help
+    the model converge more quickly, but the estimates become only slightly more
+    accurate.
 
@@
     days' worth of flights per snow event would be sufficient. If the same
-    observer flies twice (or more often) in one day, perhaps they could act as
-    a substitute for an additional observer.  This is because it is probable
-    that, with one flight's worth of experience observing the given river, the
+    observer flies twice (or more often) in one day, perhaps they could act as a
+    substitute for an additional observer. This is because it is probable that,
+    with one flight's worth of experience observing the given river, the
     observer's \(p\) and \(E\) would change due to their newfound familiarity
@@
 \bibliography{Bibliography}
+
 \end{document}

### 2010-02-24 ChrisnaSections.tex
Making my section look much longer.
Added stub for WinBugs/R discussion.
Removed all stuff moved to Paper.tex.

--- 2010-02-24 ChrisnaSections.tex
@@
-\documentclass{article}
+\documentclass[12pt]{article}
 \usepackage{amsmath}
@@
 \begin{document}
+\section{Theory}
+    \subsection{Computer Help}
+    All of our statistical analysis was performed in R.
 
-%%% *STANDARD HIERARCHICAL MODEL*:
+    ALl of our modeling was performed using BRugs, which allowed us to use
+    WinBUGS commands and other stuff in R.
+
 \section{Standard Hierarchical Model}
-
-    %% MODEL PARAMETERS:
     \subsection{Model Parameters}
-
-    %% SIMULATED DATA, PT. 1:
     \subsection{Simulated Data}
-
     \noindent{\begin{tabular}{|l|l|l|l|}
@@
 
-%%% *DISCUSSION*:
-\section{Discussion}
-
-    %% GETTING THE ACTUAL DATA TO CONVERGE:
-    \subsection{Obtaining Convergence on the Actual Data}
-    Recall that our standard hierarchical model contains the following
-    parameters: \(\psi\) (the rate of occupancy), \(\theta\) (the probability of
-    laying down a track given occupancy), \(p\) (the probability of detecting a
-    track given that it is present, which is unique to each observer and makes
-    \(1-p\) the false negative rate of detecting a track), and \(E\) (the false
-    positive rate of detecting a track, which is unique to each observer).
-
-    Once both the standard hierarchical model and the addition of a CAR prior
-    distribution on \(\theta\) to the standard model exhibited convergence on
-    our simulated data, we moved onto running both models on the data given to
-    us by the Minnesota DNR.  Recall that data was gathered during the winters
-    of 2003 and 2004 along the Mississippi, Whitewater, and Zumbro Rivers and
-    aggregated with ArcGIS into vectors containing ones and zeroes representing
-    the detection and non-detection of a track, respectively, within plots of
-    length 402 meters, 804 meters, and 1608 meters.  The covariates recorded
-    were: the observer, the date, and the number of days since the last snow
-    event.  In general, the sparseness of the data gathered in 2004 caused us to
-    simplify the model in order to obtain results.  Unfortunately, the
-    simplification results in estimates of \(\psi\) that do not seem accurate
-    and are not directly comparable to estimates of \(\psi\) on 2003's data.
-
-    The standard model converged reliably on all data gathered in 2003, but it
-    did not converge on any of the data gathered in 2004.  Considering that the
-    amount of data gathered in 2004 pales in comparison to the amount of data
-    gathered in 2003, this result was not surprising.  The vast majority of
-    2004's data consists of just one observer taking observations once after
-    each snow event, which causes \(p\) and \(E\) in particular to not converge,
-    regardless of plot size.
-
-    To obtain convergence on 2004's relatively sparse data, we removed \(E\),
-    which adds to the model the assumption that there are no false positives
-    and, consequently, simplifies it greatly.  Although the model still did not
-    converge on data with plot sizes of 804m and 1608m, both the Whitewater and
-    Zumbro Rivers' data at the 402m plot size converged.  Although none of the
-    Mississippi River's 2004 data converged, we still include estimates of
-    \(p\), \(\psi\), and \(\theta\) at the 402m plot size to accompany the
-    estimates corresponding to the Whitewater and Zumbro Rivers.  For the most
-    part, however, we feel that the absence of \(E\) from the model makes the
-    the estimates of \(\psi\) much less dependable.
-
-    The addition of a CAR prior distribution on \(\theta\) to the standard model
-    caused the model to fail to converge on all of the data.  Although the
-    standard model and the CAR model performed similarly on simulated data
-    whenever both models converged, the standard model converged much more often
-    than did the CAR model.  This is most likely due to the fact that the CAR
-    model attempts to estimate too many parameters.  Whereas the standard model
-    estimates between eight and twelve parameters on the actual data (\(\psi\),
-    \(\theta\), the four parameters of the beta distributions corresponding to
-    \(p\) and \(E\), \(p\) for each observer, and \(E\) for each observer), the
-    CAR model estimates \(\theta\) for each site.  Because there are at least
-    thirty-five sites in each data set (regardless of plot size), the CAR prior
-    distribution adds a considerable amount of complexity to the model.
-
-    We also tried to simplify this model by eliminating \(E\) because this was
-    somewhat successful with respect to convergence with the standard model.  We
-    first attempted to fit this model to data from 2003 at the 402m plot size.
-    Although the model successfully converged on all of this data, its estimates
-    of \(\psi\), all of which were greater than 0.9, did not seem reasonable,
-    especially in comparison to the standard model's estimates when given the
-    same data.  As a result, we did not attempt to fit this model to the rest of
-    the data and do not include this model's estimates within our results.
-
-    %% RESULTS:
-    \subsection{Results}
-    Below, in tables, lie our estimates for all of the Minnesota DNR's data from
-    2003 and 2004.  Below each table is detailed commentary about the ability of
-    the standard hierarchical model to converge, given the corresponding data.
-
-    With respect to the model's ability to converge on the actual data, the most
-    important factors are plot size, the number of flights recorded, and the
-    number of estimated parameters.  In general, both plot size and the number
-    of flights recorded must be sufficiently small and large, respectively, for
-    the standard model with \(E\) to converge.
-
-    2003's data worked relatively well with the model.  The presence of numerous
-    observations that occur in the second and third days since the most recent
-    snow event allows one of the more crucial assumptions of our hierarchical
-    model -- that otter tracks laid on each day after a snow event are visible
-    throughout the entire observation period and therefore, that it is more
-    likely for an observer to detect a track on later days -- to be thoroughly
-    tested.  Most of the \(\psi\) and \(\theta\) estimates of 2003's data fall
-    into ranges where our model is at its best, both with respect to convergence
-    and to accurately estimating \(\psi\).
-
-    2004's data, which suffered from the lack of multiple observers and
-    observations later than one day after each snow event, does not provide
-    enough data with respect to time (measured by the number of days since the
-    last snow event) for the parameters \(p\) and \(E\) to converge, regardless
-    of the number of Markov Chain Monte Carlo (MCMC) iterations performed.
-
-    % Wow, this stuff SUCKS.
-    As plot size increases, every estimated parameter tends to increase as well.
-    Although the vast majority of these increases, even at the 1608m plot size,
-    lie well within the 95\% confidence interval of the accompanying estimate at
-    the 402m plot size, we believe that this trend makes sense.  Only one track
-    detection within a plot is necessary to assign it a 1.  Therefore, once 402m
-    plots are aggregated into 1608m plots, only one of the four 402m plots needs
-    to be assigned a 1 in order for the entire 1608m plot to receive a 1.  The
-    amount of correlation between 1's determines the increase in the proportion
-    of 1's to the number of plots -- as the amount of correlation increases, the
-    increase in the proportion decreases.
-
-    We believe that this increase leads to higher \(\psi\) and \(\theta\)
-    estimates because the potential for multiple otters to lay tracks within the
-    same plot increases.  In southeastern Minnesota, most otters tolerate the
-    presence of other otters in the areas they use most frequently, suggesting
-    that they are neither solitary nor territorial \cite{Gorman2006}.  This is
-    consistent with the noticeable but modest increase in \(\theta\) across plot
-    % I have no idea how to explain this... :(.
-    sizes for all three rivers.  However, estimates of \(\psi\) along the
-    Whitewater and Zumbro Rivers remain constant, despite large increases in
-    \(\theta\).
-
-    % More detail, maybe?
-    Similarly, the increases in estimates of \(p\) and \(E\) seem to result from
-    the general increase of tracks laid per plot as the plot size grows.  As
-    plot size grows, the influence of a 1 increases, regardless of whether an
-    otter track or a track from a different animal was observed.
-
-    Lastly, we decided to use the data from every plot even though we believe
-    that, as a result, our data violates the assumption of independence between
-    sites.  However, the model remains effective with spatially correlated data.
-    Since the actual data is relatively sparse as it is, we consider the removal
-    of half of it in order to preserve independence to be more harmful to the
-    model than supplying it with correlated data.  Therefore, all estimates come
-    from all of the data that we were given.
-
-    % Mississippi River!
-    \begin{center}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
-        \hline
-            & 402m & 804m & 1608m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.01954 & 0.04808 & 0.06885 \\
-            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.05216 & 0.07771 & 0.13650 \\
-            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.02668 & 0.06140 & 0.10230 \\
-            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.65240 & 0.67340 & 0.74960 \\
-            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.67770 & 0.77580 & 0.82540 \\
-            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.53670 & 0.54220 & 0.65330 \\
-            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.59380 & 0.69370 & 0.75820 \\
-            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.15280 & 0.20730 & 0.27690 \\
-            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    This data converged well at the 402m and 804m plot sizes.  At the 1608m plot
-    size, \(E\), \(\psi\), and \(\theta\) did not converge well.
-
-    \begin{center}
-    \begin{tabular}{|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Mississippi River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.3129 \\
-            & (0.2174, 0.8107) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.7613 \\
-            & (0.6609, 0.8833) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.8511 \\
-            & (0.2426, 0.9973) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    This data did not converge.  We do not think that these estimates are
-    accurate, but they are included for the sake of completeness.
-
-    % Whitewater River!
-    \begin{center}
-    \begin{tabular}{|l|l|l|}
-        \hline
-        \multicolumn{3}{|l|}{\textbf{Whitewater River, 2003}} \\
-        \hline
-            & 402m & 804m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.05608 & 0.0555 \\
-            & (0.02474, 0.09554) & (0.002859, 0.1390) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.16570 & 0.1931 \\
-            & (0.10130, 0.23510) & (0.065910, 0.3384) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.15350 & 0.2241 \\
-            & (0.09758, 0.21660) & (0.110000, 0.3623) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.66990 & 0.6985 \\
-            & (0.56080, 0.79040) & (0.573100, 0.8366) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.92040 & 0.9877 \\
-            & (0.84270, 0.97590) & (0.945100, 1.0000) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.75910 & 0.8575 \\
-            & (0.65060, 0.85800) & (0.745400, 0.9600) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.87590 & 0.9126 \\
-            & (0.72910, 0.98880) & (0.746900, 0.9955) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.24100 & 0.3779 \\
-            & (0.17890, 0.30900) & (0.286500, 0.4721) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    While the 402m data converged well, \(E\), \(\psi\), and \(\theta\) did not
-    converge well for the 804m data.
-
-    \begin{center}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Whitewater River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.8237 \\
-            & (0.5323, 0.9804) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.6066 \\
-            & (0.4246, 0.8382) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.2068 \\
-            & (0.1211, 0.3550) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    This data converged well.  However, we feel that the absence of \(E\) from
-    the model makes the estimates much less dependable.  The estimates are also
-    much less precise than those from 2003 -- the confidence intervals of
-    \(\psi\) and of \(p[Danny]\) are twice as large as those from 2003.
-
-    % Zumbro River!
-    \begin{center}
-    \begin{tabular}{|l|l|l|}
-        \hline
-        \multicolumn{3}{|l|}{\textbf{Zumbro River, 2003}} \\
-        \hline
-            & 402m & 804m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.02922 & 0.007066 \\
-            & (0.006786, 0.05994) & (0.0000, 0.03403) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.03388 & 0.008312 \\
-            & (0.008584, 0.06880) & (0.0000, 0.04188) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.02228 & 0.012960 \\
-            & (0.002430, 0.05385) & (0.0000, 0.05642) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.81890 & 0.837100 \\
-            & (0.749500, 0.88540) & (0.7615, 0.90700) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.81770 & 0.852500 \\
-            & (0.740100, 0.88670) & (0.7597, 0.92230) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.73120 & 0.805400 \\
-            & (0.625400, 0.82440) & (0.6885, 0.89780) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.66550 & 0.651600 \\
-            & (0.549100, 0.78150) & (0.5098, 0.78210) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.29950 & 0.395800 \\
-            & (0.236400, 0.36550) & (0.3100, 0.48520) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    Again, the 402m data converged well.  With respect to the 804m data, while
-    \(\psi\) and \(\theta\) converged well, none of the other parameters did.
-
-    \begin{center}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Zumbro River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.9166 \\
-            & (0.7739, 0.9918) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.4773 \\
-            & (0.3586, 0.6114) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.4232 \\
-            & (0.3073, 0.5520) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    This data converged well.  However, as is the case with all of the 2004
-    data, we feel that the absence of \(E\) from the model makes the estimates
-    much less dependable.
-
-    %% COVARIATES:
-    \subsection{Covariates}
-    To determine the relative importance of our covariates (the observer, the
-    date, and the number of days since the most recent snow event), we ran the
-    standard hierarchical model with \(E\) on simulated data with values of
-    \(\psi\) and \(\theta\) similar to our estimates of \(\psi\) and \(\theta\)
-    on 2003's data with varying numbers of observers, snow events, and
-    observations after each snow event.
-
-    \subsection{Best Practices for Measuring Otter Occupancy}
-
-\bibliographystyle{alpha}
-\bibliography{Bibliography}
 \end{document}

### 2010-02-24 Paper.tex
Added rest of my stuff.
Added Kiva's citations.

@@
     \subsection{WinBUGS/R}
-    To be filled in...
+    All of our statistical analysis was performed in R.
+
+    All of our modeling was performed using BRugs, which allowed us to use
+    WinBUGS commands and other stuff in R.
 
 \section{Standard Hierarchical Model}
-Also to be filled in... :(.
+
+    \subsection{Model Parameters}
+
+    \subsection{Simulated Data}
+    \noindent{\begin{tabular}{|l|l|l|l|}
+        \hline
+        \multicolumn{4}{|l|}{\textbf{The Model vs. Simulated Data}} \\
+        \hline
+            \(\psi\) & \(\theta\) & Convergence & 95\% CI (\(\psi\)) \\
+        \hline
+        \multirow{5}{*}{0.20}
+            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.112, 0.305) \\
+            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.142, 0.341) \\
+            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.098, 0.284) \\
+            & \cellcolor{Green}0.80 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.097, 0.281) \\
+            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.117, 0.302) \\
+        \hline
+        \multirow{5}{*}{0.40}
+            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.388, 0.721) \\
+            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.270, 0.494) \\
+            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.287, 0.512) \\
+            & \cellcolor{Red}0.80 & \cellcolor{Red}Bad &
+              \cellcolor{Red}(0.277, 0.507) \\
+            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.279, 0.503) \\
+        \hline
+        \multirow{5}{*}{0.60}
+            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
+              \cellcolor{Yellow}(0.312, 0.581) \\
+            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.400, 0.630) \\
+            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.462, 0.687) \\
+            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.502, 0.727) \\
+            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
+              \cellcolor{Red}(0.318, 0.549) \\
+        \hline
+        \multirow{5}{*}{0.80}
+            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.578, 0.838) \\
+            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.736, 0.915) \\
+            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.587, 0.829) \\
+            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.599, 0.811) \\
+            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
+              \cellcolor{Red}(0.612, 0.830) \\
+        \hline
+        \multirow{5}{*}{1.00}
+            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
+              \cellcolor{Yellow}(0.895, 0.999) \\
+            & \cellcolor{Yellow}0.40 & \cellcolor{Yellow}Okay &
+              \cellcolor{Yellow}(0.927, 0.999) \\
+            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
+              \cellcolor{Green}(0.949, 1.000) \\
+            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Good &
+              \cellcolor{Yellow}(0.802, 0.954) \\
+            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
+              \cellcolor{Red}(0.800, 0.955) \\
+        \hline
+    \end{tabular}}
 
@@
     variance, $c_{si}$ is the weight defining how strong the correlation is
-    between sites $s$ and $i$, and $c_{ii}=0$ for all sites $i$ (Arab et al.,
-    year?). Essentially, this model means that if site $s$'s neighbor has a
+    between sites $s$ and $i$, and $c_{ii}=0$ for all sites $i$ \cite{Arab2008}.
+    Essentially, this model means that if site $s$'s neighbor has a
     particularly high or low track-laying probability, that will increase or
@@
     $\gamma(0.5,0.0005)$ prior distribution for $(\tau^2)^{-1}$, a common choice
-    (GeoBUGS manual). In our model we define $\text{logit}(\theta_s)=\alpha_0+
+    \cite{Thomas2004}. In our model we define $\text{logit}(\theta_s)=\alpha_0+
     \alpha_{1_s}$ where $\alpha_0$ has an improper uniform distribution and

### 2010-02-24 Paper.tex
Added WinBUGS/R section.

@@
 \usepackage[all]{xy}
+\usepackage{hyperref}
 
@@
     \subsection{WinBUGS/R}
-    All of our statistical analysis was performed in R.
 
-    All of our modeling was performed using BRugs, which allowed us to use
-    WinBUGS commands and other stuff in R.
+    To implement a Bayesian hierarchical model, we make use of the software
+    packages WinBUGS \cite{Lunn2000}, R \cite{R2009}, BRugs \cite{Thomas2008},
+    and, by extension, OpenBUGS \cite{Thomas2006}, as one can describe the model
+    using an ``intuitive `pseudo-code' representation'' in WinBUGS
+    \cite{MacKenzie2006} and then run the model within the familiar environment
+    of R using the combination of BRugs and OpenBUGS. OpenBUGS, the open source,
+    cross-platform successor to WinBUGS, uses the same syntax. All of the
+    aforementioned software packages are open source and available
+    freely\footnote{WinBUGS: \url{http://www.mrc-bsu.cam.ac.uk/bugs/} and R:
+    \url{http://www.r-project.org/} are available as standalone programs. The
+    BRugs/OpenBUGS package can be installed by typing
+    \texttt{install.packages("BRugs")} in R's interactive shell. Note that while
+    R and OpenBUGS include official support for Mac OS X and Unix/Unix-like
+    operating systems, WinBUGS and the BRugs package are only available for
+    Windows.}.
+
+    WinBUGS requires a representation of a model, data, and initial values on
+    which a predetermined number of iterations of the MCMC algorithm are
+    executed \cite{MacKenzie2006}. Because WinBUGS explicitly estimates missing
+    data that is `padded out' by inserting NA values \cite{Spiegelhalter2003},
+    we implement a nested indexing strategy to prevent this extraneous
+    estimation on the actual data. This is a particularly important
+    implementation detail in this case because the data we received from the
+    Minnesota Department of Natural Resources was not consistent with respect to
+    our covariates. Neither the number of observers nor number of observations
+    after each snow event exhibited consistency. Our simulations, on the other
+    hand, involve balanced data sets that are much less difficult to implement
+    within WinBUGS and R.
+
+    % Should we say anything about making maps/other CAR-related stuff?
 

### 2010-02-25 Paper.tex
Changed bibliography style to 'amsplain'.
Prettified simple hierarchical model figure.
Added about a third of hierarchical model stuff.

@@
     Here is a visualization of how the Hierarchical model works:
-    % Either the margins need to be increased or we need a prettier diagram :(.
-    \begin{displaymath}
-        \xymatrix{
-        & & & \text{Site}\ar[dl]\ar[dr] & \\
-        & & \text{Occupied}\ar[dl]\ar[d] & & \text{Unoccupied}\ar[d] \\
-        & \text{Presence Indicated}\ar[dl]\ar[d] & 
-          \text{Presence Not Indicated}\ar[d] & & \text{Not Detected} \\
-        \text{Detected} & \text{Not Detected} & \text{Not Detected} & &}
-    \end{displaymath}
+
+    \begin{center}
+        \includegraphics[scale=0.75]{SimpleHierarchicalModel}
+    \end{center}
 
@@
 
-    \subsection{Model Parameters}
+    \subsection{Our Hierarchical Model}
+    For our model, we specified uniform \((0, 1)\) prior distributions for
+    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
+    formal information about the prior or expert consensus \cite{MacKenzie2006}.
+    We implemented models with normal and beta prior distributions on \(p\) and
+    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
+    range of \(E\) to the interval \((0, 0.3)\) in both models. We set the
+    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
+    parameters are tracked and therefore are estimated.
 
     \subsection{Simulated Data}
+    We simulated both balanced and unbalanced datasets meant for the standard
+    hierarchical model with given values of \(\psi\), \(\theta\), \(p\) for each
+    observer, and \(E\) for each observer.  Balanced data sets, in this context,
+    contain the same number of flights by the same observers after each snow
+    event.  Naturally, unbalanced data sets do not.
+
+    To simulate balanced data, we provided the following additional variables:
+    the number of sites, the number of snow events, the number of observers, and
+    the number of observations following each snow event. These variables are
+    all that are required to allow us to generate data in which each snow event
+    receives the same number of flights by the same number of observers, which
+    greatly simplifies the accompanying hierarchical model. First, the
+    simulation assigns a site a 0 (not occupied) or 1 (occupied) based upon the
+    given \(psi\) value.
+
     \noindent{\begin{tabular}{|l|l|l|l|}
@@
 
-\bibliographystyle{alpha}
+\bibliographystyle{amsplain}
 \bibliography{Bibliography}

### 2010-02-25 Paper.tex
Finished standard model stuff, added box plot figure.

@@
 
-    % INSERT BOXPLOT FROM STATUS REPORT HERE OF p-(1-q) Mississippi
+    \begin{center}
+        \includegraphics[scale=0.55]{PQBoxPlots}
+    \end{center}
+
     We found no obvious trend indicating one site size to be more independent
@@
     \subsection{WinBUGS/R}
-
     To implement a Bayesian hierarchical model, we make use of the software
     packages WinBUGS \cite{Lunn2000}, R \cite{R2009}, BRugs \cite{Thomas2008},
-    and, by extension, OpenBUGS \cite{Thomas2006}, as one can describe the model
+    and, by extension, OpenBUGS \cite{Thomas2006}. One can describe the model
     using an ``intuitive `pseudo-code' representation'' in WinBUGS
@@
 
-    % Should we say anything about making maps/other CAR-related stuff?
-
 \section{Standard Hierarchical Model}
@@
     greatly simplifies the accompanying hierarchical model. First, the
-    simulation assigns a site a 0 (not occupied) or 1 (occupied) based upon the
-    given \(psi\) value.
-
-    \noindent{\begin{tabular}{|l|l|l|l|}
+    simulation determines a site's occupancy based upon the given \(psi\) value.
+    Then, for each day after each snow event, the simulation decides whether or
+    not an otter laid down a track. For a given day after a snow event, a track
+    is considered to be present if an otter lays a track down on any of the days
+    after the snow event up to and including the given day.  Once that has been
+    decided, the given \(p\) and \(E\) values are used to assign a 0 (not
+    detected) or 1 (detected).
+
+    Because of limitations with respect to WinBUGS' syntax, simulating
+    unbalanced data requires much more additional information and a reformatting
+    of the data as well as the corresponding hierarchical model. While the
+    resulting code's structure differs significantly from that used for balanced
+    data, the underlying model remains the same.
+
+    To determine the efficacy of the standard model over the range of possible
+    \(\psi\) and \(\theta\) values, we ran simulations with several combinations
+    of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
+    the model's estimates to the simulation's parameters. A table detailing the
+    model's performance lies below:
+
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|}
         \hline
@@
         \hline
-    \end{tabular}}
+    \end{tabular}
+    \end{center}
+
+    Each of the simulations represented in the table above ran with the
+    following parameters: seventy sites, five snow events, three observers, and
+    three observations after each snow event. The three observers had \(p\)
+    values of 0.70, 0.80, 0.90 and \(E\) values of 0.05, 0.20, and 0.30,
+    respectively.
+
+    Green rows represent values of \(\psi\) and \(\theta\) for which our
+    standard hierarchical model exhibits good convergence and provides a 95\%
+    confidence interval for \(\psi\) that includes the simulated data's actual
+    \(\psi\) value.  Yellow rows represent values of \(\psi\) and \(\theta\) for
+    which our model exhibits either mediocre convergence with a good confidence
+    interval or good convergence with a bad confidence interval.  Red rows
+    represent values of \(\psi\) and \(\theta\) for which our model does not
+    converge.  The 95\% confidence interval is ignored in these cases.
+
+    Generally, the model is most effective for lower values of \(\psi\) and
+    \(\theta\) but will either converge well or provide useful estimates for the
+    vast majority of combinations of \(\psi\) and \(\theta\) values.
 

### 2010-02-25 Paper.tex
added figure, edited false positive section

@@
     among the proportion of sites in which the scientists detected a track
-    (Figure x). The difference in estimated na\"ive occupancy among the
+    (Figure \ref{obsPlots}). The difference in estimated na\"ive occupancy among 
+    the
     observers indicates that the scientists did not always correctly identify
     the tracks.
+
+    \begin{figure}
+        	\centering
+	    \includegraphics[width=5in]{observerPlots.pdf}
+	    \caption{The box plots show how the proportion of sites in which an 
+	    observer saw tracks (``occupancy'') differs among the three observers in 
+	    2003. Only those dates on which all three observers flew are included.}
+	    \label{obsPlots}
+    \end{figure}
 
@@
     the number of flights increases, the occupancy rate estimate will approach
-    one (Royle and Link 2006). We introduced a false detection process and used
-    the observer as a covariate (by means of indicator variables) for both the
-    probability of correctly identifying a track that exists ($p$) and the
-    probability of identifying a track that does not exist ($E$). Including
-    false positives introduces a duality whereby $(p=x,E=y,\psi=z)$ is as good a
-    solution as $(p=y,E=x,\psi=1-z)$. To address this problem we assumed
-    $p\gg0.5$ and $E\ll0.5$, an approach recommended by Royle and Link (2006).
+    1 \cite{Royle2006}. Unlike most occupancy models where scientists assume 
+    that detection of the species in question guarantees its presence \cite
+    {MacKenzie2006}, we introduced a false detection process whereby an observer 
+    can mark down the existence of a track when in reality there is none. This 
+    false detection process can occur as long as there is no track, whether or 
+    not the site is occupied.
 
@@
     greater than \(E\), we solved this problem by restricting the distribution
-    of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to \(0\text{ to }0.4\).
+    of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to \(0\text{ to }0.4\),  
+    an approach recommended by Royle and Link \cite{Royle2006}.
 
@@
 
-    Figure x caption:
-    The box plots show how the proportion of sites in which an observer saw
-    tracks (``occupancy'') differs among the three observers. Only those dates
-    in which all three observers flew are included.
-
 \section{Discussion}

### 2010-02-25 Paper.tex
Added tables and figure.

--- 2010-02-25 Paper.tex
@@
     trends for this statistic with respect to observer, days since snow, snow
-    event, or plot size.
-
-    % FIGURES OR TABLES??? WHICH ONES? from status report compilation
+    event, or site size.
+
+    % Mississippi River!
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|l|l|}
+        \hline
+        \multicolumn{6}{|l|}{\textbf{Mississippi River}} \\
+        \hline
+        \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
+        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} &
+        \multirow{2}{*}{\(E[|z_{all}|-|z_{alt}|]\)} \\
+        \cline{4-5}
+        & & & All Plots & Alt. Plots & \\
+        \hline
+        \multirow{3}{*}{2003} & 402m & \multirow{3}{*}{20} & 7 & 0 & 0.915 \\
+        \cline{2-2} \cline{4-6}
+        & 804m & & 5 & 0 & 0.651 \\
+        \cline{2-2} \cline{4-6}
+        & 1608m & & 1 & 0 & 0.383 \\
+        \hline
+        \multirow{3}{*}{2004} & 402m & \multirow{3}{*}{6} & 5 & 3 & 0.625 \\
+        \cline{2-2} \cline{4-6}
+        & 804m & & 3 & 0 & 1.428 \\
+        \cline{2-2} \cline{4-6}
+        & 1608m & & 2 & 0 & 0.053 \\
+        \hline
+    \end{tabular}
+    \end{center}
+
+    % Whitewater River!
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|l|l|}
+        \hline
+        \multicolumn{6}{|l|}{\textbf{Whitewater River}} \\
+        \hline
+        \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
+        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} &
+        \multirow{2}{*}{\(E[|z_{all}|-|z_{alt}|]\)} \\
+        \cline{4-5}
+        & & & All Plots & Alt. Plots & \\
+        \hline
+        \multirow{2}{*}{2003} & 402m & \multirow{2}{*}{15} & 6 & 2 & 0.722 \\
+        \cline{2-2} \cline{4-6}
+        & 804m & & 4 & 1 & 0.634 \\
+        \hline
+        \multirow{2}{*}{2004} & 402m & \multirow{2}{*}{6} & 3 & 1 & 1.189 \\
+        \cline{2-2} \cline{4-6}
+        & 804m & & 2 & 2 & 0.612 \\
+        \hline
+    \end{tabular}
+    \end{center}
+
+    % Zumbro River!
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|l|l|}
+        \hline
+        \multicolumn{6}{|l|}{\textbf{Zumbro River}} \\
+        \hline
+        \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
+        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} &
+        \multirow{2}{*}{\(E[|z_{all}|-|z_{alt}|]\)} \\
+        \cline{4-5}
+        & & & All Plots & Alt. Plots & \\
+        \hline
+        \multirow{2}{*}{2003} & 402m & \multirow{2}{*}{11} & 10 & 6 & 1.883 \\
+        \cline{2-2} \cline{4-6}
+        & 804m & & 8 & 5 & 1.287 \\
+        \hline
+        \multirow{2}{*}{2004} & 402m & \multirow{2}{*}{4} & 3 & 3 & 2.425 \\
+        \cline{2-2} \cline{4-6}
+        & 804m & & 3 & 1 & 2.343 \\
+        \hline
+    \end{tabular}
+    \end{center}
 
@@
     \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
-    range of \(E\) to the interval \((0, 0.3)\) in both models. We set the
+    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
     burn-in time for the MCMC simulation to 10,000 iterations. Each of these
-    parameters are tracked and therefore are estimated.
+    parameters are tracked and therefore are estimated. The covariates of this
+    model are: observer, date, and the number of days since the last snow event.
 

### 2010-02-25 Paper.tex
Messed with titles.

@@
 
-\section{Overview of Occupancy Modeling}
-
-    \subsection{What is Occupancy Rate?}
+\section{Overview of occupancy modeling}
+
+    \subsection{What is occupancy rate?}
     Occupancy has two different definitions. It can mean ``the probability that
@@
 
-    \subsection{Different Ways of Sampling Data}
+    \subsection{Different ways of sampling data}
     In order to determine occupancy rate, the data is often collected in the
@@
 
-    \subsection{Types of Occupancy Models}
+    \subsection{Types of occupancy models}
     All of the models discussed in this paper are single-species, single-season
@@
 
-    \subsection{Markov Chains and Testing the Independence Assumption}
+    \subsection{Testing the independence assumption}
     \label{shat}
@@
 
-\section{Theoretical Stuff}
-
-    \subsection{Bayesian Statistics}
+\section{Bayesian modeling theory and practice}
+
+    \subsection{Bayesian statistics}
     Bayes' Theorem: \(P(\theta|X) = P(X|\theta)P(\theta) / P(X)\) \\
@@
 
-\section{Standard Hierarchical Model}
-
-    \subsection{Our Hierarchical Model}
+\section{Standard hierarchical model}
+
+    \subsection{Our hierarchical model}
     For our model, we specified uniform \((0, 1)\) prior distributions for
@@
 
-    \subsection{Simulated Data}
+    \subsection{Simulated data}
     We simulated both balanced and unbalanced datasets meant for the standard
@@
 
-    \subsection{Obtaining Convergence on the Actual Data}
+    \subsection{Obtaining convergence on the data}
     Recall that our standard hierarchical model contains the following
@@
 
-    \subsection{Future Work}
+    \subsection{Future work}
     Additional observations are more helpful than additional snow events at

### 2010-02-25 Paper.tex
Some edits to my sections

@@
 time. We used the concepts from occupancy theory to develop models to estimate
-abundance.
+the occupancy rate in order to get a sense of otter abundance.
 
@@
     people should have seen the same tracks. While there is some error in
-    classifying which site the observer saw the actual track in, each observer
+    classifying in which plot the observer saw the actual track, each observer
     should at least be seeing the same, or nearly the same, number of sites with
@@
     variation in the data not necessarily due to these classification issues.
-    This is difficult to quantify, as an algorithm for finding classification
-    errors would be complicated. Correlation between flights is a bit higher for
+    This is difficult to quantify. Correlation between flights is a bit higher 
+    for
     flights on the same day or flights during the same snow event, but still not
-    as high as one might hope. The proportion of sites with tracks detected
+    as high as one might hope. Correlation rarely exceeds 0.5. The proportion of 
+    sites with tracks detected
     generally increases with days since snow, likely because the otters have a
@@
     not the site is occupied.
-
     \subsection{Testing the independence assumption}
@@
     model}, the model almost never converged. We tested independence of the data
-    using the $\hat{s}$ statistic we developed in section \ref{shat} and found
+    using the $\hat{s}$ statistic we developed (section \ref{shat}) and found
     that the simulated data displayed much more independence than the actual
@@
 
-    To increase the correlation of the tracks we simulated data with spatial
+    To increase the correlation of the observed tracks we simulated data with 
+    spatial
     correlation in both $\theta$ and $\psi$. Under this simulation method we

### 2010-02-27 Presentation/Presentation.tex
Added presentation.

\documentclass{beamer}
\usetheme{}
\title{Using occupancy theory and Bayesian models to estimate otter populations in southeastern Minnesota}
\author[Aing, Halls, Oken]{Chrisna Aing, Sarah Halls, Kiva Oken}
\institute{Carleton College}
\usepackage{graphpap,graphics}
%\includeonlyframes{}

\begin{document}
	{
	\usebackgroundtemplate{\includegraphics[width=\paperwidth]{ottertracks1.jpg}}

\begin{frame}
	\titlepage
\end{frame}
}
\begin{frame}
	\tableofcontents[pausesections]
\end{frame}

\section{The data and questions}
\begin{frame}{Otters}
	\begin{columns}
		\column{5cm}
		\begin{itemize}[<+->]
			\item Otters endemic to all of Minnesota
			\item Indicator species for quality of aquatic habitat
			\item Populations crashed, now rebounding
			\item In SE Minnesota interest in resuming trapping
			\item Government needs tools to monitor changes in populations
		\end{itemize}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{cuteOtter1.pdf}}
	\end{columns}
\end{frame}

\begin{frame}{Collecting data}
	\begin{columns}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{cuteOtter1.pdf}}
		\column{5cm}
		\begin{itemize}[<+->]
			\item Researchers flew over rivers during winters of 2003, 2004
			\item Flew one, two, or three days after snow events
			\item Marked GPS coordinates of tracks
			\item Rivers divided into 400, 800, or 1600 m plots
			\item Presence of track in each plot determined
		\end{itemize}
	\end{columns}
\end{frame}

{
	\usebackgroundtemplate{\includegraphics[width=\paperwidth]{SnowTracks12.jpg}}
	\begin{frame}{}
	\end{frame}
}

{
	\usebackgroundtemplate{\includegraphics[width=\paperwidth]{SNOWTRACKS9.jpg}}
	\begin{frame}
	\end{frame}
}

\begin{frame}{Issues related to data}
	\begin{itemize}
		\item Tracks near plot boundary can be assigned to different plots
		\item Track in one site means more likely to be track in neighboring site- not independent
	\end{itemize}
\end{frame}

\section{Occupancy Theory}
\begin{frame}{Definitions of occupancy}
	\begin{itemize}
		\item The probability that a randomly selected site or sampling unit in an area of interest is occupied by a species
		\item The proportion of area, patches, or sampling units that is occupied 	\end{itemize}
	(MacKenzie, et al. 2006) 
\end{frame}

\begin{frame}{Estimating occupancy}
	\begin{itemize}
		\item Ecologists often conduct surveys to find the occupancy rate
		\item Not observing the species can mean two things:
		\begin{itemize}
			\item The species does not occupy the area
			\item The species went undetected during surveying
		\end{itemize}
		\item Need to be able to estimate probability of observing the species given its presence
	\end{itemize}
\end{frame}

\begin{frame}{Assumptions of occupancy models}
	\begin{itemize}
		\item No false positives
		\item Presence and detection probabilities are constant across sites and surveys (or can be modeled by covariates)
		\item Independence of detection between sites
		\item Survey is conducted on a closed population	\end{itemize}
\end{frame}

\begin{frame}{Types of occupancy models}
	\begin{itemize}
		\item Single species, single season
		\item Na\"ive occupancy: proportion of sites found to be occupied
		\item Hierarchical models
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{SimpleHierarchicalModel.pdf}}
	\end{center}
\end{frame}

\begin{frame}{Parameters}
	\begin{itemize}
		\item $\psi=$ The probability a track is occupied
		\item $\theta=$ The probability a track is laid down in a site on a given day, given that the site is occupied
		\item $p= $ The probability an observer sees a track given that a track has been made
	\end{itemize}
\end{frame}

\section{Exploring the data}
\begin{frame}{Effects of multiple observers}
	\begin{itemize}
		\item In 2003, three different observers
		\item Detection rates varied among the observers
		%include graph here
		\item Solution: use observer indicators as covariates for detection probability
	\end{itemize}
\end{frame}

\begin{frame}{Correlation between flights}
	\begin{itemize}
		\item Two observers on same day should have the same na\"ive occupancy rates; not the case
		\item Also looked at correlation between flights
		\begin{itemize}
			\item Very low, overall
			\item Slightly higher during the same snow event or day
		\end{itemize}
		\item Solution: include false detection in model
	\end{itemize}
\end{frame}

%Include frame with new hierarchical diagram

\begin{frame}{Independence between plots}
	\begin{itemize}
		\item Assumption: independence between plots
		\item %Sarah going to review
		\item Solution: kept issue in mind, explored spatially explicit models
	\end{itemize}
\end{frame}

\section{Modeling the data}
\begin{frame}{Bayesian statistics}
\end{frame}

\begin{frame}{Our model}
\end{frame}

\begin{frame}{Testing the model- data simulations}
\end{frame}

\begin{frame}{Modeling spatial correlation}
\end{frame}

\begin{frame}{Testing the CAR model}
\end{frame}

\section{Discussion}
\begin{

\end{document}

### 2010-02-27 Paper.tex
Edited Occupancy and Independence sections

@@
     really is no animal is also a possible problem, but is thought to be much
-    less likely and is often ignored. We will discuss this issue more below.
+    less likely and is often ignored. We will discuss this issue more in later sections.
 
     \subsection{Assumptions}
-    There are some assumptions that are made in most models of occupancy.
-    \cite{MacKenzie2006}. Without these assumptions, the models become much more
-    complicated. One common assumption is no false positives, which means if a
+    As is true for all modeling, there are some assumptions made in most models of occupancy 
+in order to create simple enough models \cite{MacKenzie2006}. One common assumption is no false positives, which means if a
     site is recorded as occupied we say it must actually be occupied. A second
@@
     sites. A fourth common assumption is that the surveys are being conducted on
-    a closed population. This means that no animals leave the area or join the
+    a closed population. This means that no animals leave the area or joins the
     area during the time of the surveys.
 
-    Obviously, not all of these assumptions are always true, but they are often
-    necessary in order to get a model that is simple enough to use. There often
-    are possible ways to get rid of these assumptions, but it complicates the
-    model. In this paper, we will only talk about attempting to get rid of the
-    assumption of no false positives, and the assumption of independence of
+    There are different methods for dealing with violations of these assumptions, some 
+more developed than others.  In this paper, we will only talk about attempting to get rid of the
+    assumption of no false positives and the assumption of independence of
     parameters across sites.
@@
     models. It is possible to modify all of these models to be used for multiple
-    seasons or multiple species, but it would make them more complicated.
+    seasons or multiple species, at the cost of added complexity.
 
@@
     accepted that this estimate is too low because animals are not detected at
-    all sites where they are present, which is the issue of false negatives
+    all sites where they are present, due to the issue of false negatives
     mentioned above. Other occupancy models attempt to do better than this
@@
 
-    One way to create more accurate models is to use a Hierarchical model as a
+    One way to create a more accurate model is to attempt to model the biological 
+proccess that is occuring.  In order to do this, we can use a Hierarchical model as a
     way to think about the parameters that affect occupancy. This allows us to
@@
     not the site is occupied.
+
     \subsection{Testing the independence assumption}
     \label{shat}
-    First we considered using Markov Chains to come up with a long term
-    probability of detection, which we called \(r\). We let \(p\) be the number
-    of \(0\)'s followed by a \(0\) as a fraction of all successive pairs that
-    start with a \(0\). We let \(q\) be the number of \(1\)'s followed by a
-    \(1\) as a fraction of all successive pairs that start with a \(1\). These
-    probabilities can be represented by the matrix:
-    \[M=\begin{bmatrix}
-        p & 1-p \\
-        1-q & q
-    \end{bmatrix}\text{.}\]
-
-    The long term probability of detection can be found from this by finding
-    \(\lim_{n \to \inf}M^n\), which returns
-    \[\begin{bmatrix}
-        1-r & r \\
-        1-r & r
-    \end{bmatrix}\text{,}\]
-    where \(r\) is the probability of a \(1\), which is the long term
-    probability of detection.
-
-    We calculated \(p\), \(q\), and \(r\) for every flight in our data, using
-    \(n = 500\). We then realized that \(r\) is actually not so useful in
-    analyzing our data. It does not take into account whether or not a site is
-    occupied, or whether or not a track was laid down on that site, or any of
-    the ideas behind the hierarchical model. These are the parameters that we
-    eventually want to find out something about, and the value of \(r\) cannot
-    help us get there. The idea to use Markov Chains was still useful, however.
-
-    Although we rejected the idea of using \(r\) as a way to analyze the data,
-    we realized that \(p\) and \(q\) could be used to test the viability of the
-    independence assumption. If there is independence between sites, the
-    difference of the proportions \(p\) and \(1-q\) should be approximately
-    \(0\). This would mean that the probability that a \(0\) follows a \(0\) is
-    about the same as the probability that a \(1\) follows a \(0\), meaning that
-    it does not depend on what the observation is at the previous site. Looking
-    at boxplots of \(p-(1-q)\), we see that data from flights that record
-    alternating sites may be more independent than data from flights that record
-    all sites, which suggests a possible benefit to using the alternating
-    sampling method. The boxplots for the Whitewater and Zumbro Rivers looked
-    similar to the Mississippi:
-
-    \begin{center}
-        \includegraphics[scale=0.55]{PQBoxPlots}
-    \end{center}
-
-    We found no obvious trend indicating one site size to be more independent
-    than the others.
-
-    While this is an interesting observation, we would like to be able to
-    quantify it better by finding the distribution of our statistic. This would
-    allow us to make statements about how independent the data is, and if the
-    difference in independence between different types of data is significant.
-    This statistic seems hard to get, but we considered a modification. Let
-    \(S\) be the number of successive \(0\), \(1\)'s minus the number of
+    Since one of the assumptions of occupancy modeling is that there is 
+	independence between sites, we wanted to see if this assumption held true 
+	for our data.  In order to do this, we calculated the statistic \(S\), which
+	we define as the number of successive \(0\), \(1\)'s minus the number of
     successive \(1\), \(1\)'s per flight. Under the null hypothesis that each
-    site is an independent Bernoulli trial with probability \(p\) we were able
+    site is an independent Bernoulli trial with probability of a \(1 = p\) we were able
     to compute the mean and standard deviation of \(S\), which we used to
-    calculate a Z-score. In order to compute these values, we needed to find the
-    expected value and variance of each term in \(S\). We let \(A = \text{\# of
-    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).
-
-    In order to find \(E[\hat{s}]\), we first needed to find \(E[A]\) and
-    \(E[B]\).
-
-    % E[A]
-    \begin{equation*}
-        \begin{split}
-        E[A]& =\sum_{i=1}^{n-1}P(X_i=0,X_{i+1}=1) \\
-            & =\sum_{i=1}^{n-1}(1-p)p \\
-            & =(n-1)(1-p)p \\
-        \end{split}
-    \end{equation*}
-
-    % E[B]
-    \begin{equation*}
-        \begin{split}
-        E[B]& =\sum_{i=1}^{n-1}P(X_i=1,X_{i+1}=1) \\
-            & =\sum_{i=1}^{n-1}p*p \\
-            & =(n-1)p^2
-        \end{split}
-    \end{equation*}
-
-    % E[s-hat]
-    \begin{equation*}
-        \begin{split}
-        E[\hat{s}]& =E[A]-E[B] \\
-                  & =(n-1)\left((1-p)p-p^2\right)
-        \end{split}
-    \end{equation*}
-
-    The standard deviation was a little more difficult.
-
-    % Var[A]
-    \begin{equation*}
-        \begin{split}
-        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[I_i=(0,1),I_{i+1}=
-                 (1,1)] \\
-              & =(n-1)P((0,1))(1-P((0,1))) \\
-              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(0,1),I_{i+1}=(0,1)]-E[I_i=
-                (0,1)]E[I_{i+1}=(0,1)]\bigg) \\
-              & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
-                \bigg) \\
-              & =(1-p)p(-1+n+5p-3np-5p^2+3np^2)
-        \end{split}
-    \end{equation*}
-
-    % Var[B]
-    \begin{equation*}
-        \begin{split}
-        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[I_i=(1,1),I_{i+1}=
-                (1,1)] \\
-              & =(n-1)P((1, 1))(1-P((1, 1))) \\
-              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,1),I_{i+1}=(1,1)]-E[I_i=
-                (1,1)]E[I_{i+1}=(1,1)]\bigg) \\
-              & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
-              & =(1-p)p^2(-1+n-5p+3np)
-        \end{split}
-    \end{equation*}
-
-    % Cov[A, B]
-    \begin{equation*}
-        \begin{split}
-        Cov[A,B]& =\sum_{i=1}^{n-1}\sum_{j=1}^{n-1}Cov[A,B] \\
-                & =\sum_{i=1}^{n-2}Cov[A,B]+\sum_{i=1}^{n-2}Cov[B,A] \\
-                & \quad +\sum_{i=1}^{n-1}Cov[X_i=0\text{ or }1,X_{i+1}=1] \\
-                & =\sum_{i=1}^{n-2}E[AB]-E[A]E[B]+\sum_{i=1}^{n-2}E[BA]-
-                  E[B]E[A] \\
-                & \quad +\sum_{i=1}^{n-1}E[X_i=0\text{ or }1,X_{i+1}=1]-
-                  E[B]E[A] \\
-                & =(n-2)(p-1)^2p^2-(n-2)p^3(1-p)-(n-1)p^3(1-p)
-        \end{split}
-    \end{equation*}
-
-    % Var[s-hat]
-    \begin{equation*}
-        \begin{split}
-        Var[\hat{s}]& =Var[A-B] \\
-                    & =Var[A]+Var[B]-2Cov[A,B] \\
-                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
-        \end{split}
-    \end{equation*}
+    calculate a Z-score, where \(Z = (S – E[S])/SD[S]\). In order to compute these values, we needed to find the
+    expected value and variance of each term in \(S\). These caclulations can be
+	found in the appendix.
 
@@
     computed a Z-score for each flight, where we use a plug-in estimate for
-    \(p\) to be the proportion of \(1\)'s per flight. We said that Z-scores with
-    an absolute value greater than \(2\) suggest that the assumption of
+    \(p\) to be the proportion of \(1\)'s per flight. We said that \(|Z| > 2\) suggests that the assumption of
     independence between sites is violated for that flight. We found that, in
@@
     For instance, six of eleven flights along the Zumbro River during 2003 still
-    have significant Z-scores with alternating sites. There were no noticeable
-    trends for this statistic with respect to observer, days since snow, snow
-    event, or site size.
+    have significant Z-scores with alternating sites. It also appears that there 
+	may be slightly more independence when larger site sizes are used, although it is
+not clear if this trend is significant.  There 
+	were no noticable trends for this statistic with respect to observer, days 
+	since snow, or snow event.
 
@@
 
+%Needs to be fixed/formatted better... I didn't know how to make an appendix
+\appendix
+	We let \(A = \text{\# of
+    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).
+    In order to find \(E[s]\), we first needed to find \(E[A]\) and
+    \(E[B]\).
+
+    % E[A]
+    \begin{equation*}
+        \begin{split}
+        E[A]& =\sum_{i=1}^{n-1}P(X_i=0,X_{i+1}=1) \\
+            & =\sum_{i=1}^{n-1}(1-p)p \\
+            & =(n-1)(1-p)p \\
+        \end{split}
+    \end{equation*}
+
+    % E[B]
+    \begin{equation*}
+        \begin{split}
+        E[B]& =\sum_{i=1}^{n-1}P(X_i=1,X_{i+1}=1) \\
+            & =\sum_{i=1}^{n-1}p*p \\
+            & =(n-1)p^2
+        \end{split}
+    \end{equation*}
+
+    % E[s]
+    \begin{equation*}
+        \begin{split}
+        E[s]& =E[A]-E[B] \\
+                  & =(n-1)\left((1-p)p-p^2\right)
+        \end{split}
+    \end{equation*}
+
+    The standard deviation was a little more difficult.
+
+    % Var[A]
+    \begin{equation*}
+        \begin{split}
+        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[I_i=(0,1),I_{i+1}=
+                 (1,1)] \\
+              & =(n-1)P((0,1))(1-P((0,1))) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(0,1),I_{i+1}=(0,1)]-E[I_i=
+                (0,1)]E[I_{i+1}=(0,1)]\bigg) \\
+              & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
+                \bigg) \\
+              & =(1-p)p(-1+n+5p-3np-5p^2+3np^2)
+        \end{split}
+    \end{equation*}
+
+    % Var[B]
+    \begin{equation*}
+        \begin{split}
+        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[I_i=(1,1),I_{i+1}=
+                (1,1)] \\
+              & =(n-1)P((1, 1))(1-P((1, 1))) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,1),I_{i+1}=(1,1)]-E[I_i=
+                (1,1)]E[I_{i+1}=(1,1)]\bigg) \\
+              & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
+              & =(1-p)p^2(-1+n-5p+3np)
+        \end{split}
+    \end{equation*}
+
+    % Cov[A, B]
+    \begin{equation*}
+        \begin{split}
+        Cov[A,B]& =\sum_{i=1}^{n-1}\sum_{j=1}^{n-1}Cov[A,B] \\
+                & =\sum_{i=1}^{n-2}Cov[A,B]+\sum_{i=1}^{n-2}Cov[B,A] \\
+                & \quad +\sum_{i=1}^{n-1}Cov[X_i=0\text{ or }1,X_{i+1}=1] \\
+                & =\sum_{i=1}^{n-2}E[AB]-E[A]E[B]+\sum_{i=1}^{n-2}E[BA]-
+                  E[B]E[A] \\
+                & \quad +\sum_{i=1}^{n-1}E[X_i=0\text{ or }1,X_{i+1}=1]-
+                  E[B]E[A] \\
+                & =(n-2)(p-1)^2p^2-(n-2)p^3(1-p)-(n-1)p^3(1-p)
+        \end{split}
+    \end{equation*}
+
+    % Var[s]
+    \begin{equation*}
+        \begin{split}
+        Var[s]& =Var[A-B] \\
+                    & =Var[A]+Var[B]-2Cov[A,B] \\
+                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
+        \end{split}
+    \end{equation*}
+
 \end{document}

### 2010-02-27 Paper.tex
Edited my sections

--- 2010-02-25 Paper.tex
@@
     days that all three people observed, there appear to be major differences
-    among the proportion of sites in which the scientists detected a track
+    among the proportion of sites in which the observers detected a track
     (Figure \ref{obsPlots}). The difference in estimated na\"ive occupancy among 
     the
-    observers indicates that the scientists did not always correctly identify
+    observers indicates that the observers did not always correctly identify
     the tracks.
@@
     The lack of consistency among the flights is strong evidence that the data
-    contain false positives where scientists think they see a track when it does
+    contain false positives where observers think they see a track when it does
     not exist. Not accounting for such a process can lead to major biases; as
     the number of flights increases, the occupancy rate estimate will approach
-    1 \cite{Royle2006}. Unlike most occupancy models where scientists assume 
+    1 \cite{Royle2006}. Unlike most occupancy models where observers assume 
     that detection of the species in question guarantees its presence \cite
@@
 \section{Spatial correlation of the data}
-One approach to dealing with the significant correlation in the probability of
+One approach to dealing with the correlation in the probability of
 observing a track in neighboring sites was to use a conditional autoregressive
@@
     that the simulated data displayed much more independence than the actual
-    data. We tried simulating spatial correlation only in $\psi$ with similar
-    results.
-
-    To increase the correlation of the observed tracks we simulated data with 
-    spatial
-    correlation in both $\theta$ and $\psi$. Under this simulation method we
+    data. We tried modeling spatial correlation in the occupancy process, $\psi
+    $, instead of in the track-laying process, but the simulated data were again 
+    more independent than the real data.
+
+    To increase the correlation between sites we simulated data
+    correlated in both the track-laying and occupancy processes. Under this 
+    simulation method we
     again had the variables $\theta$ and $\alpha$, but we also had analogous
@@
 
+\section{Acknowledgments}
+We would like to thank our adviser, Bob Dobrow, for all of his help and advice and John Fieberg and John Erb from the Minnesota Department of Natural Resources for collecting and sharing the data as well as for offering their specific knowledge of the subject.
+
 \bibliographystyle{amsplain}

### 2010-02-28 Paper.tex
edited beginning and bayesian statistics

@@
 \maketitle
-
-\section{Overview of occupancy modeling}
-
-    \subsection{What is occupancy rate?}
-    Occupancy has two different definitions. It can mean ``the probability that
-    a randomly selected site or sampling unit in an area of interest is occupied
-    by a species'' \cite{MacKenzie2006}. It also can be used to describe ``the
-    proportion of area, patches, or sample units that is occupied''
-    \cite{MacKenzie2006}. Occupancy is not the same as the abundance, or
-    population size, of a species. The number of members of a species is not
-    important for either definition of occupancy.
-
-    While there is a difference between these two definitions of occupancy, in
-    general the probability is unknown and the observed proportion is used as an
-    estimate of the underlying probability. Because of this, these two
-    definitions of occupancy are oftentimes used interchangeably.
-
-    The main reasons that people are interested in measuring occupancy are
-    science or conservation and management \cite{MacKenzie2006}. This statistic
-    can give insight into the way a system works. It also can be used to monitor
-    animals habitats and discover trends over time in groups of animals.
-
-    \subsection{Different ways of sampling data}
-    In order to determine occupancy rate, the data is often collected in the
-    form of a presence-absence survey. This means that observers visit sites and
-    look for the animals themselves, tracks, vocalizations, or some other
-    indicator that the animal is present. In some cases, camera traps and sound
-    recording are used instead of human observers \cite{MacKenzie2006}.
-
-    The main problem with these methods of sampling data is that it is possible
-    for an animal to occupy a site, but not be detected by the observer. This
-    means that the site is recorded as unoccupied, but there really is an animal
-    there. This is called a false negative. Marking an animal present when there
-    really is no animal is also a possible problem, but is thought to be much
-    less likely and is often ignored. We will discuss this issue more in later sections.
-
-    \subsection{Assumptions}
-    As is true for all modeling, there are some assumptions made in most models of occupancy 
-in order to create simple enough models \cite{MacKenzie2006}. One common assumption is no false positives, which means if a
-    site is recorded as occupied we say it must actually be occupied. A second
-    common assumption is that presence and detection probabilities are constant
-    across sites and surveys. This simplifies the estimation of each parameter.
-    A third common assumption is that there is independence of detection between
-    sites. A fourth common assumption is that the surveys are being conducted on
-    a closed population. This means that no animals leave the area or joins the
-    area during the time of the surveys.
-
-    There are different methods for dealing with violations of these assumptions, some 
-more developed than others.  In this paper, we will only talk about attempting to get rid of the
-    assumption of no false positives and the assumption of independence of
-    parameters across sites.
-
-    \subsection{Types of occupancy models}
-    All of the models discussed in this paper are single-species, single-season
-    models. It is possible to modify all of these models to be used for multiple
-    seasons or multiple species, at the cost of added complexity.
-
-    The most simple way to measure occupancy is to simply divide the number of
-    sites where an animal was found to be present by the total number of sites.
-    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
-    accepted that this estimate is too low because animals are not detected at
-    all sites where they are present, due to the issue of false negatives
-    mentioned above. Other occupancy models attempt to do better than this
-    simple estimate.
-
-    One way to create a more accurate model is to attempt to model the biological 
-proccess that is occuring.  In order to do this, we can use a Hierarchical model as a
-    way to think about the parameters that affect occupancy. This allows us to
-    take into account that when an animal is not detected, it could either not
-    be there, or it could be there and just was not detected. This could either
-    be due to a false negative, where the observer did not see the indication --
-    for example they did not see the animal, track, etc. -- or it could be due
-    to the fact that the animal does occupy the site but there was no indication
-    of the animal at the site -- for example, no track, no vocalization, etc.
-
-    Here is a visualization of how the Hierarchical model works:
-
-    \begin{center}
-        \includegraphics[scale=0.75]{SimpleHierarchicalModel}
-    \end{center}
-
-    From this diagram, it is apparent that we need separate parameters for the
-    probability that a site is occupied, the probability that presence is
-    indicated given that a site is occupied, and the probability of detection
-    given that presence is indicated and the site is occupied. We will call
-    these \(\psi\), \(\theta\), and \(p\), respectively.
 
@@
 
+\section{Overview of occupancy modeling}
+
+    \subsection{What is occupancy rate?}
+    Occupancy has two different definitions. It can mean ``the probability that
+    a randomly selected site or sampling unit in an area of interest is occupied
+    by a species'' \cite{MacKenzie2006}. It also can be used to describe ``the
+    proportion of area, patches, or sample units that is occupied''
+    \cite{MacKenzie2006}. Occupancy is not the same as the abundance, or
+    population size, of a species. The number of members of a species is not
+    important for either definition of occupancy.
+
+    While there is a difference between these two definitions of occupancy, in
+    general the probability is unknown and the observed proportion is used as an
+    estimate of the underlying probability. Because of this, these two
+    definitions of occupancy are oftentimes used interchangeably.
+
+    The main reasons that people are interested in measuring occupancy are
+    science or conservation and management \cite{MacKenzie2006}. This statistic
+    can give insight into the way a system works. It also can be used to monitor
+    animals habitats and discover trends over time in groups of animals.
+
+    \subsection{Different ways of sampling data}
+    In order to determine occupancy rate, the data is often collected in the
+    form of a presence-absence survey. This means that observers visit sites and
+    look for the animals themselves, tracks, vocalizations, or some other
+    indicator that the animal is present. In some cases, camera traps and sound
+    recording are used instead of human observers \cite{MacKenzie2006}.
+
+    The main problem with these methods of sampling data is that it is possible
+    for an animal to occupy a site, but not be detected by the observer. This
+    means that the site is recorded as unoccupied, but there really is an animal
+    there. This is called a false negative. Marking an animal present when there
+    really is no animal is also a possible problem, but is thought to be much
+    less likely and is often ignored. We will discuss this issue more in later sections.
+
+    \subsection{Assumptions}
+    As is true for all modeling, there are some assumptions made in most models of occupancy 
+in order to create simple enough models \cite{MacKenzie2006}. One common assumption is no false positives, which means if a
+    site is recorded as occupied we say it must actually be occupied. A second
+    common assumption is that presence and detection probabilities are constant
+    across sites and surveys. This simplifies the estimation of each parameter.
+    A third common assumption is that there is independence of detection between
+    sites. A fourth common assumption is that the surveys are being conducted on
+    a closed population. This means that no animals leave the area or joins the
+    area during the time of the surveys.
+
+    There are different methods for dealing with violations of these assumptions, some 
+more developed than others.  In this paper, we will only talk about attempting to get rid of the
+    assumption of no false positives and the assumption of independence of
+    parameters across sites.
+
+    \subsection{Types of occupancy models}
+    All of the models discussed in this paper are single-species, single-season
+    models. It is possible to modify all of these models to be used for multiple
+    seasons or multiple species, at the cost of added complexity.
+
+    The most simple way to measure occupancy is to simply divide the number of
+    sites where an animal was found to be present by the total number of sites.
+    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
+    accepted that this estimate is too low because animals are not detected at
+    all sites where they are present, due to the issue of false negatives
+    mentioned above. Other occupancy models attempt to do better than this
+    simple estimate.
+
+    One way to create a more accurate model is to attempt to model the biological 
+proccess that is occuring.  In order to do this, we can use a Hierarchical model as a
+    way to think about the parameters that affect occupancy. This allows us to
+    take into account that when an animal is not detected, it could either not
+    be there, or it could be there and just was not detected. This could either
+    be due to a false negative, where the observer did not see the indication --
+    for example they did not see the animal, track, etc. -- or it could be due
+    to the fact that the animal does occupy the site but there was no indication
+    of the animal at the site -- for example, no track, no vocalization, etc.
+
+    Here is a visualization of how the Hierarchical model works:
+
+    \begin{center}
+        \includegraphics[scale=0.75]{SimpleHierarchicalModel}
+    \end{center}
+
+    From this diagram, it is apparent that we need separate parameters for the
+    probability that a site is occupied, the probability that presence is
+    indicated given that a site is occupied, and the probability of detection
+    given that presence is indicated and the site is occupied. We will call
+    these \(\psi\), \(\theta\), and \(p\), respectively.
+
     \subsection{Descriptive statistics}
@@
     \subsection{Bayesian statistics}
-    Bayes' Theorem: \(P(\theta|X) = P(X|\theta)P(\theta) / P(X)\) \\
-
-    This states that the posterior distribution \(\propto\) the likelihood
-    \(\times\) the prior. Usually, the mean of the posterior distribution is
-    then used as a point estimate for the parameter.
-
-    We can use Bayesian statistics in the Hierarchical model as long as we set a
-    prior for every parameter that we include. We can then use Markov Chain
-    Monte Carlo (MCMC) to find a posterior distributions for every parameter. We
-    run into a problem of duality with the probability of correct detection
+	Most occupancy models utitlize maximum likelihood estimation, but we chose 
+	to use a Bayesian approach to model the data for several reasons. 
+	First, the occupancy models developed using maximum likelihood methods 
+	assume that the sampled locations are a random sample from a large, 
+	theoretically infinite, population.  In this study, the entire area of 
+	interest was observed, so this assumption does not hold true.  Bayesian 
+	models do not make such an assumption.  Second, because our model added extra
+	parameters, formulating the likelihood function was cumbersome and maximizing
+	it would have been difficult.  We added an additional random process, track 
+	laying, and also included the possibility of false detection.  A Bayesian
+	model seems much more intuitive and natural.
+
+    Once we had a model formulated, we could then use Markov Chain Monte Carlo 
+	(MCMC) to find a posterior distributions for every parameter. Due to the 
+	nature of occupancy models, certain models can have multiple interpretations.
+	We run into a problem of duality with the probability of correct detection
     \(p\), and the probability of false detection \(E\). These two variables
-    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is the same
-    as \(p=y,E=x,\psi=1-z\). Since we know that is reasonable for \(p\) to be
-    greater than \(E\), we solved this problem by restricting the distribution
-    of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to \(0\text{ to }0.4\),  
-    an approach recommended by Royle and Link \cite{Royle2006}.
+    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is 
+	equivalent in terms of likelihood to \(p=y,E=x,\psi=1-z\) due to the 
+	symmetry of the model \cite{Royle2006}.  Since we know that is reasonable 
+	for \(p\) to be greater than \(E\), we solved this problem by restricting 
+	the distribution of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to 
+	\(0\text{ to }0.4\), an adaptation of the approach recommended by Royle and 
+	Link \cite{Royle2006}.
 
@@
     that the simulated data displayed much more independence than the actual
-    data. We tried modeling spatial correlation in the occupancy process, $\psi
-    $, instead of in the track-laying process, but the simulated data were again 
-    more independent than the real data.
+    data. We would like our simulated data to have the same level of independence 
+	as our actual data.  We tried modeling spatial correlation in the occupancy 
+	process, $\psi$, instead of in the track-laying process, but the simulated 
+	data were again more independent than the real data.
 
@@
 \section{Acknowledgments}
-We would like to thank our adviser, Bob Dobrow, for all of his help and advice and John Fieberg and John Erb from the Minnesota Department of Natural Resources for collecting and sharing the data as well as for offering their specific knowledge of the subject.
+We would like to thank our adviser, Bob Dobrow, for all of his help and advice 
+and John Fieberg and John Erb from the Minnesota Department of Natural Resources
+for collecting and sharing the data as well as for offering their specific 
+knowledge of the subject.
 

### 2010-02-28 Paper.tex
Added some edits to my sections.

--- 2010-02-28 Paper.tex
@@
     which a predetermined number of iterations of the MCMC algorithm are
-    executed \cite{MacKenzie2006}. Because WinBUGS explicitly estimates missing
-    data that is `padded out' by inserting NA values \cite{Spiegelhalter2003},
-    we implement a nested indexing strategy to prevent this extraneous
-    estimation on the actual data. This is a particularly important
-    implementation detail in this case because the data we received from the
-    Minnesota Department of Natural Resources was not consistent with respect to
-    our covariates. Neither the number of observers nor number of observations
-    after each snow event exhibited consistency. Our simulations, on the other
-    hand, involve balanced data sets that are much less difficult to implement
+    executed \cite{MacKenzie2006}.
+    The data we received from the
+    Minnesota Department of Natural Resources was not fixed with respect to
+    our covariates and required us to implement one of the three common
+    strategies used to deal with irregular datasets \cite{Spiegelhalter2003}.
+    Our simulations, on the other
+    hand, ran on datasets with a fixed number of observers, snow events, and
+    observations after each snow event, which are much easier to implement
     within WinBUGS and R.
@@
     \subsection{Our hierarchical model}
-    For our model, we specified uniform \((0, 1)\) prior distributions for
+    For our model, we specified noninformative uniform \((0, 1)\) prior
+    distributions for
     parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
@@
     model are: observer, date, and the number of days since the last snow event.
+    It was necessary to preprocess the Minnesota DNR's data to account for its
+    irregularities.
 
@@
     of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
-    the model's estimates to the simulation's parameters. A table detailing the
-    model's performance lies below:
+    the model's estimates to the simulation's parameters. The model with beta
+    prior distributions on \(p\) and \(E\) performed better than the model with
+    normal prior distributions on the same parameters. A table detailing the
+    model's performance with the beta prior distributions lies below:
 
@@
     \(\theta\) but will either converge well or provide useful estimates for the
-    vast majority of combinations of \(\psi\) and \(\theta\) values.
+    vast majority of combinations of \(\psi\) and \(\theta\) values. 21 of the
+    model's 25 confidence intervals for \(\psi\) include its actual value. With
+    respect to \(p\), all of the model's confidence intervals for each of the
+    three values of \(p\) include its actual value for all \(\psi\) and
+    \(\theta\) values. On the other hand, the model accurately estimates \(E\)
+    for low values of \(\theta\), but consistently overestimates \(E\) when
+    \(\psi\geq0.4\) and \(\theta\geq0.6\) regardless of \(E\)'s actual value.
 
@@
 
-    \subsection{Results}
+    \subsection{Our estimates}
     Below, in tables, lie our estimates for all of the Minnesota DNR's data from

### 2010-02-28 Paper.tex
Edited appendix and table

--- 2010-02-28 Paper.tex
@@
     to compute the mean and standard deviation of \(S\), which we used to
-    calculate a Z-score, where \(Z = (S – E[S])/SD[S]\). In order to compute these values, we needed to find the
+    calculate a Z-score, where \(Z = (\hat{S} - E[S])/SD[S]\). In order to compute these values, we needed to find the
     expected value and variance of each term in \(S\). These caclulations can be
-	found in the appendix.
+	found in the appendix \ref{math}.
 
@@
     \begin{center}
-    \begin{tabular}{|l|l|l|l|l|l|}
-        \hline
-        \multicolumn{6}{|l|}{\textbf{Mississippi River}} \\
+    \begin{tabular}{|l|l|l|l|l|}
+        \hline
+        \multicolumn{5}{|l|}{\textbf{Mississippi River}} \\
         \hline
         \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
-        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} &
-        \multirow{2}{*}{\(E[|z_{all}|-|z_{alt}|]\)} \\
+        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} \\
         \cline{4-5}
-        & & & All Plots & Alt. Plots & \\
-        \hline
-        \multirow{3}{*}{2003} & 402m & \multirow{3}{*}{20} & 7 & 0 & 0.915 \\
-        \cline{2-2} \cline{4-6}
-        & 804m & & 5 & 0 & 0.651 \\
-        \cline{2-2} \cline{4-6}
-        & 1608m & & 1 & 0 & 0.383 \\
-        \hline
-        \multirow{3}{*}{2004} & 402m & \multirow{3}{*}{6} & 5 & 3 & 0.625 \\
-        \cline{2-2} \cline{4-6}
-        & 804m & & 3 & 0 & 1.428 \\
-        \cline{2-2} \cline{4-6}
-        & 1608m & & 2 & 0 & 0.053 \\
+        & & & All Plots & Alt. Plots \\
+        \hline
+        \multirow{3}{*}{2003} & 402m & \multirow{3}{*}{20} & 7 & 0 \\
+        \cline{2-2} \cline{4-5}
+        & 804m & & 5 & 0 \\
+        \cline{2-2} \cline{4-5}
+        & 1608m & & 1 & 0 \\
+        \hline
+        \multirow{3}{*}{2004} & 402m & \multirow{3}{*}{6} & 5 & 3 \\
+        \cline{2-2} \cline{4-5}
+        & 804m & & 3 & 0 \\
+        \cline{2-2} \cline{4-5}
+        & 1608m & & 2 & 0 \\
         \hline
@@
     \begin{center}
-    \begin{tabular}{|l|l|l|l|l|l|}
-        \hline
-        \multicolumn{6}{|l|}{\textbf{Whitewater River}} \\
+    \begin{tabular}{|l|l|l|l|l|}
+        \hline
+        \multicolumn{5}{|l|}{\textbf{Whitewater River}} \\
         \hline
         \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
-        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} &
-        \multirow{2}{*}{\(E[|z_{all}|-|z_{alt}|]\)} \\
+        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} \\
         \cline{4-5}
-        & & & All Plots & Alt. Plots & \\
-        \hline
-        \multirow{2}{*}{2003} & 402m & \multirow{2}{*}{15} & 6 & 2 & 0.722 \\
-        \cline{2-2} \cline{4-6}
-        & 804m & & 4 & 1 & 0.634 \\
-        \hline
-        \multirow{2}{*}{2004} & 402m & \multirow{2}{*}{6} & 3 & 1 & 1.189 \\
-        \cline{2-2} \cline{4-6}
-        & 804m & & 2 & 2 & 0.612 \\
+        & & & All Plots & Alt. Plots \\
+        \hline
+        \multirow{2}{*}{2003} & 402m & \multirow{2}{*}{15} & 6 & 2 \\
+        \cline{2-2} \cline{4-5}
+        & 804m & & 4 & 1 \\
+        \hline
+        \multirow{2}{*}{2004} & 402m & \multirow{2}{*}{6} & 3 & 1 \\
+        \cline{2-2} \cline{4-5}
+        & 804m & & 2 & 2 \\
         \hline
@@
     \begin{center}
-    \begin{tabular}{|l|l|l|l|l|l|}
-        \hline
-        \multicolumn{6}{|l|}{\textbf{Zumbro River}} \\
+    \begin{tabular}{|l|l|l|l|l|}
+        \hline
+        \multicolumn{5}{|l|}{\textbf{Zumbro River}} \\
         \hline
         \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
-        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} &
-        \multirow{2}{*}{\(E[|z_{all}|-|z_{alt}|]\)} \\
+        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} \\
         \cline{4-5}
-        & & & All Plots & Alt. Plots & \\
-        \hline
-        \multirow{2}{*}{2003} & 402m & \multirow{2}{*}{11} & 10 & 6 & 1.883 \\
-        \cline{2-2} \cline{4-6}
-        & 804m & & 8 & 5 & 1.287 \\
-        \hline
-        \multirow{2}{*}{2004} & 402m & \multirow{2}{*}{4} & 3 & 3 & 2.425 \\
-        \cline{2-2} \cline{4-6}
-        & 804m & & 3 & 1 & 2.343 \\
+        & & & All Plots & Alt. Plots \\
+        \hline
+        \multirow{2}{*}{2003} & 402m & \multirow{2}{*}{11} & 10 & 6 \\
+        \cline{2-2} \cline{4-5}
+        & 804m & & 8 & 5 \\
+        \hline
+        \multirow{2}{*}{2004} & 402m & \multirow{2}{*}{4} & 3 & 3 \\
+        \cline{2-2} \cline{4-5}
+        & 804m & & 3 & 1 \\
         \hline
@@
 \appendix
+	\section{Deriving \(E[S]\) and \(SD[S]\)}
+	\label{math}
 	We let \(A = \text{\# of

### 2010-02-28 Presentation/Presentation.tex
Edited my sections
Presentation edits

@@
 \usepackage{graphpap,graphics}
-%\includeonlyframes{}
+\includeonlyframes{currentslide}
 
@@
 	\begin{itemize}
-		\item The probability that a randomly selected site or sampling unit in an area of interest is occupied by a species
-		\item The proportion of area, patches, or sampling units that is occupied 	\end{itemize}
+		\item The probability that a randomly selected site or sampling unit in 
+		an area of interest is occupied by a species
+		\item The proportion of area, patches, or sampling units that is 
+		occupied 	
+	\end{itemize}
 	(MacKenzie, et al. 2006) 
@@
 		\end{itemize}
-		\item Need to be able to estimate probability of observing the species given its presence
+		\item Need to be able to estimate probability of observing the species 
+		given its presence
 	\end{itemize}
@@
 		\item No false positives
-		\item Presence and detection probabilities are constant across sites and surveys (or can be modeled by covariates)
+		\item Presence and detection probabilities are constant across sites 
+		and surveys (or can be modeled by covariates)
 		\item Independence of detection between sites
@@
 		\item $\psi=$ The probability a track is occupied
-		\item $\theta=$ The probability a track is laid down in a site on a given day, given that the site is occupied
-		\item $p= $ The probability an observer sees a track given that a track has been made
+		\item $\theta=$ The probability a track is laid down in a site on a 
+		given day, given that the site is occupied
+		\item $p= $ The probability an observer sees a track given that a track 
+		has been made
 	\end{itemize}
@@
 		%include graph here
-		\item Solution: use observer indicators as covariates for detection probability
+		\item Solution: use observer indicators as covariates for detection 
+		probability
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Two observers on same day should have the same na\"ive occupancy rates; not the case
+		\item Two observers on same day should have the same na\"ive occupancy 
+		rates; not the case
 		\item Also looked at correlation between flights
@@
 
-%Include frame with new hierarchical diagram
+\begin{frame}{Hierarchical model with false detection}
+	\begin{center}
+		\resizebox{10cm}{!}{\includegraphics{../ModelWithE.pdf}}
+	\end{center}
+\end{frame}
 
@@
 		\item Assumption: independence between plots
-		\item %Sarah going to review
+		\item $S=(\text{\# of successive 0-1s})-(\text{\# of successive 1-1s})$
+		\item Calculated $Z$-scores for each flight
+			\begin{itemize}
+				\item $Z=\frac{\hat{S}-E[S]}{SD[S]}$
+				\item $|Z_i|>2$ suggests assumption violated for flight $i$.
+			\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}[label=currentslide]{Independence, continued}
+
+	\begin{itemize}
+		\item Looking at alternating plots increases the independence
+		\item Also saw trends with respect to plot size
 		\item Solution: kept issue in mind, explored spatially explicit models
@@
 \section{Modeling the data}
-\begin{frame}{Bayesian statistics}
+\begin{frame}{Introduction to Bayesian statistics}
+	\begin{itemize}
+		\item Under Bayesian statistics, parameters are random variables
+		\item According to Bayes' theorem
+		\begin{itemize}
+			\item $\text{posterior distribution} \propto \text{prior 
+			distribution}\times \text{likelihood}$
+		\end{itemize}
+	\end{itemize}
 \end{frame}
@@
 
-\section{Discussion}
-\begin{
-
 \end{document}

### 2010-03-01 Paper.tex
Changed interval of p in CAR section
Tons of edits.

@@
 \title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
-Population in Southeastern Minnesota}
-\date{}
+Population in Southeastern Minnesota \\
+{\large Senior Comps Project}}
 \author{Chrisna Aing, Sarah Halls, Kiva Oken}
@@
 where hunters would like to be able to trap otters again. In order for that to
-be possible governmental organizations must be able to monitor the population
-health. In addition, otter presence can indicate high quality aquatic habitat.
-Due to these reasons, there is interest in monitoring changes in otter abundance
+be possible, governmental organizations must be able to monitor the health of
+the population.
+In addition, otter presence can indicate the quality of the aquatic habitat.
+Due to these reasons, there is interest in monitoring changes in otter
+population
 over time. Unfortunately, otters are not easy to observe. Thus, this study
 investigated the potential of using wintertime aerial surveys of otter tracks on
-rivers to make inferences about the abundance of otters and how it changes over
-time. We used the concepts from occupancy theory to develop models to estimate
-the occupancy rate in order to get a sense of otter abundance.
+rivers to make inferences about the population of otters and how it changes over
+time. We used concepts from occupancy theory to develop models to estimate
+the occupancy rate in order to get a sense of otter population.
 
@@
     river on the same day. Surveys were conducted during multiple snow events
-    throughout both winters. The rivers were then divided into 400 and 800 meter
+    throughout both winters. The rivers were then divided into 402 and 804 meter
     plots and the presence (1) or absence (0) of a waypoint in a plot on a given
-    flight was determined. The Mississippi River was also divided into 1600
-    meter plots.
+    flight was determined. The Mississippi River was also divided into 1608
+    meter plots.  For each river and plot size, the data set with every other
+    plot removed was considered.  We refer to this as ``alternating plots''.
 
@@
     the boundary between two plots, a person on one flight may have recorded the
-    track in one site whereas someone on a later flight during that same snow
+    track in one plot whereas someone on a later flight during that same snow
     event may have seen the same track but recorded it in the neighboring plot.
@@
     In addition, there is concern that the probability of an otter laying down a
-    track in one site is not independent of the probability that an otter lays
-    down a track in neighboring sites, violating a principle assumption of
+    track in one plot is not independent of the probability that an otter lays
+    down a track in neighboring plots, violating a principle assumption of
     occupancy theory. That is, a single track can extend across multiple plots.
@@
 
-    The main reasons that people are interested in measuring occupancy are
-    science or conservation and management \cite{MacKenzie2006}. This statistic
-    can give insight into the way a system works. It also can be used to monitor
-    animals habitats and discover trends over time in groups of animals.
-
-    \subsection{Different ways of sampling data}
-    In order to determine occupancy rate, the data is often collected in the
-    form of a presence-absence survey. This means that observers visit sites and
-    look for the animals themselves, tracks, vocalizations, or some other
-    indicator that the animal is present. In some cases, camera traps and sound
-    recording are used instead of human observers \cite{MacKenzie2006}.
-
-    The main problem with these methods of sampling data is that it is possible
-    for an animal to occupy a site, but not be detected by the observer. This
-    means that the site is recorded as unoccupied, but there really is an animal
-    there. This is called a false negative. Marking an animal present when there
-    really is no animal is also a possible problem, but is thought to be much
-    less likely and is often ignored. We will discuss this issue more in later sections.
-
     \subsection{Assumptions}
-    As is true for all modeling, there are some assumptions made in most models of occupancy 
-in order to create simple enough models \cite{MacKenzie2006}. One common assumption is no false positives, which means if a
-    site is recorded as occupied we say it must actually be occupied. A second
+    As is true for all modeling, there are some assumptions made in most models
+    of occupancy
+    in order to create simple enough models \cite{MacKenzie2006}. One common
+    assumption is no false positives, which means if a
+    plot is recorded as occupied we say it must actually be occupied. A second
     common assumption is that presence and detection probabilities are constant
-    across sites and surveys. This simplifies the estimation of each parameter.
+    across plots and surveys. This simplifies the estimation of each parameter.
     A third common assumption is that there is independence of detection between
-    sites. A fourth common assumption is that the surveys are being conducted on
-    a closed population. This means that no animals leave the area or joins the
+    % Question: Independence of occupancy as an assumption as well?
+    plots. A fourth common assumption is that the surveys are being conducted on
+    a closed population. This means that no animals leave or join the
     area during the time of the surveys.
 
-    There are different methods for dealing with violations of these assumptions, some 
-more developed than others.  In this paper, we will only talk about attempting to get rid of the
+    There are different methods for dealing with violations of these
+    assumptions, some more developed than others.  In this paper, we will only
+    talk about attempting to eliminate the
     assumption of no false positives and the assumption of independence of
-    parameters across sites.
+    parameters across plots.
 
@@
 
-    The most simple way to measure occupancy is to simply divide the number of
-    sites where an animal was found to be present by the total number of sites.
+    The most simple way to measure occupancy is to divide the number of
+    plots where an animal was found to be present by the total number of plots.
     This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
-    accepted that this estimate is too low because animals are not detected at
-    all sites where they are present, due to the issue of false negatives
-    mentioned above. Other occupancy models attempt to do better than this
-    simple estimate.
-
-    One way to create a more accurate model is to attempt to model the biological 
-proccess that is occuring.  In order to do this, we can use a Hierarchical model as a
-    way to think about the parameters that affect occupancy. This allows us to
-    take into account that when an animal is not detected, it could either not
-    be there, or it could be there and just was not detected. This could either
-    be due to a false negative, where the observer did not see the indication --
-    for example they did not see the animal, track, etc. -- or it could be due
-    to the fact that the animal does occupy the site but there was no indication
-    of the animal at the site -- for example, no track, no vocalization, etc.
-
-    Here is a visualization of how the Hierarchical model works:
-
+    accepted that this estimate is too low because observers usually do not
+    detect animals at
+    all plots where they are present.
+
+    One way to create a more accurate model is to attempt to model the
+    biological process that is occurring.  In order to do this, we can use a
+    hierarchical model as a
+    way to think about the parameters that affect occupancy. A plot can be
+    designated as unoccupied for two reasons.
+    There could either be
+    a false negative, where the observer did not see the track,
+    or
+    the animal may occupy the plot but did not leave a track before a survey.
+
+    \begin{figure}
     \begin{center}
         \includegraphics[scale=0.75]{SimpleHierarchicalModel}
+        \caption{A visualization of how the hierarchical model works.}
+        \label{simplehierarchicalmodel}
     \end{center}
-
-    From this diagram, it is apparent that we need separate parameters for the
-    probability that a site is occupied, the probability that presence is
-    indicated given that a site is occupied, and the probability of detection
-    given that presence is indicated and the site is occupied. We will call
+    \end{figure}
+
+    Figure \ref{simplehierarchicalmodel} shows that we need separate parameters
+    for the
+    probability that a plot is occupied, the probability that a track is laid
+    down given that a plot is occupied, and the probability of detection
+    given that a track is laid down and the plot is occupied. We will call
     these \(\psi\), \(\theta\), and \(p\), respectively.
+
+\section{Exploring the data}
 
@@
     classifying in which plot the observer saw the actual track, each observer
-    should at least be seeing the same, or nearly the same, number of sites with
+    should at least be seeing the same, or nearly the same, number of plots with
     tracks. However, we did not find that to be the case. If we look only at
     days that all three people observed, there appear to be major differences
-    among the proportion of sites in which the observers detected a track
+    among the proportion of plots in which the observers detected a track
     (Figure \ref{obsPlots}). The difference in estimated na\"ive occupancy among
@@
         \includegraphics[width=5in]{observerPlots.pdf}
-	    \caption{The box plots show how the proportion of sites in which an 
+        \caption{The box plots show how the proportion of plots in which an
         observer saw tracks (``occupancy'') differs among the three observers in
@@
     Correlation among the flight sequences is low. For example, average
-    correlation between flights on the Mississippi in 2003 is 0.176 for 400
+    correlation between flights on the Mississippi in 2003 is 0.176 for 402
     meter plots. Some of this is due to the difficulties in determining to which
-    site each track belongs. However, we have also found quite a bit of
+    plot each track belongs. However, we have also found quite a bit of
     variation in the data not necessarily due to these classification issues.
@@
     flights on the same day or flights during the same snow event, but still not
-    as high as one might hope. Correlation rarely exceeds 0.5. The proportion of 
-    sites with tracks detected
+    as high as one might expect. Correlation rarely exceeds 0.5. The proportion
+    of plots with tracks detected
     generally increases with days since snow, likely because the otters have a
@@
     false detection process can occur as long as there is no track, whether or
-    not the site is occupied.
+    not the plot is occupied.
 
@@
     Since one of the assumptions of occupancy modeling is that there is
-	independence between sites, we wanted to see if this assumption held true 
+    independence between plots, we wanted to see if this assumption held true
     for our data.  In order to do this, we calculated the statistic \(S\), which
-	we define as the number of successive \(0\), \(1\)'s minus the number of
-    successive \(1\), \(1\)'s per flight. Under the null hypothesis that each
-    site is an independent Bernoulli trial with probability of a \(1 = p\) we were able
+    we define as the number of \(0\)--\(1\)'s minus the number of
+    \(1\)--\(1\)'s per flight. Under the null hypothesis that each
+    plot is an independent Bernoulli trial where \(p\) is the probability of a
+    \(1\) we were able
     to compute the mean and standard deviation of \(S\), which we used to
-    calculate a Z-score, where \(Z = (\hat{S} - E[S])/SD[S]\). In order to compute these values, we needed to find the
-    expected value and variance of each term in \(S\). These caclulations can be
-	found in the appendix \ref{math}.
+    calculate a Z-score, where \(Z = (\hat{S} - E[S])/SD[S]\). In order to
+    compute these values, we needed to find the
+    expected value and variance of each term in \(S\). These calculations can be
+    found in Appendix \ref{math}.
 
     Simulations of \(S\) show that the distribution of \(S\) is roughly normal.
-    So using our calculations for the mean and the standard deviation, we
+    Using our calculations for the mean and the standard deviation, we
     computed a Z-score for each flight, where we use a plug-in estimate for
-    \(p\) to be the proportion of \(1\)'s per flight. We said that \(|Z| > 2\) suggests that the assumption of
-    independence between sites is violated for that flight. We found that, in
-    general, alternating sites are more "independent" than all sites as measured
-    by this statistic. However, alternating sites do not guarantee independence.
+    \(p\) to be the proportion of \(1\)'s per flight. \(|Z| > 2\) suggests that
+    the assumption of
+    independence between plots is violated for that flight. We found that, in
+    general, data with alternating plots are more ``independent'' than all plots
+    as measured
+    by this statistic. However, alternating plots do not guarantee independence.
     For instance, six of eleven flights along the Zumbro River during 2003 still
-    have significant Z-scores with alternating sites. It also appears that there 
-	may be slightly more independence when larger site sizes are used, although it is
-not clear if this trend is significant.  There 
-	were no noticable trends for this statistic with respect to observer, days 
+    have significant Z-scores with alternating plots (Table \ref{Z-scores}). It
+    also appears that there
+    may be slightly more independence when larger plot sizes are used, although
+    it is not clear if this trend is significant. There
+    were no noticeable trends for this statistic with respect to observer, days
     since snow, or snow event.
@@
     % Mississippi River!
+    \begin{table}
+    \caption{Summary of \(\hat{S}\) Z-scores}
+    \label{Z-scores}
     \begin{center}
@@
     \end{center}
+    \end{table}
 
     We think that the lack of independence results from strings of ones where an
-    otter left a track across multiple sites. When alternating sites are used,
-    the strings are cut in half and the occurrence of \(1\), \(1\)'s is lower
-    while the occurrence of \(0\), \(1\)'s stays about the same. It is
+    otter left a track across multiple plots. When alternating plots are used,
+    the strings are cut in half and the occurrence of \(1\)--\(1\)'s is lower
+    while the occurrence of \(0\)--\(1\)'s stays about the same. It is
     noteworthy that virtually all the violations of independence are in the
-    direction of more \(1\), \(1\)'s than would be expected. This reinforces our
-    thoughts about successive \(1\)'s representing the same otter. We will
-    discuss later whether this violation of independence actually matters in our
+    direction of more \(1\)--\(1\)'s than would be expected. We believe that
+    successive \(1\)'s could represent a single track made by the same otter. We
+    will
+    discuss later whether this violation of independence matters in our
     model, and how this might affect the estimation of actual occupancy rates.
@@
     \subsection{Bayesian statistics}
-	Most occupancy models utitlize maximum likelihood estimation, but we chose 
+    Most occupancy models utilize maximum likelihood estimation, but we chose
     to use a Bayesian approach to model the data for several reasons.
@@
     interest was observed, so this assumption does not hold true.  Bayesian
-	models do not make such an assumption.  Second, because our model added extra
-	parameters, formulating the likelihood function was cumbersome and maximizing
+    models do not make such an assumption.  Second, because our model added
+    extra
+    parameters, formulating the likelihood function was cumbersome and
+    maximizing
     it would have been difficult.  We added an additional random process, track
@@
     Once we had a model formulated, we could then use Markov Chain Monte Carlo
-	(MCMC) to find a posterior distributions for every parameter. Due to the 
-	nature of occupancy models, certain models can have multiple interpretations.
-	We run into a problem of duality with the probability of correct detection
+    (MCMC) to find a posterior distribution for every parameter. Due to the
+    nature of occupancy models, certain models can have multiple
+    interpretations.
+    We run into one such duality with the probability of correct detection
     \(p\), and the probability of false detection \(E\). These two variables
-    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is 
-	equivalent in terms of likelihood to \(p=y,E=x,\psi=1-z\) due to the 
+    have bimodal posterior distributions, because \((p=x,E=y,\psi=z)\) is
+    equivalent in terms of likelihood to \((p=y,E=x,\psi=1-z)\) due to the
     symmetry of the model \cite{Royle2006}.  Since we know that is reasonable
     for \(p\) to be greater than \(E\), we solved this problem by restricting
-	the distribution of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to 
-	\(0\text{ to }0.4\), an adaptation of the approach recommended by Royle and 
+    the distribution of \(p\) to the range \((0.6,1)\), and \(E\) to
+    \((0,0.4)\), an adaptation of the approach recommended by Royle and
     Link \cite{Royle2006}.
@@
     our covariates and required us to implement one of the three common
-    strategies used to deal with irregular datasets \cite{Spiegelhalter2003}.
+    strategies used to deal with irregular data sets \cite{Spiegelhalter2003}.
     Our simulations, on the other
-    hand, ran on datasets with a fixed number of observers, snow events, and
+    hand, ran on data sets with a fixed number of observers, snow events, and
     observations after each snow event, which are much easier to implement
@@
     burn-in time for the MCMC simulation to 10,000 iterations. Each of these
-    parameters are tracked and therefore are estimated. The covariates of this
-    model are: observer, date, and the number of days since the last snow event.
+    parameters are tracked for 10,000 additional iterations and are estimated
+    by the mean of the posterior distribution. The covariates of this
+    model are the observer and the number of days since the last snow event.
     It was necessary to preprocess the Minnesota DNR's data to account for its
@@
     \subsection{Simulated data}
-    We simulated both balanced and unbalanced datasets meant for the standard
+    We simulated both balanced and unbalanced data sets meant for the standard
     hierarchical model with given values of \(\psi\), \(\theta\), \(p\) for each
@@
     To simulate balanced data, we provided the following additional variables:
-    the number of sites, the number of snow events, the number of observers, and
+    the number of plots, the number of snow events, the number of observers, and
     the number of observations following each snow event. These variables are
@@
     greatly simplifies the accompanying hierarchical model. First, the
-    simulation determines a site's occupancy based upon the given \(psi\) value.
+    simulation determines a plot's occupancy based upon the given \(\psi\)
+    value.
     Then, for each day after each snow event, the simulation decides whether or
@@
     of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
-    the model's estimates to the simulation's parameters. The model with beta
+    the model's estimates to the simulation's parameters (Table \ref{Simulated
+    data}). The model with beta
     prior distributions on \(p\) and \(E\) performed better than the model with
-    normal prior distributions on the same parameters. A table detailing the
-    model's performance with the beta prior distributions lies below:
-
+    normal prior distributions on the same parameters.  Therefore, we limit
+    further discussion to the beta model.
+
+    \begin{table}
+    \caption{Beta model estimates from simulated data. All simulations
+    ran with 70 plots, 5 snow events, 3 observers, and for 3 days after each
+    snow event. The 3 observers had \(p\) values of 0.70, 0.80, 0.90 and \(E\)
+    values of 0.05, 0.20, and 0.30, respectively. Green rows represent values of
+    \(\psi\) and \(\theta\) for which the model exhibits good convergence and
+    the 95\% confidence interval captures the actual \(\psi\) value.  Yellow
+    rows represent values for which our model exhibits either mediocre
+    convergence or a bad confidence interval.  Red rows represent values for
+    which our model does not converge.}
+    \label{Simulated data}
     \begin{center}
     \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{4}{|l|}{\textbf{The Model vs. Simulated Data}} \\
         \hline
@@
     \end{center}
-
-    Each of the simulations represented in the table above ran with the
-    following parameters: seventy sites, five snow events, three observers, and
-    three observations after each snow event. The three observers had \(p\)
-    values of 0.70, 0.80, 0.90 and \(E\) values of 0.05, 0.20, and 0.30,
-    respectively.
-
-    Green rows represent values of \(\psi\) and \(\theta\) for which our
-    standard hierarchical model exhibits good convergence and provides a 95\%
-    confidence interval for \(\psi\) that includes the simulated data's actual
-    \(\psi\) value.  Yellow rows represent values of \(\psi\) and \(\theta\) for
-    which our model exhibits either mediocre convergence with a good confidence
-    interval or good convergence with a bad confidence interval.  Red rows
-    represent values of \(\psi\) and \(\theta\) for which our model does not
-    converge.  The 95\% confidence interval is ignored in these cases.
+    \end{table}
 
@@
     \(\theta\) but will either converge well or provide useful estimates for the
-    vast majority of combinations of \(\psi\) and \(\theta\) values. 21 of the
-    model's 25 confidence intervals for \(\psi\) include its actual value. With
-    respect to \(p\), all of the model's confidence intervals for each of the
+    vast majority of combinations of \(\psi\) and \(\theta\) values. Twenty-one
+    of the
+    model's 25 confidence intervals for \(\psi\) include its actual value.
+    The model's confidence intervals for each of the
     three values of \(p\) include its actual value for all \(\psi\) and
@@
     \(\psi\geq0.4\) and \(\theta\geq0.6\) regardless of \(E\)'s actual value.
+    This may help to explain why the model does not estimate \(\psi\) as
+    successfully for higher values of \(\psi\) and \(\theta\).
 
@@
 One approach to dealing with the correlation in the probability of
-observing a track in neighboring sites was to use a conditional autoregressive
+observing a track in neighboring plots is to use a conditional autoregressive
 (CAR) prior distribution for $\theta$, the probability that an otter lays down a
-track in a site on a given day.
+track in a plot on a given day.
 
     \subsection{Conditional autoregressive (CAR) models}
-    Let $\theta_s$ be the probability of a track being laid down in site $s$ on
+    Let $\theta_s$ be the probability of a track being laid down in plot $s$ on
     a given day. Then the Gaussian CAR model can be defined using the following
@@
     where $N_s$ is the set of neighbors of area $s$ (in this case, the two
-    bordering sites), $E(\theta_s)=\mu_s$, $\tau_s^2$ is the conditional
+    bordering plots), $E(\theta_s)=\mu_s$, $\tau_s^2$ is the conditional
     variance, $c_{si}$ is the weight defining how strong the correlation is
-    between sites $s$ and $i$, and $c_{ii}=0$ for all sites $i$ \cite{Arab2008}.
-    Essentially, this model means that if site $s$'s neighbor has a
+    between plots $s$ and $i$, and $c_{ii}=0$ for all plots $i$ \cite{Arab2008}.
+    Essentially, this model means that if plot $s$'s neighbor has a
     particularly high or low track-laying probability, that will increase or
-    decrease the track-laying probability for site $s$. This makes it more
-    likely to have strings of tracks that cross over multiple sites.
+    decrease the track-laying probability for plot $s$. This makes it more
+    likely to have strings of tracks that cross over multiple plots.
 
@@
     For our model, both neighbors are weighted equally. Since $\tau_s$ for any
-    site $s$ is unknown, we followed convention and used the same value of
-    $\tau$ for all sites and placed a prior distribution on it. We chose a
+    plot $s$ is unknown, we followed convention and used the same value of
+    $\tau$ for all plots and placed a prior distribution on it. We chose a
     $\gamma(0.5,0.0005)$ prior distribution for $(\tau^2)^{-1}$, a common choice
@@
     distribution on $\psi$ and the same beta prior distributions on $p$ and $E$.
-    In this case, we limited the range of $p$ to the interval $(0.7,1)$ and the
+    In this case, we limited the range of $p$ to the interval $(0.6,1)$ and the
     range of $E$ to $(0,0.3)$. We increased the burn-in time to 12,000
     iterations. With this model, instead of a single estimate for $\theta$ there
-    are now $S$ estimates for $\theta$ where $S$ is the total number of sites.
+    are now $S$ estimates for $\theta$ where $S$ is the total number of plots.
     While ideally we would estimate an overall value for $\theta$, it seems more
@@
     \subsection{Simulating spatially correlated data}
+    \label{spatial simulation}
     Because we modeled spatial correlation in $\theta$, we began by simulating
@@
     simulation method, $\theta$ is the probability of an otter laying down a
-    track in site $i$ given site $i$ is occupied and site $i-1$ does
-    \textit{not} have a track. If site $i$ is occupied and site $i-1$ does have
-    a track, then the probability of an otter laying down a track in site $i$ is
-    equal to $\theta+\alpha$. This means if there is a track in one site, there
-    is more likely to be a track in the next site. However, when we ran data
-    simulated using this new method on the model described in section \ref{car
+    track in plot $i$ given plot $i$ is occupied and plot $i-1$ does
+    \textit{not} have a track. If plot $i$ is occupied and plot $i-1$ does have
+    a track, then the probability of an otter laying down a track in plot $i$ is
+    equal to $\theta+\alpha$. This means if there is a track in one plot, there
+    is more likely to be a track in the next plot. However, when we ran data
+    simulated using this new method on the model described in Section \ref{car
     model}, the model almost never converged. We tested independence of the data
-    using the $\hat{s}$ statistic we developed (section \ref{shat}) and found
+    using the $\hat{S}$ statistic we developed (Section \ref{shat}) and found
     that the simulated data displayed much more independence than the actual
-    data. We would like our simulated data to have the same level of independence 
+    data. We would like our simulated data to have the same level of
+    independence
     as our actual data.  We tried modeling spatial correlation in the occupancy
@@
 
-    To increase the correlation between sites we simulated data
+    To increase the correlation between plots we simulated data
     correlated in both the track-laying and occupancy processes. Under this 
@@
     variables $ \psi$ and $\beta$. In this case, $\psi$ is the probability an
-    otter occupies site $i$ given that site $i-1$ is not occupied and $\psi+
-    \beta$ is the probability that an otter occupies site $i$ given that site
+    otter occupies plot $i$ given that plot $i-1$ is not occupied and $\psi+
+    \beta$ is the probability that an otter occupies plot $i$ given that plot
     $i-1$ is occupied. With this increased spatial correlation we were able to
-    more closely match the $\hat{s}$ $z$-scores found in the actual data. In
+    more closely match the $\hat{S}$ Z-scores found in the actual data. In
     addition, the CAR model converged more often with data simulated using this
-    method, but it still did not converge consistently.
+    method, but it still did not converge consistently. When it did converge,
+    estimates were comparable to those found using the standard model on the
+    same simulated data.
 
@@
     sense of the overall occupancy rate in the simulated data we counted the
-    number of sites that were occupied (known because the data are simulated)
-    and divided that by the total number of sites. This proportion is the value
+    number of plots that were occupied (known because the data are simulated)
+    and divided that by the total number of plots. This proportion is the value
     we compared to the $\psi$ estimate from both hierarchical models.
@@
     \subsection{Obtaining convergence on the data}
-    Recall that our standard hierarchical model contains the following
-    parameters: \(\psi\) (the rate of occupancy), \(\theta\) (the probability of
-    laying down a track given occupancy), \(p\) (the probability of detecting a
-    track given that it is present, which is unique to each observer and makes
-    \(1-p\) the false negative rate of detecting a track), and \(E\) (the false
-    positive rate of detecting a track, which is unique to each observer).
-
     Once both the standard hierarchical model and the addition of a CAR prior
     distribution on \(\theta\) to the standard model exhibited convergence on
-    our simulated data, we moved onto running both models on the data given to
-    us by the Minnesota DNR. Recall that data was gathered during the winters of
-    2003 and 2004 along the Mississippi, Whitewater, and Zumbro Rivers and
-    aggregated with ArcGIS into vectors containing ones and zeroes representing
-    the detection and non-detection of a track, respectively, within plots of
-    length 402 meters, 804 meters, and 1608 meters. The covariates recorded
-    were: the observer, the date, and the number of days since the last snow
-    event. In general, the sparseness of the data gathered in 2004 caused us to
+    our simulated data, we moved onto running both models on the data.
+    In general, the sparseness of the data gathered in 2004 caused us to
     simplify the model in order to obtain results. Unfortunately, the
@@
     \(p\) and \(E\), \(p\) for each observer, and \(E\) for each observer), the
-    CAR model estimates \(\theta\) for each site. Because there are at least
-    thirty-five sites in each data set (regardless of plot size), the CAR prior
+    CAR model estimates \(\theta\) for each plot. Because there are at least
+    thirty-five plots in each data set (regardless of plot size), the CAR prior
     distribution adds a considerable amount of complexity to the model.
@@
 
-    \subsection{Our estimates}
-    Below, in tables, lie our estimates for all of the Minnesota DNR's data from
-    2003 and 2004. Below each table is detailed commentary about the ability of
-    the standard hierarchical model to converge, given the corresponding data.
-
+    % Mississippi River!
+    \begin{table}
+    \caption{Mississippi River. 2003 data converged well at the 402m and 804m
+    plot sizes. At the 1608m plot size, \(E\), \(\psi\), and \(\theta\) did not
+    converge well. 2004 data did not converge. We do not think that these
+    estimates are accurate, but they are included for the sake of completeness.}
+    \label{Mississippi}
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|}
+        \hline
+        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
+        \hline
+            & 402m & 804m & 1608m \\
+        \hline
+        \multirow{2}{*}{E[Danny]}
+            & 0.01954 & 0.04808 & 0.06885 \\
+            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
+        \hline
+        \multirow{2}{*}{E[John]}
+            & 0.05216 & 0.07771 & 0.13650 \\
+            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
+        \hline
+        \multirow{2}{*}{E[Tom]}
+            & 0.02668 & 0.06140 & 0.10230 \\
+            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
+        \hline
+        \multirow{2}{*}{p[Danny]}
+            & 0.65240 & 0.67340 & 0.74960 \\
+            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
+        \hline
+        \multirow{2}{*}{p[John]}
+            & 0.67770 & 0.77580 & 0.82540 \\
+            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
+        \hline
+        \multirow{2}{*}{p[Tom]}
+            & 0.53670 & 0.54220 & 0.65330 \\
+            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
+        \hline
+        \multirow{2}{*}{\(\psi\)}
+            & 0.59380 & 0.69370 & 0.75820 \\
+            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
+        \hline
+        \multirow{2}{*}{\(\theta\)}
+            & 0.15280 & 0.20730 & 0.27690 \\
+            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
+        \hline
+    \end{tabular}
+    \end{center}
+
+    \begin{center}
+    \begin{tabular}{|l|l|}
+        \hline
+        \multicolumn{2}{|l|}{\textbf{Mississippi River, 2004}} \\
+        \hline
+            & 402m \\
+        \hline
+        \multirow{2}{*}{p[Danny]}
+            & 0.3129 \\
+            & (0.2174, 0.8107) \\
+        \hline
+        \multirow{2}{*}{\(\psi\)}
+            & 0.7613 \\
+            & (0.6609, 0.8833) \\
+        \hline
+        \multirow{2}{*}{\(\theta\)}
+            & 0.8511 \\
+            & (0.2426, 0.9973) \\
+        \hline
+    \end{tabular}
+    \end{center}
+    \end{table}
+
+    % Whitewater River!
+    \begin{table}
+    \caption{Whitewater River. In 2003, while the 402m data converged well,
+    \(E\), \(\psi\), and \(\theta\) did not converge well for the 804m data.
+    2004 data converged well. However, we feel that the absence of \(E\) from
+    the model makes the estimates much less dependable. The estimates are also
+    much less precise than those from 2003 -- the confidence intervals of
+    \(\psi\) and of \(p[Danny]\) are twice as large as those from 2003.}
+    \label{Whitewater}
+    \begin{center}
+    \begin{tabular}{|l|l|l|}
+        \hline
+        \multicolumn{3}{|l|}{\textbf{Whitewater River, 2003}} \\
+        \hline
+            & 402m & 804m \\
+        \hline
+        \multirow{2}{*}{E[Danny]}
+            & 0.05608 & 0.0555 \\
+            & (0.02474, 0.09554) & (0.002859, 0.1390) \\
+        \hline
+        \multirow{2}{*}{E[John]}
+            & 0.16570 & 0.1931 \\
+            & (0.10130, 0.23510) & (0.065910, 0.3384) \\
+        \hline
+        \multirow{2}{*}{E[Tom]}
+            & 0.15350 & 0.2241 \\
+            & (0.09758, 0.21660) & (0.110000, 0.3623) \\
+        \hline
+        \multirow{2}{*}{p[Danny]}
+            & 0.66990 & 0.6985 \\
+            & (0.56080, 0.79040) & (0.573100, 0.8366) \\
+        \hline
+        \multirow{2}{*}{p[John]}
+            & 0.92040 & 0.9877 \\
+            & (0.84270, 0.97590) & (0.945100, 1.0000) \\
+        \hline
+        \multirow{2}{*}{p[Tom]}
+            & 0.75910 & 0.8575 \\
+            & (0.65060, 0.85800) & (0.745400, 0.9600) \\
+        \hline
+        \multirow{2}{*}{\(\psi\)}
+            & 0.87590 & 0.9126 \\
+            & (0.72910, 0.98880) & (0.746900, 0.9955) \\
+        \hline
+        \multirow{2}{*}{\(\theta\)}
+            & 0.24100 & 0.3779 \\
+            & (0.17890, 0.30900) & (0.286500, 0.4721) \\
+        \hline
+    \end{tabular}
+    \end{center}
+
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|}
+        \hline
+        \multicolumn{2}{|l|}{\textbf{Whitewater River, 2004}} \\
+        \hline
+            & 402m \\
+        \hline
+        \multirow{2}{*}{p[Danny]}
+            & 0.8237 \\
+            & (0.5323, 0.9804) \\
+        \hline
+        \multirow{2}{*}{\(\psi\)}
+            & 0.6066 \\
+            & (0.4246, 0.8382) \\
+        \hline
+        \multirow{2}{*}{\(\theta\)}
+            & 0.2068 \\
+            & (0.1211, 0.3550) \\
+        \hline
+    \end{tabular}
+    \end{center}
+    \end{table}
+
+    % Zumbro River!
+    \begin{table}
+    \caption{Zumbro River. In 2003, the 402m data converged well. While \(\psi\)
+    and \(\theta\) converged well for the 804m data, none of the other
+    parameters did. 2004 data converged well. However, as is the case with all
+    of the 2004 data, we feel that the absence of \(E\) from the model makes the
+    estimates much less dependable.}
+    \label{Zumbro}
+    \begin{center}
+    \begin{tabular}{|l|l|l|}
+        \hline
+        \multicolumn{3}{|l|}{\textbf{Zumbro River, 2003}} \\
+        \hline
+            & 402m & 804m \\
+        \hline
+        \multirow{2}{*}{E[Danny]}
+            & 0.02922 & 0.007066 \\
+            & (0.006786, 0.05994) & (0.0000, 0.03403) \\
+        \hline
+        \multirow{2}{*}{E[John]}
+            & 0.03388 & 0.008312 \\
+            & (0.008584, 0.06880) & (0.0000, 0.04188) \\
+        \hline
+        \multirow{2}{*}{E[Tom]}
+            & 0.02228 & 0.012960 \\
+            & (0.002430, 0.05385) & (0.0000, 0.05642) \\
+        \hline
+        \multirow{2}{*}{p[Danny]}
+            & 0.81890 & 0.837100 \\
+            & (0.749500, 0.88540) & (0.7615, 0.90700) \\
+        \hline
+        \multirow{2}{*}{p[John]}
+            & 0.81770 & 0.852500 \\
+            & (0.740100, 0.88670) & (0.7597, 0.92230) \\
+        \hline
+        \multirow{2}{*}{p[Tom]}
+            & 0.73120 & 0.805400 \\
+            & (0.625400, 0.82440) & (0.6885, 0.89780) \\
+        \hline
+        \multirow{2}{*}{\(\psi\)}
+            & 0.66550 & 0.651600 \\
+            & (0.549100, 0.78150) & (0.5098, 0.78210) \\
+        \hline
+        \multirow{2}{*}{\(\theta\)}
+            & 0.29950 & 0.395800 \\
+            & (0.236400, 0.36550) & (0.3100, 0.48520) \\
+        \hline
+    \end{tabular}
+    \end{center}
+
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|}
+        \hline
+        \multicolumn{2}{|l|}{\textbf{Zumbro River, 2004}} \\
+        \hline
+            & 402m \\
+        \hline
+        \multirow{2}{*}{p[Danny]}
+            & 0.9166 \\
+            & (0.7739, 0.9918) \\
+        \hline
+        \multirow{2}{*}{\(\psi\)}
+            & 0.4773 \\
+            & (0.3586, 0.6114) \\
+        \hline
+        \multirow{2}{*}{\(\theta\)}
+            & 0.4232 \\
+            & (0.3073, 0.5520) \\
+        \hline
+    \end{tabular}
+    \end{center}
+    \end{table}
+
+    \subsection{Summary and interpretation of estimates}
+    In Tables \ref{Mississippi}, \ref{Whitewater}, and \ref{Zumbro} lie our
+    estimates for all of the data from
+    2003 and 2004.
     With respect to the model's ability to converge on the actual data, the most
@@
 
-    2003's data worked relatively well with the model. The presence of numerous
+    The 2003 data worked relatively well with the model. The presence of
+    numerous
     observations that occur in the second and third days since the most recent
@@
 
-    2004's data, which suffered from the lack of multiple observers and
+    The 2004 data, which suffered from the lack of multiple observers and
     observations later than one day after each snow event, does not provide
@@
     last snow event) for the parameters \(p\) and \(E\) to converge, regardless
-    of the number of Markov Chain Monte Carlo (MCMC) iterations performed.
-
-    % Wow, this stuff SUCKS.
+    of the number of MCMC iterations performed.
+
     As plot size increases, every estimated parameter tends to increase as well.
@@
     amount of correlation between 1's determines the increase in the proportion
-    of 1's to the number of plots -- as the amount of correlation increases, the
+    of 1's to the number of plots -- as the amount of correlation increases,
+    there will be more strings of 1's which are aggregated together, so the
     increase in the proportion decreases.
@@
     We believe that this increase leads to higher \(\psi\) and \(\theta\)
-    estimates because the potential for multiple otters to lay tracks within the
-    same plot increases. In southeastern Minnesota, most otters tolerate the
+    estimates because the potential for otters to lay tracks within a larger
+    plot increases. There is also an increased potential for multiple otters to
+    reside within a larger plot. In southeastern Minnesota, most otters tolerate
+    the
     presence of other otters in the areas they use most frequently, suggesting
@@
     consistent with the noticeable but modest increase in \(\theta\) across plot
-    % I have no idea how to explain this... :(. sizes for all three rivers.
+    sizes for all three rivers.
     However, estimates of \(\psi\) along the Whitewater and Zumbro Rivers remain
-    constant, despite large increases in \(\theta\).
-
-    % More detail, maybe?
+    constant, despite large increases in \(\theta\).  We have not thought of an
+    explanation for this.
+
     Similarly, the increases in estimates of \(p\) and \(E\) seem to result from
     the general increase of tracks laid per plot as the plot size grows. As plot
-    size grows, the influence of a 1 increases, regardless of whether an otter
-    track or a track from a different animal was observed.
+    size grows and there are fewer plots, the influence of a correct or
+    incorrect detection increases.  If the observer fails to see a track, he may
+    observe another track in the same plot, thus correctly assigning that plot a
+    1 despite failing to observe a track.  With smaller plot sizes, this case
+    could result in one detected plot and one undetected plot, decreasing his
+    accuracy.  A similar scenario is possible with incorrectly identified
+    tracks, which would increase \(E\).
 
@@
     that, as a result, our data violates the assumption of independence between
-    sites. However, the model remains effective with spatially correlated data.
+    plots. However, our simulations, described in Section \ref{spatial
+    simulation}, indicate that the model remains effective with spatially
+    correlated data.
     Since the actual data is relatively sparse as it is, we consider the removal
     of half of it in order to preserve independence to be more harmful to the
-    model than supplying it with correlated data. Therefore, all estimates come
+    model than supplying the model with correlated data. Therefore, all
+    estimates come
     from all of the data that we were given.
 
-    % Mississippi River!
-    \begin{center}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
-        \hline
-            & 402m & 804m & 1608m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.01954 & 0.04808 & 0.06885 \\
-            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.05216 & 0.07771 & 0.13650 \\
-            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.02668 & 0.06140 & 0.10230 \\
-            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.65240 & 0.67340 & 0.74960 \\
-            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.67770 & 0.77580 & 0.82540 \\
-            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.53670 & 0.54220 & 0.65330 \\
-            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.59380 & 0.69370 & 0.75820 \\
-            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.15280 & 0.20730 & 0.27690 \\
-            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    This data converged well at the 402m and 804m plot sizes. At the 1608m plot
-    size, \(E\), \(\psi\), and \(\theta\) did not converge well.
-
-    \begin{center}
-    \begin{tabular}{|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Mississippi River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.3129 \\
-            & (0.2174, 0.8107) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.7613 \\
-            & (0.6609, 0.8833) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.8511 \\
-            & (0.2426, 0.9973) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    This data did not converge. We do not think that these estimates are
-    accurate, but they are included for the sake of completeness.
-
-    % Whitewater River!
-    \begin{center}
-    \begin{tabular}{|l|l|l|}
-        \hline
-        \multicolumn{3}{|l|}{\textbf{Whitewater River, 2003}} \\
-        \hline
-            & 402m & 804m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.05608 & 0.0555 \\
-            & (0.02474, 0.09554) & (0.002859, 0.1390) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.16570 & 0.1931 \\
-            & (0.10130, 0.23510) & (0.065910, 0.3384) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.15350 & 0.2241 \\
-            & (0.09758, 0.21660) & (0.110000, 0.3623) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.66990 & 0.6985 \\
-            & (0.56080, 0.79040) & (0.573100, 0.8366) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.92040 & 0.9877 \\
-            & (0.84270, 0.97590) & (0.945100, 1.0000) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.75910 & 0.8575 \\
-            & (0.65060, 0.85800) & (0.745400, 0.9600) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.87590 & 0.9126 \\
-            & (0.72910, 0.98880) & (0.746900, 0.9955) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.24100 & 0.3779 \\
-            & (0.17890, 0.30900) & (0.286500, 0.4721) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    While the 402m data converged well, \(E\), \(\psi\), and \(\theta\) did not
-    converge well for the 804m data.
-
-    \begin{center}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Whitewater River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.8237 \\
-            & (0.5323, 0.9804) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.6066 \\
-            & (0.4246, 0.8382) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.2068 \\
-            & (0.1211, 0.3550) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    This data converged well. However, we feel that the absence of \(E\) from
-    the model makes the estimates much less dependable. The estimates are also
-    much less precise than those from 2003 -- the confidence intervals of
-    \(\psi\) and of \(p[Danny]\) are twice as large as those from 2003.
-
-    % Zumbro River!
-    \begin{center}
-    \begin{tabular}{|l|l|l|}
-        \hline
-        \multicolumn{3}{|l|}{\textbf{Zumbro River, 2003}} \\
-        \hline
-            & 402m & 804m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.02922 & 0.007066 \\
-            & (0.006786, 0.05994) & (0.0000, 0.03403) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.03388 & 0.008312 \\
-            & (0.008584, 0.06880) & (0.0000, 0.04188) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.02228 & 0.012960 \\
-            & (0.002430, 0.05385) & (0.0000, 0.05642) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.81890 & 0.837100 \\
-            & (0.749500, 0.88540) & (0.7615, 0.90700) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.81770 & 0.852500 \\
-            & (0.740100, 0.88670) & (0.7597, 0.92230) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.73120 & 0.805400 \\
-            & (0.625400, 0.82440) & (0.6885, 0.89780) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.66550 & 0.651600 \\
-            & (0.549100, 0.78150) & (0.5098, 0.78210) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.29950 & 0.395800 \\
-            & (0.236400, 0.36550) & (0.3100, 0.48520) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    Again, the 402m data converged well. With respect to the 804m data, while
-    \(\psi\) and \(\theta\) converged well, none of the other parameters did.
-
-    \begin{center}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{2}{|l|}{\textbf{Zumbro River, 2004}} \\
-        \hline
-            & 402m \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.9166 \\
-            & (0.7739, 0.9918) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.4773 \\
-            & (0.3586, 0.6114) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.4232 \\
-            & (0.3073, 0.5520) \\
-        \hline
-    \end{tabular}
-    \end{center}
-
-    This data converged well. However, as is the case with all of the 2004 data,
-    we feel that the absence of \(E\) from the model makes the estimates much
-    less dependable.
-
-    \subsection{Covariates}
-    To determine the relative importance of our covariates (observer, date, and
+    \subsection{Recommended sampling practices}
+    To determine the relative importance of the different sampling variables
+    (number of observers, number of snow events, and
     number of days since the most recent snow event), we ran the standard
-    hierarchical model with \(E\) on simulated data with values of \(\psi\) and
-    \(\theta\) similar to our estimates of \(\psi\) and \(\theta\) on 2003's
-    data with varying numbers of observers, snow events, and observations after
-    each snow event. We assume that observations after each snow event will
+    hierarchical model with false detection on simulated data with values of
+    \(\psi\) and
+    \(\theta\) similar to the estimates of \(\psi\) and \(\theta\) from 2003
+    data. We varied the numbers of observers, snow events, and observations
+    after
+    each snow event. We assumed that observations after each snow event will
     occur on consecutive days, starting from the first day after the snow event.
@@
     The number of observers proved to be the least important of the three
-    covariates. We were able to obtain accurate estimates regardless of the
+    sampling variables. We were able to obtain accurate estimates regardless of
+    the
     number of observers, as long as there were sufficient numbers of snow events
@@
     The number of snow events proved to be the second-most important of the
-    three covariates. Although two snow events are required, only with one
+    three sampling variables. Although two snow events are required, only with
+    one
     observer does the addition of a third snow event noticeably affect the
@@
     The number of observations after each snow event proved to be the most
-    important of the three covariates. Because the model relies upon the
+    important of the three sampling variables. Because the model relies upon the
     assumption that the probability of seeing a track increases each day because
     tracks made on previous days within the same snow event can still be seen,
-    this is not surprising. Two observations after each snow event on
+    this is not surprising. At least two observations after each snow event on
     consecutive days are required. Additional observations after each event help
@@
 
-    \subsection{Future work}
-    Additional observations are more helpful than additional snow events at
-    first, although both are much more helpful than the addition of an observer.
+    Additional observations after each snow event are more helpful than
+    additional snow events for the first three days of observation.
+    Both are much more helpful than the addition of an observer.
     Once the numbers of snow events and observations after each snow event both
@@
     less challenging than increasing the number of observations after each snow
-    event at that point, especially in southeastern Minnesota.
+    event at that point.
 
@@
 
-%Needs to be fixed/formatted better... I didn't know how to make an appendix
 \appendix

### 2010-03-01 Presentation/Presentation.tex
Added z-score boxplot.
added some slides

@@
 
-\begin{frame}[label=currentslide]{Independence, continued}
+\begin{frame}{Independence, continued}
 
@@
 
-\begin{frame}{Our model}
+\begin{frame}{Bayesian vs. likelihood occupancy models}
+	\begin{itemize}
+		\item Maximum likelihood assumes random sample from large population
+		\begin{itemize}
+			\item Entire section of river was sampled
+		\end{itemize}
+		\item Most occupancy models have two random processes- occupancy and 
+		detection
+		\begin{itemize}
+			\item We added false detection and track-laying
+			\item Extra parameters made likelihood model complicated
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Issue with false detection}
+	\begin{itemize}
+		\item $(p=x, E=y, \psi=z)$ equivalent in terms of likelihood to 
+		$(p=y,E=x,\psi=1-z)$.
+		\item Leads to bimodal posterior distributions for $p$ and $E$.
+		\item Solution: restrict $p$ to interval $(0.6,1)$, $E$ to interval 
+		$(0,0.4)$.
+	\end{itemize}
+\end{frame}
+
+%This slide for DNR only
+\begin{frame}{Bayesian model specifics}
+	\begin{itemize}
+		\item $\psi,\theta \sim \text{Uniform}(0,1)$
+		\item $p_{\text{obs}} \sim \text{Beta}(a,b)$
+		\begin{itemize}
+			\item $a, b \sim \text{Gamma}(1,0.1)$
+			\item E similar
+			\item $p$ and $E$ had restricted distributions
+		\end{itemize}
+		\item Burn-in period 10,000 iterations
+		\item Covariates: observer, days since snow
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{Testing the model- data simulations}
+	\begin{itemize}
+		\item Simulated data with known $\psi$, $\theta$, $p$, $E$.
+		\item To simplify, every observer flew every day, flights ran all 
+		three days after a snow event
+		\item Ran data on our model
+		\item When regularity restriction lifted, similar results
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Results from simulated data}
+	\begin{itemize}
+		\item Lower values for $\psi$ and $\theta$ had better convergence and 
+		estimates
+		\item 95\% confidence interval about 0.2 units wide
+		\item Estimates usually within 0.1 of actual value
+		\item Overall, model estimates parameters well
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{Modeling spatial correlation}
+	\begin{itemize}
+		\item Conditional autoregressive (CAR) models
+		\item Allows extreme values to influence estimates for neighboring sites
+		\item Placed CAR prior on $\theta$, track-laying probability
+		\item With CAR model, each site has a unique estimate of $\theta$
+	\end{itemize}
 \end{frame}
 
-\begin{frame}{Testing the CAR model}
+%DNR only slide
+\begin{frame}{CAR model specifics}
+	\begin{itemize}
+		\item $\tau^2 =$ conditional variance
+		\item $(\tau^2)^{-1} \sim \text{Gamma}(0.5,0.0005)$
+		\item $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$
+		\item $\alpha_0 \sim$ Improper uniform, $\alpha_1 \sim$ Gaussian CAR
+	\end{itemize}
 \end{frame}
 
+%DNR slide
+\begin{frame}[label=currentslide]{Testing the CAR model}
+	\begin{itemize}
+		\item Simulated data with correlation in $\theta$
+		\begin{itemize} 
+			\item $P(\text{track in site }i|\text{no track in site }i-1)=\theta$
+			\item $P(\text{track in site }i|\text{track in site }i-1)=\theta+
+			\alpha$
+		\end{itemize}
+		\item CAR model did not converge
+		\item Simulated data more independent than actual data (from $\hat{S}$)
+		\item Added similar correlation in $\psi$
+		\begin{itemize}
+			\item Data converged more often
+			\item Independence more similar to that of actual data
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+%Carleton slide
 \end{document}

### 2010-03-01 Presentation/Presentation.tex
finished presentation draft

@@
 \usepackage{graphpap,graphics}
-\includeonlyframes{currentslide}
+%\includeonlyframes{currentslide}
 
@@
 		\item Detection rates varied among the observers
-		%include graph here
+	\end{itemize}
+	\begin{center}
+		\resizebox{5cm}{!}{\includegraphics{../observerPlots.pdf}}
+	\end{center}
+	\begin{itemize}
 		\item Solution: use observer indicators as covariates for detection 
@@
 		\item $\psi,\theta \sim \text{Uniform}(0,1)$
-		\item $p_{\text{obs}} \sim \text{Beta}(a,b)$
+		\item $p[{\text{obs}] \sim \text{Beta}(a,b)$
 		\begin{itemize}
@@
 		\item Conditional autoregressive (CAR) models
-		\item Allows extreme values to influence estimates for neighboring sites
+		\item Allows extreme values to influence estimates of neighboring sites
 		\item Placed CAR prior on $\theta$, track-laying probability
@@
 		\item $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$
-		\item $\alpha_0 \sim$ Improper uniform, $\alpha_1 \sim$ Gaussian CAR
+		\item $\alpha_0 \sim$ Improper uniform
+		\item $\alpha_1 \sim$ Gaussian CAR
 	\end{itemize}
@@
 %DNR slide
-\begin{frame}[label=currentslide]{Testing the CAR model}
+\begin{frame}{Testing the CAR model}
 	\begin{itemize}
@@
 		\begin{itemize}
-			\item Data converged more often
+			\item CAR model converged more often
 			\item Independence more similar to that of actual data
@@
 
-%Carleton slide
+\begin{frame}{Testing CAR model, continued}
+	\begin{itemize}
+		\item With correlated data, no constant $\psi$.
+		\item To find true $\psi$ from simulated data, took proportion of plots 
+		occupied
+		\item When CAR model converged, estimates from CAR and standard models 
+		similar for the same data
+	\end{itemize}
+\end{frame}
+
+%Carleton slide goes here
+\section{Discussion}
+\begin{frame}{Obtaining convergence on the data}
+	\begin{itemize}
+		\item All of 2003 data converged with standard model
+		\item Nothing in 2004 converged
+		\begin{itemize}
+			\item Removed false detection, 402m in 2004 converged
+		\end{itemize}
+		\item CAR model never converged with $E$. Without $E$, poor 
+		estimates for $\psi$
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Summary and interpretation of estimates}
+	\begin{itemize}
+		\item Estimates for all parameters increase with plot size
+		\item Possible hypotheses...
+		\item Used data from all plots, not alternating plots
+		\begin{itemize}
+			\item More data more useful than more independence
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Recommended sampling practices}
+	\begin{itemize}
+		\item Investigated effects of changing the sampling variables
+		\item In order of decreasing importance, maximize:
+		\begin{enumerate}
+			\item Number of consecutive days after snow event sampled
+			\item Number of snow events
+			\item Number of observers
+		\end{enumerate}
+		\item All conclusions come from regular data
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Acknowledgments}
+	\begin{itemize}
+		\item Bob Dobrow, adviser
+		\item John Fieberg, MN DNR
+		\item John Erb, MN DNR
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Questions?}
+\end{frame}
+
 \end{document}

### 2010-03-01 Paper.tex
Added Bob's changes to Sarah's sections

@@
     \subsection{What is occupancy rate?}
-    Occupancy has two different definitions. It can mean ``the probability that
+    Occupancy has two different definitions in the literature. It can mean ``the
+	probability that
     a randomly selected site or sampling unit in an area of interest is occupied
@@
     population size, of a species. The number of members of a species is not
-    important for either definition of occupancy.
+    relevant for either definition of occupancy.
 
     While there is a difference between these two definitions of occupancy, in
-    general the probability is unknown and the observed proportion is used as an
+    general the probability of occupancy is unknown and the observed proportion
+	is used as an
     estimate of the underlying probability. Because of this, these two
-    definitions of occupancy are oftentimes used interchangeably.
+    definitions are oftentimes used interchangeably.
 
@@
     As is true for all modeling, there are some assumptions made in most models
-    of occupancy
-    in order to create simple enough models \cite{MacKenzie2006}. One common
+    of occupancy \cite{MacKenzie2006}. One common
     assumption is no false positives, which means if a
     plot is recorded as occupied we say it must actually be occupied. A second
-    common assumption is that presence and detection probabilities are constant
-    across plots and surveys. This simplifies the estimation of each parameter.
-    A third common assumption is that there is independence of detection between
-    % Question: Independence of occupancy as an assumption as well?
-    plots. A fourth common assumption is that the surveys are being conducted on
+    assumption is that presence and detection probabilities are constant
+    across plots and surveys.
+    Third, we assume that there is independence of detection between
+    plots. And finally, we assume that the surveys are being conducted on
     a closed population. This means that no animals leave or join the
@@
     talk about attempting to eliminate the
-    assumption of no false positives and the assumption of independence of
-    parameters across plots.
+    assumption of no false positives and the independence assumption across
+	plots.
 
@@
 
-    The most simple way to measure occupancy is to divide the number of
+    The simplest way to measure occupancy is to divide the number of
     plots where an animal was found to be present by the total number of plots.
@@
 
-    Figure \ref{simplehierarchicalmodel} shows that we need separate parameters
-    for the
-    probability that a plot is occupied, the probability that a track is laid
-    down given that a plot is occupied, and the probability of detection
-    given that a track is laid down and the plot is occupied. We will call
-    these \(\psi\), \(\theta\), and \(p\), respectively.
+    Figure \ref{simplehierarchicalmodel} suggests that we need separate 
+	parameters for (i) the probability that a plot is occupied, (ii) the 
+	probability that a track is laid
+    down given that a plot is occupied, and (iii) the probability of detection
+    given that a track is laid down and the plot is occupied. We notate
+    these as \(\psi\), \(\theta\), and \(p\), respectively.
 
@@
     independence between plots, we wanted to see if this assumption held true
-    for our data.  In order to do this, we calculated the statistic \(S\), which
-    we define as the number of \(0\)--\(1\)'s minus the number of
-    \(1\)--\(1\)'s per flight. Under the null hypothesis that each
+    for our data.  In order to do this, we proposed the statistic \(S\), which
+    we define as the number of successive \(0\)--\(1\)'s minus the number of
+    successive \(1\)--\(1\)'s per flight. Under the null hypothesis that each
     plot is an independent Bernoulli trial where \(p\) is the probability of a
@@
     to compute the mean and standard deviation of \(S\), which we used to
-    calculate a Z-score, where \(Z = (\hat{S} - E[S])/SD[S]\). In order to
+    calculate a Z-score, where \(Z = (S - E[S])/SD[S]\). In order to
     compute these values, we needed to find the
-    expected value and variance of each term in \(S\). These calculations can be
+    expected value and variance of \(S\). These calculations can be
     found in Appendix \ref{math}.
 
-    Simulations of \(S\) show that the distribution of \(S\) is roughly normal.
+    Simulations show that the distribution of \(S\) is roughly normal.
     Using our calculations for the mean and the standard deviation, we
@@
     the assumption of
-    independence between plots is violated for that flight. We found that, in
+    independence between plots is violated for that flight, and we found that a
+	large number of flights show a significant departure from independence. We
+	found that, in
     general, data with alternating plots are more ``independent'' than all plots
@@
     will
-    discuss later whether this violation of independence matters in our
-    model, and how this might affect the estimation of actual occupancy rates.
+    discuss later how sensitive our model is to this violation of independence,
+	and how this might affect the estimation of actual occupancy rates.
 
@@
     equivalent in terms of likelihood to \((p=y,E=x,\psi=1-z)\) due to the
-    symmetry of the model \cite{Royle2006}.  Since we know that is reasonable
+    symmetry of the model \cite{Royle2006}.  Since it is reasonable
     for \(p\) to be greater than \(E\), we solved this problem by restricting
@@
 \appendix
-\section{Deriving \(E[S]\) and \(SD[S]\)}
+\section{Deriving the mean and standard deviation of our \(S\) statistic}
 \label{math}
-    We let \(A = \text{\# of
-    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).
-    In order to find \(E[s]\), we first needed to find \(E[A]\) and
+    Given a binary sequence \(X_1, ..., X_n\) of independently and indentically
+	distributed \(0\)'s and \(1\)'s, let \(A = \text{\# of successive 0-1's}\),
+	\(B=\text{\# of successive 
+	1-1's}\) and \(p = \text{ the proportion of 1's in the sequence}\).
+    In order to find \(E[S]\), we first needed to find \(E[A]\) and
     \(E[B]\).
@@
 
-    % E[s]
+    % E[S]
     \begin{equation*}
         \begin{split}
-        E[s]& =E[A]-E[B] \\
+        E[S]& =E[A]-E[B] \\
                   & =(n-1)\left((1-p)p-p^2\right)
@@
 
-    The standard deviation was a little more difficult.
+	For the standard deviation,
 
@@
         \begin{split}
-        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[I_i=(0,1),I_{i+1}=
-                 (1,1)] \\
-              & =(n-1)P((0,1))(1-P((0,1))) \\
-              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(0,1),I_{i+1}=(0,1)]-E[I_i=
-                (0,1)]E[I_{i+1}=(0,1)]\bigg) \\
+        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[\{X_i=0, 
+				 X_{i+1}=1\}, \{X_{i+1}=0, X_{i+2}=1\}] \\
+              & =(n-1)p(1-(1-p)p) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[\{X_i=0, X_{i+1}=1\},
+			  	\{X_{i+1}=0, X_{i+2}=1\}]  \\
+			  &	\qquad \qquad \quad - E[X_i=0, X_{i+1}=1]E[ X_{i+1}=0, 
+				 X_{i+2}=1]\bigg) \\
               & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
@@
         \begin{split}
-        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[I_i=(1,1),I_{i+1}=
-                (1,1)] \\
+        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[\{X_i=1, 
+				 X_{i+1}=1\}, \{X_{i+1}=1, X_{i+2}=1\}] \\
               & =(n-1)P((1, 1))(1-P((1, 1))) \\
-              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,1),I_{i+1}=(1,1)]-E[I_i=
-                (1,1)]E[I_{i+1}=(1,1)]\bigg) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[\{X_i=1, X_{i+1}=1\},
+				 \{X_{i+1}=1, X_{i+2}=1\}]  \\
+			  & \qquad \qquad \quad -E[X_i=1, X_{i+1}=1]E[X_{i+1}=1,
+				 X_{i+2}=1]\bigg) \\
               & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
@@
 
-    % Var[s]
+    % Var[S]
     \begin{equation*}
         \begin{split}
-        Var[s]& =Var[A-B] \\
+        Var[S]& =Var[A-B] \\
                     & =Var[A]+Var[B]-2Cov[A,B] \\

### 2010-03-01 Paper.tex
Edits to standard model and discussion sections.

--- 2010-03-01 Paper.tex
@@
     The data we received from the
-    Minnesota Department of Natural Resources was not fixed with respect to
-    our covariates and required us to implement one of the three common
-    strategies used to deal with irregular data sets \cite{Spiegelhalter2003}.
-    Our simulations, on the other
+    Minnesota Department of Natural Resources was unbalanced with regard to
+    covariates.
+    Our initial simulations, on the other
     hand, ran on data sets with a fixed number of observers, snow events, and
@@
     range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
-    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
+    burn-in time for the MCMC simulation to 10,000 iterations, which seemed
+    necessary to handle convergence. Each of these
     parameters are tracked for 10,000 additional iterations and are estimated
@@
     contain the same number of flights by the same observers after each snow
-    event.  Naturally, unbalanced data sets do not.
+    event.
 
@@
     receives the same number of flights by the same number of observers, which
-    greatly simplifies the accompanying hierarchical model. First, the
-    simulation determines a plot's occupancy based upon the given \(\psi\)
-    value.
+    greatly simplifies the accompanying hierarchical model. For the simulation,
+    we first determine each plot's occupancy status based
+    based upon the given \(\psi\) value.
     Then, for each day after each snow event, the simulation decides whether or
@@
     Because of limitations with respect to WinBUGS' syntax, simulating
-    unbalanced data requires much more additional information and a reformatting
+    unbalanced data requires additional information and a reformatting
     of the data as well as the corresponding hierarchical model. While the
-    resulting code's structure differs significantly from that used for balanced
+    resulting code's structure differs from that used for balanced
     data, the underlying model remains the same.
@@
     prior distributions on \(p\) and \(E\) performed better than the model with
-    normal prior distributions on the same parameters.  Therefore, we limit
-    further discussion to the beta model.
+    normal prior distributions on the same parameters. The beta distribution
+    gives a flexible two-parameter family of distributions on \((0,1)\). We
+    included two hyper-parameters (\(\alpha\) and \(\beta\)) on the beta
+    distribution as part of our hierarchical modeling, which seemed to work
+    well.
 
     \begin{table}
-    \caption{Beta model estimates from simulated data. All simulations
+    \caption{Model estimates from simulated data. All simulations
     ran with 70 plots, 5 snow events, 3 observers, and for 3 days after each
@@
     of the
-    model's 25 confidence intervals for \(\psi\) include its actual value.
+    model's 25 confidence intervals for \(\psi\) capture the actual value.
     The model's confidence intervals for each of the
@@
     distribution on \(\theta\) to the standard model exhibited convergence on
-    our simulated data, we moved onto running both models on the data.
+    our simulated data, we moved on to running both models on the actual data.
     In general, the sparseness of the data gathered in 2004 caused us to
@@
     gathered in 2003, this result was not surprising. The vast majority of
-    2004's data consists of just one observer taking observations once after
+    2004 data consists of just one observer taking observations once after
     each snow event, which causes \(p\) and \(E\) in particular to not converge,
@@
 
-    To obtain convergence on 2004's relatively sparse data, we removed \(E\),
-    which adds to the model the assumption that there are no false positives
-    and, consequently, simplifies it greatly. Although the model still did not
+    To obtain convergence on 2004 data, we removed \(E\), thus disallowing false
+    positives. Although the model still did not
     converge on data with plot sizes of 804m and 1608m, both the Whitewater and
@@
     part, however, we feel that the absence of \(E\) from the model makes the
-    the estimates of \(\psi\) much less dependable.
+    the estimates of \(\psi\) less dependable.
 
@@
     than did the CAR model. This is most likely due to the fact that the CAR
-    model attempts to estimate too many parameters. Whereas the standard model
+    model attempts to estimate many more parameters. Whereas the standard model
     estimates between eight and twelve parameters on the actual data (\(\psi\),
@@
     \subsection{Summary and interpretation of estimates}
-    In Tables \ref{Mississippi}, \ref{Whitewater}, and \ref{Zumbro} lie our
+    Tables \ref{Mississippi}, \ref{Whitewater}, and \ref{Zumbro} contain our
     estimates for all of the data from
@@
     occur on consecutive days, starting from the first day after the snow event.
-    This information helped us to find the most efficient ways (with respect to
+    This information helped us to find more efficient ways (with respect to
     number of flights) to gather aerial occupancy data given one, two, and three
@@
 
-    The number of observers proved to be the least important of the three
-    sampling variables. We were able to obtain accurate estimates regardless of
+    \textbf{The number of observers appears to be the least important of the
+    three
+    sampling variables.} We were able to obtain accurate estimates regardless of
     the
@@
 
-    The number of snow events proved to be the second-most important of the
-    three sampling variables. Although two snow events are required, only with
+    \textbf{The number of snow events appears to be the second-most important of
+    the
+    three sampling variables.} Although two snow events are required, only with
     one
@@
 
-    The number of observations after each snow event proved to be the most
-    important of the three sampling variables. Because the model relies upon the
+    \textbf{The number of observations after each snow event appears to be the
+    most
+    important of the three sampling variables.} Because the model relies upon
+    the
     assumption that the probability of seeing a track increases each day because
@@
 
-    In summary, for future data gathering trips, we strongly suggest using one
+    In summary, for future data gathering trips, we suggest using one
     observer, as long as they are relatively confident that they can correctly

### 2010-03-01 Paper.tex
Kiva's edits.

--- 2010-03-01 Paper.tex
@@
 \usepackage{hyperref}
+\usepackage{moreverb}
 
@@
 Population in Southeastern Minnesota \\
-{\large Senior Comps Project}}
+{\large Senior Comprehensive Exercise: Carleton College}}
 \author{Chrisna Aing, Sarah Halls, Kiva Oken}
@@
 time. We used concepts from occupancy theory to develop models to estimate
-the occupancy rate in order to get a sense of otter population.
+the occupancy rate for the otter population.
 
     \subsection{Data collection}
-    During the winters of 2003 and 2004 people flew over the Whitewater, Zumbro,
-    and Mississippi Rivers. When they saw what looked like an otter track, they
-    pushed a button recording the GPS coordinates. They pushed the button every
+    During the winters of 2003 and 2004 The Minnesota Department of Natural 
+    Resources (MDNR) sponsored flights over the Whitewater, Zumbro,
+    and Mississippi Rivers to survey otters. When observers saw what looked like 
+    an otter track, they
+    pushed a button recording on a handheld GPS unit to record the GPS 
+    coordinates (waypoint). They pushed the button every
     five seconds until the track ended. Flights occurred one, two, or three days
@@
     flight was determined. The Mississippi River was also divided into 1608
-    meter plots.  For each river and plot size, the data set with every other
-    plot removed was considered.  We refer to this as ``alternating plots''.
+    meter plots.  In addition, for each river and plot size, a data set with 
+    every other
+    plot removed was considered.  We refer to these as ``alternating plots''.
 
@@
     track in one plot is not independent of the probability that an otter lays
-    down a track in neighboring plots, violating a principle assumption of
-    occupancy theory. That is, a single track can extend across multiple plots.
-    This leads to dependence among the plots. It was important to quantify this
-    dependence and find an optimum way to handle it.
+    down a track in neighboring plots. That is, a single track can extend across 
+    multiple plots. This leads to dependence among the plots, violating a 
+    principle assumption of many
+    occupancy models. We tried to quantify this
+    dependence and find a way to handle it.
 
@@
     the number of flights increases, the occupancy rate estimate will approach
-    1 \cite{Royle2006}. Unlike most occupancy models where observers assume
-    that detection of the species in question guarantees its presence \cite
-    {MacKenzie2006}, we introduced a false detection process whereby an observer
+    1 \cite{Royle2006}. We thus introduced a false detection process whereby an 
+    observer
     can mark down the existence of a track when in reality there is none. This
@@
 \section{Spatial correlation of the data}
-One approach to dealing with the correlation in the probability of
-observing a track in neighboring plots is to use a conditional autoregressive
+One approach to dealing with the correlation in the
+observed tracks in neighboring plots is to use a conditional autoregressive
 (CAR) prior distribution for $\theta$, the probability that an otter lays down a
@@
     iterations. With this model, instead of a single estimate for $\theta$ there
-    are now $S$ estimates for $\theta$ where $S$ is the total number of plots.
+    are now $n$ estimates for $\theta$ where $n$ is the total number of plots.
     While ideally we would estimate an overall value for $\theta$, it seems more
@@
     \label{spatial simulation}
-    Because we modeled spatial correlation in $\theta$, we began by simulating
+    We began by simulating
     spatial correlation in the track-laying process. To do this, we added an
-    additional variable to the simulated data, $\alpha$. Under the new
+    additional variable, $\alpha$, to the simulated data. Under the new
     simulation method, $\theta$ is the probability of an otter laying down a
@@
     model}, the model almost never converged. We tested independence of the data
-    using the $\hat{S}$ statistic we developed (Section \ref{shat}) and found
-    that the simulated data displayed much more independence than the actual
-    data. We would like our simulated data to have the same level of
-    independence
+    using the $S$ statistic we developed (Section \ref{shat}) and found
+    that the simulated data displayed much less dependence than the actual
+    data. We would like our simulated data to have the same level of dependence
     as our actual data.  We tried modeling spatial correlation in the occupancy
@@
 
-    To increase the correlation between plots we simulated data
-    correlated in both the track-laying and occupancy processes. Under this 
+    To increase the correlation between plots we simulated the correlation
+    in both the track-laying and occupancy processes. Under this 
     simulation method we
@@
     $i-1$ is occupied. With this increased spatial correlation we were able to
-    more closely match the $\hat{S}$ Z-scores found in the actual data. In
+    more closely match the $S$-related Z-scores found in the actual data. In
     addition, the CAR model converged more often with data simulated using this
@@
 
+\section{Standard hierarchical model}
+\scriptsize
+\begin{verbatimtab}
+otterModel <- function() {
+    # Initialization of variables to estimate.
+    psi ~ dunif(0,1)
+    theta ~ dunif(0,1)
+    a ~ dgamma(1,.1)
+    b ~ dgamma(1,.1)
+    c ~ dgamma(1,.1)
+    d ~ dgamma(1,.1)
+    for (observer in 1:observers) {
+        #p[observer] ~ dbeta(a,b)
+        p[observer] ~ dbeta(a,b) %_% I(.5, 1)
+        E[observer] ~ dbeta(c,d) %_% I(0, .5)
+    }
+
+    # Execution of all site- and track-related work.
+    for (site in 1:sites) {
+        z[site] ~ dbern(psi)
+        lambda[site] <- theta * z[site]
+        for (event in 1:snowEvents) {
+            for (day in 1:daysAfterTotal) {
+                track[site, event, day] ~ dbern(lambda[site])
+                trackSum[site, event, day] <- sum(track[site, event, 1:day])
+            }
+        }
+    }
+
+    # Super difficult stuff.  (Not really.)
+    for (i in 1:numDataPts) {
+        # equals(track.count[...], 0) == 1 when there is no track to observe.
+        mu[i] <- 
+            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
+        # probability of false detection, given track and occupancy state
+        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
+        prob[i] <- mu[i] + nu[i]
+        #prob[i] <- mu[i]
+        y[i] ~ dbern(prob[i])
+    }
+}
+
+writeModel(otterModel, "irregularHierarchicalModel.txt")
+modelCheck("irregularHierarchicalModel.txt")
+bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
+           "numDataPts", "snowEvents", "daysAfterTotal"), 
+         "irregularOtterData.txt")
+modelData("irregularOtterData.txt")
+modelCompile(numChains=2)
+modelGenInits()
+
+modelUpdate(10000)
+samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi", "theta"))
+#samplesSet(c("a", "b", "p[]", "psi", "theta"))
+modelUpdate(10000)
+\end{verbatimtab}
+%}
+
+\section{CAR model}
+\scriptsize
+\begin{verbatimtab}
+otterModel <- function() {
+    for (site in 1:sites) {
+        logit(theta[site]) <- alpha1[site] + alpha0
+    }
+    for (i in 1:sumNumNeigh) {
+        weights[i] <- 1
+    }
+    alpha1[1:sites] ~ car.normal(adj[], weights[], num[], tau)
+    alpha0 ~ dflat()
+    tau ~ dgamma(.5,.0005)
+    psi ~ dunif(0,1) 
+    a ~ dgamma(1,.1)
+    b ~ dgamma(1,.1)
+    #c ~ dgamma(1,.1)
+    #d ~ dgamma(1,.1)
+    for (observer in 1:observers) {
+        p[observer] ~ dbeta (a,b)
+	  #p[observer] ~ dbeta(a,b) %_% I(.6, 1)
+        #E[observer] ~ dbeta(c,d) %_% I(0, .3)
+    }
+
+    # Execution of all site- and track-related work.
+    for (site in 1:sites) {
+        z[site] ~ dbern(psi)
+        lambda[site] <- theta[site] * z[site]
+        for (event in 1:snowEvents) {
+            for (day in 1:daysAfterTotal) {
+                track[site, event, day] ~ dbern(lambda[site])
+                trackSum[site, event, day] <- sum(track[site, event, 1:day])
+            }
+        }
+    }
+
+    # Super difficult stuff.  (Not really.)
+    for (i in 1:numDataPts) {
+        # equals(track.count[...], 0) == 1 when there is no track to observe.
+        mu[i] <- 
+            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
+        # probability of false detection, given track and occupancy state
+        #nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
+        #prob[i] <- mu[i] + nu[i]
+        prob[i] <- mu[i]
+	  y[i] ~ dbern(prob[i])
+    }
+}
+
+writeModel(otterModel, "irregularHierarchicalModel.txt")
+modelCheck("irregularHierarchicalModel.txt")
+bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
+           "numDataPts", "snowEvents", "daysAfterTotal", "num", "adj",
+           "sumNumNeigh"), 
+         "irregularOtterData.txt")
+modelData("irregularOtterData.txt")
+modelCompile(numChains=2)
+
+init.val <- function(){
+	list(
+		alpha1=runif(sites,-100,100),
+		alpha0=runif(1,-100,100),
+		tau=rgamma(1,.5,.0005),
+		psi=runif(1),
+		a=rgamma(1,1,.1),
+		b=rgamma(1,1,.1)
+		#,c=rgamma(1,1,.1),
+		#d=rgamma(1,1,.1)
+	) 
+}
+bugsInits(init.val,numChains=2,fileName=c('init1.txt','init2.txt'))
+modelInits(c('init1.txt','init2.txt'))
+
+modelGenInits()
+
+modelUpdate(12000)
+#samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi"))
+samplesSet(c("a", "b", "p[]", "psi"))
+modelUpdate(10000)
+\end{verbatimtab}
+
+\section{Simulated data, independent plots}
+\scriptsize
+\begin{verbatimtab}
+# Setting up the environment and results matrix.
+sites <- 100
+snowEvents <- 2
+daysAfterTotal <- 2
+observers <- 3
+
+eventsToDaysBegin <- c(1, 3, 5, 7, 9, 11)
+eventsToDaysEnd   <- c(2, 4, 6, 8, 10, 12)
+curDaySinceSnow <- c(1, 2, 1, 2, 1, 2, 1, 2, 1, 2, 1, 2)
+daysToObsBegin <- c(1, 4, 7, 10, 13, 16, 19, 22, 25, 28, 31, 34)
+daysToObsEnd   <- c(3, 6, 9, 12, 15, 18, 21, 24, 27, 30, 33, 36)
+curObserver <- c(1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3,
+                 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3)
+
+# Setting up the parameters.
+psi <- .7
+beta <- .4
+condlOcc <- psi + beta
+condlUnocc <- psi
+
+theta <- .5
+alpha <- .3
+condlTrack <- theta + alpha
+condlNoTrack <- theta
+p <- c(.70, .80, .90)
+E <- c(.05, .02, .03)
+
+# Retaining necessary information.
+occupancy <- NULL
+
+# Vectors; employing a nested index strategy to deal with WinBUGS and its lack
+# of if-then statements.
+y <- NULL
+siteV <- NULL
+eventV <- NULL
+dayV <- NULL
+obsV <- NULL
+numDataPts <- 0
+
+# Simulating the data!
+for (site in 1:sites) {
+    # TRUE == 1, FALSE == 0.
+	if (site == 1) {
+		occupancy[site] <- 1 * (runif(1) < (condlOcc + condlUnocc/2))
+	} else if (occupancy[i-1] == 1) {
+		occupancy[site] <- 1 * (runif(1) < condlOcc)
+	} else
+		occupancy[site] <- 1 * (runif(1) < condlUnocc)
+	}
+    for (event in 1:snowEvents) {
+        otterTrackLaid <- 0
+        currentDay <- 0
+        # Haha, this is awful.
+        for (day in 1:curDaySinceSnow[eventsToDaysBegin[event]]) {
+            if (day != 1) {
+                currentDay <- currentDay + 1
+                otterTrackLaid <- max(otterTrackLaid,
+                    occupancy[site] * (runif(1) < theta))
+            }
+        }
+        for (day in eventsToDaysBegin[event]:eventsToDaysEnd[event]) {
+            currentDay <- currentDay + 1
+            otterTrackLaid <- max(otterTrackLaid,
+                occupancy[site] * (runif(1) < theta))
+            for (colIndex in daysToObsBegin[day]:daysToObsEnd[day]) {
+                numDataPts <- numDataPts + 1
+                sawOtterTrack <- otterTrackLaid * 
+                    (runif(1) < p[curObserver[colIndex]])
+                sawOtherTrack <- (1 - otterTrackLaid) * 
+                    (runif(1) < E[curObserver[colIndex]])
+                y[numDataPts] <- sawOtterTrack + sawOtherTrack
+                siteV[numDataPts] <- site
+                eventV[numDataPts] <- event
+                dayV[numDataPts] <- currentDay
+                obsV[numDataPts] <- curObserver[colIndex]
+            }
+        }
+    }
+}
+\end{verbatimtab}
+
+\section{Simulated regular data, dependent in $\psi$, $\theta$}
+\scriptsize
+\begin{verbatimtab}
+#Define structure
+sites <- 60
+events <- 5
+snowdays <- 3
+obs <- 3
+I1 <- rep(c(1,0,0), events*snowdays)
+I2 <- rep(c(0,1,0), events*snowdays)
+
+#Define parameters
+psi <- .4
+beta <- .4
+condl.occ <- psi + beta
+condl.unocc <- psi
+theta <-.3
+alpha <-.5
+condl.track <- theta + alpha
+condl.notrack <- theta
+p <- c(.7,.7,.7)
+E <- c(.05,.05,.05)
+
+#Initialize variables
+occupancy <- NULL
+tracks <- matrix(nrow=sites, ncol=events*snowdays)
+y <- matrix(nrow=sites, ncol=events*snowdays*obs)
+
+for(i in 1:sites) {
+	#Determine occupancy of site. 1*(runif(1)<psi) is 1 around psi*100 percent of the time.
+	#Use conditional rates depending on occupancy of site i-1
+	if (i==1) 
+		{occupancy[i] <- 1*(runif(1)<(condl.occ+condl.unocc)/2)}	
+	else if (occupancy[i-1]==1) 
+		{occupancy[i] <- 1*(runif(1)<condl.occ)}
+	else 
+		{occupancy[i] <- 1*(runif(1)<condl.unocc)}
+	for(j in 1:events) {
+		is.track <- 0
+		for(k in 1:snowdays) {
+			index <- k + snowdays*(j-1)
+			#Decide if a track is laid down on day k of event j, given the occupancy state
+			if(i==1) {
+				tracks[i,index] <- (runif(1)<(condl.track+condl.notrack)/2)*occupancy[i]
+			}else if(tracks[i-1,index]==1) 
+				{tracks[i,index] <- (runif(1)<condl.track)*occupancy[i]
+			}else 
+				{tracks[i,index] <- (runif(1)<condl.notrack)*occupancy[i]}
+
+			#Determine if any tracks have been laid down for event j.
+			is.track <- max(is.track, tracks[i,index])
+			for(m in 1:obs) {
+				index <- m + obs*(k-1) + snowdays*obs*(j-1)
+				#Determine if the observer saw a track.
+				y[i,index] <- (runif(1)<p[m])*is.track + (runif(1)<E[m])*(1-is.track)
+			}
+		}
+	}
+}
+\end{verbatimtab}
+Note: This needs some more processing before it will run on either model.
 \end{document}

### 2010-03-02 Paper.tex
Readded Sarah's edits, modified the code appendix slightly.

@@
     \subsection{What is occupancy rate?}
-    Occupancy has two different definitions. It can mean ``the probability that
+    Occupancy has two different definitions in the literature. It can mean ``the
+    probability that
     a randomly selected site or sampling unit in an area of interest is occupied
@@
     population size, of a species. The number of members of a species is not
-    important for either definition of occupancy.
+    relevant for either definition of occupancy.
 
     While there is a difference between these two definitions of occupancy, in
-    general the probability is unknown and the observed proportion is used as an
+    general the probability of occupancy is unknown and the observed proportion
+    is used as an
     estimate of the underlying probability. Because of this, these two
-    definitions of occupancy are oftentimes used interchangeably.
+    definitions are oftentimes used interchangeably.
 
@@
     talk about attempting to eliminate the
-    assumption of no false positives and the assumption of independence of
-    parameters across plots.
+    assumption of no false positives and the independence assumption
+    across plots.
 
@@
 
-    Figure \ref{simplehierarchicalmodel} shows that we need separate parameters
-    for the
-    probability that a plot is occupied, the probability that a track is laid
-    down given that a plot is occupied, and the probability of detection
-    given that a track is laid down and the plot is occupied. We will call
-    these \(\psi\), \(\theta\), and \(p\), respectively.
+    Figure \ref{simplehierarchicalmodel} suggests that we need separate
+    parameters for (i) the probability that a plot is occupied, (ii) the
+    probability that a track is laid
+    down given that a plot is occupied, and (iii) the probability of detection
+    given that a track is laid down and the plot is occupied. We notate
+    these as \(\psi\), \(\theta\), and \(p\), respectively.
 
@@
     the assumption of
-    independence between plots is violated for that flight. We found that, in
+    independence between plots is violated for that flight, and we found that a
+    large number of flights show a significant departure from independence. We
+    found that, in
     general, data with alternating plots are more ``independent'' than all plots
@@
     will
-    discuss later whether this violation of independence matters in our
-    model, and how this might affect the estimation of actual occupancy rates.
+    discuss later how sensitive our model is to this violation of independence,
+    and how this might affect the estimation of actual occupancy rates.
 
@@
 \appendix
-\section{Deriving \(E[S]\) and \(SD[S]\)}
+\section{Deriving the mean and standard deviation of our \(S\) statistic}
 \label{math}
-    We let \(A = \text{\# of
-    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).
-    In order to find \(E[s]\), we first needed to find \(E[A]\) and
+    Given a binary sequence \(X_1, ..., X_n\) of independently and identically
+    distributed 0's and 1's, let \(A=\text{\# of successive 0-1's}\), \(B=
+    \text{\# of successive 1-1's}\) and \(p=\text{the proportion of 1's in the
+    sequence}\). In order to find \(E[S]\), we first needed to find \(E[A]\) and
     \(E[B]\).
@@
 
-    % E[s]
+    % E[S]
     \begin{equation*}
         \begin{split}
-        E[s]& =E[A]-E[B] \\
+        E[S]& =E[A]-E[B] \\
             & =(n-1)\left((1-p)p-p^2\right)
@@
 
-    The standard deviation was a little more difficult.
+    For the standard deviation,
 
@@
         \begin{split}
-        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[I_i=(0,1),I_{i+1}=
-                 (1,1)] \\
-              & =(n-1)P((0,1))(1-P((0,1))) \\
-              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(0,1),I_{i+1}=(0,1)]-E[I_i=
-                (0,1)]E[I_{i+1}=(0,1)]\bigg) \\
+        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[\{X_i=0, 
+                X_{i+1}=1\}, \{X_{i+1}=0, X_{i+2}=1\}] \\
+              & =(n-1)p(1-(1-p)p) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[\{X_i=0, X_{i+1}=1\},
+                \{X_{i+1}=0, X_{i+2}=1\}] \\
+              & \qquad\qquad\quad -E[X_i=0,X_{i+1}=1]E[X_{i+1}=0,X_{i+2}=1]
+                \bigg) \\
               & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
@@
         \begin{split}
-        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[I_i=(1,1),I_{i+1}=
-                (1,1)] \\
-              & =(n-1)P((1, 1))(1-P((1, 1))) \\
-              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,1),I_{i+1}=(1,1)]-E[I_i=
-                (1,1)]E[I_{i+1}=(1,1)]\bigg) \\
+        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[\{X_i=1, 
+                X_{i+1}=1\}, \{X_{i+1}=1, X_{i+2}=1\}] \\
+              & =(n-1)p^2(1-p^2) \\
+              & \quad +2\sum_{i=1}^{n-2}\bigg(E[\{X_i=1, X_{i+1}=1\},
+                \{X_{i+1}=1, X_{i+2}=1\}] \\
+              & \qquad\qquad\quad -E[X_i=1, X_{i+1}=1]E[X_{i+1}=1,X_{i+2}=1]
+                \bigg) \\
               & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
@@
 
-    % Var[s]
+    % Var[S]
     \begin{equation*}
         \begin{split}
-        Var[s]& =Var[A-B] \\
+        Var[S]& =Var[A-B] \\
               & =Var[A]+Var[B]-2Cov[A,B] \\
@@
     for (observer in 1:observers) {
-        #p[observer] ~ dbeta(a,b)
         p[observer] ~ dbeta(a,b) %_% I(.5, 1)
@@
 
-    # Super difficult stuff.  (Not really.)
     for (i in 1:numDataPts) {
@@
         prob[i] <- mu[i] + nu[i]
-        #prob[i] <- mu[i]
         y[i] ~ dbern(prob[i])
@@
 \end{verbatimtab}
-%}
 
@@
     for (observer in 1:observers) {
-        p[observer] ~ dbeta (a,b)
-	  #p[observer] ~ dbeta(a,b) %_% I(.6, 1)
-        #E[observer] ~ dbeta(c,d) %_% I(0, .3)
+        p[observer] ~ dbeta(a,b) %_% I(.6, 1)
+        E[observer] ~ dbeta(c,d) %_% I(0, .3)
     }
@@
         # probability of false detection, given track and occupancy state
-        #nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
-        #prob[i] <- mu[i] + nu[i]
-        prob[i] <- mu[i]
+        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
+        prob[i] <- mu[i] + nu[i]
         y[i] ~ dbern(prob[i])

### 2010-03-02 Presentation/Presentation.tex
fixed a mistake
Being anal, removing the duplicate model diagram.

@@
 \usetheme{}
-\title{Using occupancy theory and Bayesian models to estimate otter populations in southeastern Minnesota}
+\title{Using occupancy theory and Bayesian models to estimate otter populations
+in southeastern Minnesota}
 \author[Aing, Halls, Oken]{Chrisna Aing, Sarah Halls, Kiva Oken}
@@
 {
-	\usebackgroundtemplate{\includegraphics[width=\paperwidth]{ottertracks1.jpg}}
-
+	\usebackgroundtemplate{
+	\includegraphics[width=\paperwidth]{ottertracks1.jpg}}
 	\begin{frame}
@@
 }
-\begin{frame}
+
+\begin{frame}{Outline}
 	\tableofcontents[pausesections]
@@
 {
-	\usebackgroundtemplate{\includegraphics[width=\paperwidth]{SnowTracks12.jpg}}
+	\usebackgroundtemplate{
+	\includegraphics[width=\paperwidth]{SnowTracks12.jpg}}
 	\begin{frame}{}
@@
 {
-	\usebackgroundtemplate{\includegraphics[width=\paperwidth]{SNOWTRACKS9.jpg}}
+	\usebackgroundtemplate{
+	\includegraphics[width=\paperwidth]{SNOWTRACKS9.jpg}}
 	\begin{frame}
@@
 		\item Tracks near plot boundary can be assigned to different plots
-		\item Track in one site means more likely to be track in neighboring site- not independent
+		\item Track in one site means more likely to be track in neighboring
+		site -- not independent
 	\end{itemize}
@@
 		\item Independence of detection between sites
-		\item Survey is conducted on a closed population	\end{itemize}
+		\item Survey is conducted on a closed population
+	\end{itemize}
 \end{frame}
@@
 	\begin{center}
-		\resizebox{7cm}{!}{\includegraphics{SimpleHierarchicalModel.pdf}}
+		\resizebox{7cm}{!}{\includegraphics{../SimpleHierarchicalModel.pdf}}
 	\end{center}
@@
 		\end{itemize}
-		\item Most occupancy models have two random processes- occupancy and 
+		\item Most occupancy models have two random processes -- occupancy and
 		detection
@@
 		\item $\psi,\theta \sim \text{Uniform}(0,1)$
-		\item $p[{\text{obs}] \sim \text{Beta}(a,b)$
+		\item $p[\text{obs}] \sim \text{Beta}(a,b)$
 		\begin{itemize}
@@
 
-\begin{frame}{Testing the model- data simulations}
+\begin{frame}{Testing the model -- data simulations}
 	\begin{itemize}

### 2010-03-03 Presentation/Presentation.tex
Edited presentation
Final presentation for DNR
More presentation work.

@@
 \institute{Carleton College}
-\usepackage{graphpap,graphics}
+\usepackage{graphics}
+\usepackage{color}
+\usepackage{multirow}
 %\includeonlyframes{currentslide}
@@
 		\column{5cm}
-		\resizebox{5cm}{!}{\includegraphics{cuteOtter1.pdf}}
+		\resizebox{5cm}{!}{\includegraphics{cuteOtter2.jpg}}
 		\column{5cm}
@@
 			\item Marked GPS coordinates of tracks
-			\item Rivers divided into 400, 800, or 1600 m plots
+			\item Rivers divided into 402m, 804m, or 1608m plots
 			\item Presence of track in each plot determined
+			\item Also examined data set with every other plot excluded
+			(Alternating Plots)
 		\end{itemize}
@@
 
-\begin{frame}{Issues related to data}
-	\begin{itemize}
-		\item Tracks near plot boundary can be assigned to different plots
-		\item Track in one site means more likely to be track in neighboring
-		site -- not independent
-	\end{itemize}
+{
+\usebackgroundtemplate{\includegraphics[width=\paperwidth]{data.pdf}}
+\begin{frame}
+    \begin{picture}(310,190)(9,7.5)
+        \linethickness{1.5pt}
+        \only<2->{\color<2>{red}\put(50,120){\framebox(290,15){}}}
+        \only<3->{\color<3>{red}\put(150,-20){\framebox(15,80){}}}
+        \only<4->{\color<4>{red}\put(150,-35){\framebox(103,30){}}}
+    \end{picture}
 \end{frame}
+}
 
@@
 	(MacKenzie, et al. 2006)
+	\begin{center}
+		\resizebox{5cm}{!}{\includegraphics{cuteOtter3.jpg}}
+	\end{center}
 \end{frame}
@@
 \begin{frame}{Assumptions of occupancy models}
+	\begin{columns}
+		\column{5cm}
 		\begin{itemize}
@@
 		\end{itemize}
+		\column{5cm}
+		\resizebox{5cm}{!}{\includegraphics{cuteOtter4.jpg}}
+	\end{columns}
 \end{frame}
@@
 	\end{itemize}
+	\begin{center}
+		\resizebox{7cm}{!}{\includegraphics{../SimpleModelWithParams.pdf}}
+	\end{center}
 \end{frame}
@@
 	\begin{center}
-		\resizebox{5cm}{!}{\includegraphics{../observerPlots.pdf}}
+		\resizebox{5cm}{!}{\includegraphics{wwObsPlot.pdf}}
 	\end{center}
@@
 \begin{frame}{Correlation between flights}
+	\begin{center}
+		\resizebox{7cm}{!}{\includegraphics{dayEventGraphs.pdf}}
+	\end{center}
 	\begin{itemize}
-		\item Two observers on same day should have the same na\"ive occupancy
-		rates; not the case
 		\item Also looked at correlation between flights
@@
 
+\begin{frame}{Hierarchical model with false detection}
+	\begin{center}
+		\resizebox{10cm}{!}{\includegraphics{../ModelWithEAndParams.pdf}}
+	\end{center}
+\end{frame}
+
 \begin{frame}{Independence between plots}
@@
 			\begin{itemize}
-				\item $Z=\frac{\hat{S}-E[S]}{SD[S]}$
+				\item $Z=\frac{S-E[S]}{SD[S]}$
 				\item $|Z_i|>2$ suggests assumption violated for flight $i$.
@@
 \begin{frame}{Independence, continued}
-
+	\begin{center}
+		\includegraphics[scale=0.4]{independenceBoxplots.pdf}
+	\end{center}
 	\begin{itemize}
-		\item Looking at alternating plots increases the independence
-		\item Also saw trends with respect to plot size
-		\item Solution: kept issue in mind, explored spatially explicit models
+		\item Solution: keep issue in mind, include spatially dependent models
 	\end{itemize}
@@
 \section{Modeling the data}
-\begin{frame}{Introduction to Bayesian statistics}
-	\begin{itemize}
-		\item Under Bayesian statistics, parameters are random variables
-		\item According to Bayes' theorem
-		\begin{itemize}
-			\item $\text{posterior distribution} \propto \text{prior
-			distribution}\times \text{likelihood}$
-		\end{itemize}
-	\end{itemize}
-\end{frame}
 
@@
 
+\begin{frame}{Introduction to Bayesian statistics}
+	\begin{itemize}
+		\item Under Bayesian statistics, parameters are random variables
+		\item According to Bayes' theorem
+		\begin{itemize}
+			\item $\text{posterior distribution} \propto \text{prior
+			distribution}\times \text{likelihood}$
+		\end{itemize}
+		\item Goal: Estimate the posterior distribution
+		\begin{itemize}
+			\item Simulate a Markov Chain for many iterations
+			\item Will approach true distribution of random variable
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Convergence}
+	\begin{center}
+		\resizebox{9cm}{!}{\includegraphics{historyConvergence.pdf}} \\
+		\includegraphics[scale=0.8]{distnConvergence.pdf} \\
+	\end{center}
+\end{frame}
+
+\begin{frame}{Convergence}
+	\begin{center}
+		\resizebox{9cm}{!}{\includegraphics{historyNoConvergence.pdf}} \\
+		\includegraphics[scale=0.8]{distnNoConvergence.pdf} \\
+	\end{center}
+\end{frame}
+
 \begin{frame}{Issue with false detection}
@@
 %This slide for DNR only
-\begin{frame}{Bayesian model specifics}
-	\begin{itemize}
-		\item $\psi,\theta \sim \text{Uniform}(0,1)$
-		\item $p[\text{obs}] \sim \text{Beta}(a,b)$
-		\begin{itemize}
-			\item $a, b \sim \text{Gamma}(1,0.1)$
-			\item E similar
-			\item $p$ and $E$ had restricted distributions
-		\end{itemize}
-		\item Burn-in period 10,000 iterations
-		\item Covariates: observer, days since snow
-	\end{itemize}
-\end{frame}
+%\begin{frame}{Bayesian model specifics}
+%	\begin{itemize}
+%		\item $\psi,\theta \sim \text{Uniform}(0,1)$
+%		\item $p[\text{obs}] \sim \text{Beta}(a,b)$
+%		\begin{itemize}
+%			\item $a, b \sim \text{Gamma}(1,0.1)$
+%			\item E similar
+%			\item $p$ and $E$ had restricted distributions
+%		\end{itemize}
+%		\item Burn-in period 10,000 iterations
+%		\item Covariates: observer, days since snow
+%	\end{itemize}
+%\end{frame}
 
@@
 	\end{itemize}
+\end{frame}
+
+\begin{frame}{Testing the model, continued}
+	\begin{center}
+	\includegraphics[scale=0.5]{../SimulatedData.pdf}
+	\end{center}
 \end{frame}
@@
 %DNR only slide
-\begin{frame}{CAR model specifics}
-	\begin{itemize}
-		\item $\tau^2 =$ conditional variance
-		\item $(\tau^2)^{-1} \sim \text{Gamma}(0.5,0.0005)$
-		\item $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$
-		\item $\alpha_0 \sim$ Improper uniform
-		\item $\alpha_1 \sim$ Gaussian CAR
-	\end{itemize}
-\end{frame}
+%\begin{frame}{CAR model specifics}
+%	\begin{itemize}
+%		\item $\tau^2 =$ conditional variance
+%		\item $(\tau^2)^{-1} \sim \text{Gamma}(0.5,0.0005)$
+%		\item $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$
+%		\item $\alpha_0 \sim$ Improper uniform
+%		\item $\alpha_1 \sim$ Gaussian CAR
+%	\end{itemize}
+%\end{frame}
 
-%DNR slide
 \begin{frame}{Testing the CAR model}
@@
 		\item CAR model did not converge
-		\item Simulated data more independent than actual data (from $\hat{S}$)
+		\item Simulated data more independent than actual data (from $S$)
 		\item Added similar correlation in $\psi$
@@
 		\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Testing CAR model, continued}
-	\begin{itemize}
-		\item With correlated data, no constant $\psi$.
-		\item To find true $\psi$ from simulated data, took proportion of plots
-		occupied
 		\item When CAR model converged, estimates from CAR and standard models
@@
 
-%Carleton slide goes here
 \section{Discussion}
@@
 
+\begin{frame}{}
+    % Mississippi River!
+	\small{
+    \begin{table}
+    \begin{center}
+    \begin{tabular}{|l|l|l|l|}
+        \hline
+        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
+        \hline
+            & 402m & 804m & 1608m \\
+        \hline
+        \multirow{2}{*}{E[Danny]}
+            & 0.01954 & 0.04808 & 0.06885 \\
+            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
+        \hline
+        \multirow{2}{*}{E[John]}
+            & 0.05216 & 0.07771 & 0.13650 \\
+            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
+        \hline
+        \multirow{2}{*}{E[Tom]}
+            & 0.02668 & 0.06140 & 0.10230 \\
+            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
+        \hline
+        \multirow{2}{*}{p[Danny]}
+            & 0.65240 & 0.67340 & 0.74960 \\
+            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
+        \hline
+        \multirow{2}{*}{p[John]}
+            & 0.67770 & 0.77580 & 0.82540 \\
+            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
+        \hline
+        \multirow{2}{*}{p[Tom]}
+            & 0.53670 & 0.54220 & 0.65330 \\
+            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
+        \hline
+        \multirow{2}{*}{\(\psi\)}
+            & 0.59380 & 0.69370 & 0.75820 \\
+            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
+        \hline
+        \multirow{2}{*}{\(\theta\)}
+            & 0.15280 & 0.20730 & 0.27690 \\
+            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
+        \hline
+    \end{tabular}
+    \end{center}
+	\end{table} 
+	}
+\end{frame}
+
+% Add slide with example of estimates for our actual data, one river/year?
+
 \begin{frame}{Summary and interpretation of estimates}
+	\begin{center}
+		\includegraphics[scale=0.5]{Aggregation.pdf}
+	\end{center}
 	\begin{itemize}
 		\item Estimates for all parameters increase with plot size
-		\item Possible hypotheses...
+		\item Possible hypotheses:
+		\begin{itemize}
+			\item \(\psi\) and \(\theta\)
+			\item \(p\) and \(E\)
+		\end{itemize}
 		\item Used data from all plots, not alternating plots
@@
 		\end{enumerate}
+		\item We suggest: 1 observer, 3 snow events, 3 days since snow
 		\item All conclusions come from regular data
@@
 	\end{itemize}
+	\begin{center}
+		\resizebox{7cm}{!}{\includegraphics{cuteOtter10.jpg}}
+	\end{center}
 \end{frame}
@@
 \begin{frame}{Questions?}
+	\begin{center}
+	\includegraphics[scale=0.7]{cuteOtter7.jpg}
+	\end{center}
 \end{frame}

### 2010-03-04 Presentation/Presentation.tex
Hopefully fixed the crash-happy convergence graphic.
Edited baysian slide
Changed aggregation slide
A few more changes
poop

@@
 		\item Ecologists often conduct surveys to find the occupancy rate
+		\item Na\"ive occupancy: proportion of sites found to be occupied
 		\item Not observing the species can mean two things:
@@
 		\item Single species, single season
-		\item Na\"ive occupancy: proportion of sites found to be occupied
 		\item Hierarchical models
@@
 	\begin{itemize}
-		\item $\psi=$ The probability a track is occupied
+		\item $\psi=$ The probability a site is occupied
 		\item $\theta=$ The probability a track is laid down in a site on a
@@
 		\end{itemize}
+		\item Spatial correlation is easier with a Bayesian model
 	\end{itemize}
@@
 	\begin{center}
-		\resizebox{9cm}{!}{\includegraphics{historyConvergence.pdf}} \\
+		\resizebox{8.5cm}{!}{\includegraphics{historyConvergence.pdf}} \\
 		\includegraphics[scale=0.8]{distnConvergence.pdf} \\
@@
 	\begin{center}
-		\resizebox{9cm}{!}{\includegraphics{historyNoConvergence.pdf}} \\
+		\resizebox{8.5cm}{!}{\includegraphics{historyNoConvergence.pdf}} \\
 		\includegraphics[scale=0.8]{distnNoConvergence.pdf} \\
@@
 		$(p=y,E=x,\psi=1-z)$.
-		\item Leads to bimodal posterior distributions for $p$ and $E$.
+		\item Leads to bimodal posterior distributions.
 		\item Solution: restrict $p$ to interval $(0.6,1)$, $E$ to interval
@@
 		\item CAR model did not converge
-		\item Simulated data more independent than actual data (from $S$)
+		\item Simulated data less dependent than actual data (from $S$)
 		\item Added similar correlation in $\psi$
@@
 			\item CAR model converged more often
-			\item Independence more similar to that of actual data
+			\item Dependence more similar to that of actual data
 		\end{itemize}
@@
 		\begin{itemize}
-			\item Removed false detection, 402m in 2004 converged
+			\item Removed false detection
+			\item 402m in 2004 converged, but estimates not dependable
 		\end{itemize}
@@
 \begin{frame}{Summary and interpretation of estimates}
-	\begin{center}
-		\includegraphics[scale=0.5]{Aggregation.pdf}
-	\end{center}
 	\begin{itemize}
@@
 	\end{itemize}
+	\begin{center}
+		\includegraphics[scale=0.5]{Aggregation.pdf}
+	\end{center}
 \end{frame}

### 2010-03-04 Presentation.tex
Massive anal reorganization, added psi estimate graph.

--- 2010-03-04 Presentation/Presentation.tex
@@
 \institute{Carleton College}
-\usepackage{graphics}
+\usepackage{graphics,graphpap}
 \usepackage{color}
 \usepackage{multirow}
+\usepackage[T1]{fontenc}
+\usepackage{fix-cm}
 %\includeonlyframes{currentslide}
@@
 	\usebackgroundtemplate{
-	\includegraphics[width=\paperwidth]{ottertracks1.jpg}}
+	\includegraphics[width=\paperwidth]{Figures/Pictures/ottertracks1.jpg}}
 	\begin{frame}
@@
 		\column{5cm}
-		\resizebox{5cm}{!}{\includegraphics{cuteOtter1.pdf}}
+		\resizebox{5cm}{!}{\includegraphics{Figures/Pictures/cuteOtter1.pdf}}
 	\end{columns}
+\end{frame}
+
+\begin{frame}{Map of area}
+	\begin{picture}(310,190)(9,7.5)
+		\put(20,-5){\resizebox{10cm}{!}{\includegraphics{Figures/map.pdf}}}
+		\only<2>{\put(164,48){\circle{30}}}
+	\end{picture}
 \end{frame}
@@
 		\column{5cm}
-		\resizebox{5cm}{!}{\includegraphics{cuteOtter2.jpg}}
+		\resizebox{5cm}{!}{\includegraphics{Figures/Pictures/cuteOtter2.jpg}}
 		\column{5cm}
@@
 	\usebackgroundtemplate{
-	\includegraphics[width=\paperwidth]{SnowTracks12.jpg}}
+	\includegraphics[width=\paperwidth]{Figures/Pictures/SnowTracks12.jpg}}
 	\begin{frame}{}
@@
 	\usebackgroundtemplate{
-	\includegraphics[width=\paperwidth]{SNOWTRACKS9.jpg}}
+	\includegraphics[width=\paperwidth]{Figures/Pictures/SNOWTRACKS9.jpg}}
 	\begin{frame}
@@
 {
-\usebackgroundtemplate{\includegraphics[width=\paperwidth]{data.pdf}}
+\usebackgroundtemplate{\includegraphics[width=\paperwidth]{Figures/data.pdf}}
 \begin{frame}
@@
 	\begin{center}
-		\resizebox{5cm}{!}{\includegraphics{cuteOtter3.jpg}}
+		\resizebox{5cm}{!}{\includegraphics{Figures/Pictures/cuteOtter3.jpg}}
 	\end{center}
@@
 		\column{5cm}
-		\resizebox{5cm}{!}{\includegraphics{cuteOtter4.jpg}}
+		\resizebox{5cm}{!}{\includegraphics{Figures/Pictures/cuteOtter4.jpg}}
 	\end{columns}
@@
 	\begin{center}
-		\resizebox{7cm}{!}{\includegraphics{../SimpleHierarchicalModel.pdf}}
+		\resizebox{7cm}{!}{
+		\includegraphics{Figures/Diagrams/SimpleHierarchicalModel.pdf}}
 	\end{center}
@@
 	\begin{center}
-		\resizebox{7cm}{!}{\includegraphics{../SimpleModelWithParams.pdf}}
+		\resizebox{7cm}{!}{
+		\includegraphics{Figures/Diagrams/SimpleModelWithParams.pdf}}
 	\end{center}
@@
 	\begin{center}
-		\resizebox{5cm}{!}{\includegraphics{wwObsPlot.pdf}}
+		\resizebox{5cm}{!}{\includegraphics{Figures/R/wwObsPlot.pdf}}
 	\end{center}
@@
 	\begin{center}
-		\resizebox{7cm}{!}{\includegraphics{dayEventGraphs.pdf}}
+		\resizebox{7cm}{!}{\includegraphics{Figures/R/dayEventGraphs.pdf}}
 	\end{center}
@@
 	\begin{center}
-		\resizebox{10cm}{!}{\includegraphics{../ModelWithE.pdf}}
+		\resizebox{10cm}{!}{\includegraphics{Figures/Diagrams/ModelWithE.pdf}}
 	\end{center}
@@
 	\begin{center}
-		\resizebox{10cm}{!}{\includegraphics{../ModelWithEAndParams.pdf}}
+		\resizebox{10cm}{!}{
+		\includegraphics{Figures/Diagrams/ModelWithEAndParams.pdf}}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[scale=0.4]{independenceBoxplots.pdf}
+		\includegraphics[scale=0.4]{Figures/R/independenceBoxplots.pdf}
 	\end{center}
@@
 	\begin{center}
-		\resizebox{8.5cm}{!}{\includegraphics{historyConvergence.pdf}} \\
-		\includegraphics[scale=0.8]{distnConvergence.pdf} \\
+		\resizebox{8.5cm}{!}{
+		\includegraphics{Figures/R/historyConvergence.pdf}} \\
+		\includegraphics[scale=0.8]{Figures/R/distnConvergence.pdf} \\
 	\end{center}
@@
 	\begin{center}
-		\resizebox{8.5cm}{!}{\includegraphics{historyNoConvergence.pdf}} \\
-		\includegraphics[scale=0.8]{distnNoConvergence.pdf} \\
+		\resizebox{8.5cm}{!}{
+		\includegraphics{Figures/R/historyNoConvergence.pdf}} \\
+		\includegraphics[scale=0.8]{Figures/R/distnNoConvergence.pdf} \\
 	\end{center}
@@
 \begin{frame}{Issue with false detection}
+	\begin{columns}
+		\column{5cm}
 		\begin{itemize}
@@
 		\end{itemize}
+		\column{5cm}
+		\includegraphics[width=5cm]{Figures/Pictures/cuteOtter6.jpg}
+	\end{columns}
 \end{frame}
 
-%This slide for DNR only
+% DNR-only slide!
 %\begin{frame}{Bayesian model specifics}
@@
 	\begin{center}
-	\includegraphics[scale=0.5]{../SimulatedData.pdf}
+		\includegraphics[width=8.5cm]{Figures/Diagrams/SimulatedData.pdf}
 	\end{center}
@@
 
-%DNR only slide
+% DNR-only slide!
 %\begin{frame}{CAR model specifics}
@@
 		\item All of 2003 data converged with standard model
-		\item Nothing in 2004 converged
+		\item None of 2004 data converged with standard model
 		\begin{itemize}
-			\item Removed false detection
-			\item 402m in 2004 converged, but estimates not dependable
+			\item Removed false detection. 2004 data with 402m plot size
+			converged, but estimates not dependable
 		\end{itemize}
@@
 
-\begin{frame}{}
-    % Mississippi River!
-	\small{
-    \begin{table}
+\begin{frame}{Summary and interpretation of estimates}
 	\begin{center}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
-        \hline
-            & 402m & 804m & 1608m \\
-        \hline
-        \multirow{2}{*}{E[Danny]}
-            & 0.01954 & 0.04808 & 0.06885 \\
-            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
-        \hline
-        \multirow{2}{*}{E[John]}
-            & 0.05216 & 0.07771 & 0.13650 \\
-            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
-        \hline
-        \multirow{2}{*}{E[Tom]}
-            & 0.02668 & 0.06140 & 0.10230 \\
-            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
-        \hline
-        \multirow{2}{*}{p[Danny]}
-            & 0.65240 & 0.67340 & 0.74960 \\
-            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
-        \hline
-        \multirow{2}{*}{p[John]}
-            & 0.67770 & 0.77580 & 0.82540 \\
-            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
-        \hline
-        \multirow{2}{*}{p[Tom]}
-            & 0.53670 & 0.54220 & 0.65330 \\
-            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
-        \hline
-        \multirow{2}{*}{\(\psi\)}
-            & 0.59380 & 0.69370 & 0.75820 \\
-            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
-        \hline
-        \multirow{2}{*}{\(\theta\)}
-            & 0.15280 & 0.20730 & 0.27690 \\
-            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
-        \hline
-    \end{tabular}
+		\includegraphics[width=10cm]{Figures/Graphs/2003Psi.pdf}
 	\end{center}
-	\end{table} 
-	}
 \end{frame}
-
-% Add slide with example of estimates for our actual data, one river/year?
 
@@
 	\begin{center}
-		\includegraphics[scale=0.5]{Aggregation.pdf}
+		\includegraphics[width=10cm]{Figures/Diagrams/Aggregation.pdf}
 	\end{center}
@@
 	\begin{center}
-		\resizebox{7cm}{!}{\includegraphics{cuteOtter10.jpg}}
+		\includegraphics[width=7cm]{Figures/Pictures/cuteOtter10.jpg}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[scale=0.7]{cuteOtter7.jpg}
+		\includegraphics[width=10cm]{Figures/Pictures/cuteOtter7.jpg}
 	\end{center}

### 2010-03-04 Presentation.tex
Last revisions?

@@
 \section{Discussion}
+
+\begin{frame}{Obtaining convergence on the data}
+	\begin{itemize}
+		\item Occupancy data from the winters of 2003 and 2004
+		\item 2003 data set consists mostly of multiple observers flying on
+		multiple days after each snow event
+		\item 2004 data set consists mostly of one observer flying on one day
+		after each snow event
+	\end{itemize}
+\end{frame}
+
 \begin{frame}{Obtaining convergence on the data}
@@
 
-\begin{frame}{Summary and interpretation of estimates}
+\begin{frame}{Summary of estimates}
+	\begin{center}
+		\includegraphics[width=8cm]{Figures/R/MissResults.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Summary of estimates}
 	\begin{center}
@@
 
-\begin{frame}{Summary and interpretation of estimates}
+\begin{frame}{Interpretation of estimates}
 	\begin{itemize}
@@
 		\begin{itemize}
-			\item More data more useful than more independence
+			\item More data more useful than less dependence
 		\end{itemize}

### 2010-03-07 Paper.tex
Fixed code and whitespace problems.

@@
 
-% This should be a lot prettier.
 \title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
@@
     \begin{center}
-        \includegraphics[scale=0.75]{SimpleHierarchicalModel}
+        \includegraphics[scale=0.75]{Figures/Diagrams/SimpleHierarchicalModel}
         \caption{A visualization of how the hierarchical model works.}
@@
         \centering
-        \includegraphics[width=5in]{observerPlots.pdf}
+        \includegraphics[width=5in]{Figures/R/observerPlots.pdf}
         \caption{The box plots show how the proportion of plots in which an
@@
     for (i in 1:numDataPts) {
-        # equals(track.count[...], 0) == 1 when there is no track to observe.
+        # Positive detection.
         mu[i] <-
             p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
-        # probability of false detection, given track and occupancy state
+        # False detection.
         nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
@@
 
-writeModel(otterModel, "irregularHierarchicalModel.txt")
-modelCheck("irregularHierarchicalModel.txt")
+writeModel(otterModel, "HierarchicalModel.txt")
+modelCheck("HierarchicalModel.txt")
 bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
            "numDataPts", "snowEvents", "daysAfterTotal"),
-         "irregularOtterData.txt")
-modelData("irregularOtterData.txt")
+         "OtterData.txt")
+modelData("OtterData.txt")
 modelCompile(numChains=2)
@@
     b ~ dgamma(1,.1)
-    #c ~ dgamma(1,.1)
-    #d ~ dgamma(1,.1)
+    c ~ dgamma(1,.1)
+    d ~ dgamma(1,.1)
     for (observer in 1:observers) {
@@
     for (i in 1:numDataPts) {
-        # equals(track.count[...], 0) == 1 when there is no track to observe.
         mu[i] <-
             p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
-        # probability of false detection, given track and occupancy state
         nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
@@
 
-writeModel(otterModel, "irregularHierarchicalModel.txt")
-modelCheck("irregularHierarchicalModel.txt")
+writeModel(otterModel, "CARModel.txt")
+modelCheck("CARModel.txt")
 bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
@@
            "sumNumNeigh"),
-         "irregularOtterData.txt")
-modelData("irregularOtterData.txt")
+         "OtterData.txt")
+modelData("OtterData.txt")
 modelCompile(numChains=2)
 
-init.val <- function(){
-	list(
-		alpha1=runif(sites,-100,100),
-		alpha0=runif(1,-100,100),
-		tau=rgamma(1,.5,.0005),
-		psi=runif(1),
-		a=rgamma(1,1,.1),
-		b=rgamma(1,1,.1)
-		#,c=rgamma(1,1,.1),
-		#d=rgamma(1,1,.1)
-	) 
-}
-bugsInits(init.val,numChains=2,fileName=c('init1.txt','init2.txt'))
-modelInits(c('init1.txt','init2.txt'))
+init.val <- function() {
+list(alpha1 = runif(sites,-100,100),
+     alpha0 = runif(1,-100,100),
+     tau = rgamma(1,.5,.0005),
+     psi = runif(1),
+     a = rgamma(1, 1, .1),
+     b = rgamma(1, 1, .1),
+     c = rgamma(1, 1, .1),
+     d = rgamma(1, 1, .1))
+}
+bugsInits(init.val, numChains=2, fileName=c('init1.txt','init2.txt'))
+modelInits(c('init1.txt', 'init2.txt'))
 
@@
 modelUpdate(12000)
-#samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi"))
-samplesSet(c("a", "b", "p[]", "psi"))
+samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi"))
 modelUpdate(10000)
@@
 
+# This is an example of regular data.  Three observers, six snow events, two
+# observations on consecutive days after each snow event.
 eventsToDaysBegin <- c(1, 3, 5, 7, 9, 11)
@@
 beta <- .4
-condlOcc <- psi + beta
-condlUnocc <- psi
-
 theta <- .5
-alpha <- .3
-condlTrack <- theta + alpha
-condlNoTrack <- theta
 p <- c(.70, .80, .90)
@@
 for (site in 1:sites) {
-    # TRUE == 1, FALSE == 0.
-	if (site == 1) {
-		occupancy[site] <- 1 * (runif(1) < (condlOcc + condlUnocc/2))
-	} else if (occupancy[i-1] == 1) {
-		occupancy[site] <- 1 * (runif(1) < condlOcc)
-	} else
-		occupancy[site] <- 1 * (runif(1) < condlUnocc)
-	}
+    occupancy[site] <- 1 * (runif(1) < psi)
     for (event in 1:snowEvents) {
@@
         currentDay <- 0
-        # Haha, this is awful.
         for (day in 1:curDaySinceSnow[eventsToDaysBegin[event]]) {
@@
 \begin{verbatimtab}
-#Define structure
+# Define structure.
 sites <- 60
@@
 
-#Define parameters
+# Define parameters.
 psi <- .4
@@
 condl.notrack <- theta
-p <- c(.7,.7,.7)
-E <- c(.05,.05,.05)
-
-#Initialize variables
+p <- c(.7, .7, .7)
+E <- c(.05, .05, .05)
+
+# Initialize variables.
 occupancy <- NULL
@@
 
-for(i in 1:sites) {
-	#Determine occupancy of site. 1*(runif(1)<psi) is 1 around psi*100 percent of the time.
-	#Use conditional rates depending on occupancy of site i-1
-	if (i==1) 
-		{occupancy[i] <- 1*(runif(1)<(condl.occ+condl.unocc)/2)}	
-	else if (occupancy[i-1]==1) 
-		{occupancy[i] <- 1*(runif(1)<condl.occ)}
+for (i in 1:sites) {
+    # Determine occupancy of site. 1 * (runif(1) < psi) is 1 around psi*100
+    # percent of the time. Use conditional rates depending on occupancy of site
+    # i-1.
+    if (i == 1)
+        {occupancy[i] <- 1 * (runif(1) < (condl.occ + condl.unocc) / 2)}
+    else if (occupancy[i-1] == 1)
+        {occupancy[i] <- 1 * (runif(1) < condl.occ)}
     else
-		{occupancy[i] <- 1*(runif(1)<condl.unocc)}
-	for(j in 1:events) {
+        {occupancy[i] <- 1 * (runif(1) < condl.unocc)}
+
+    for (j in 1:events) {
         is.track <- 0
-		for(k in 1:snowdays) {
-			index <- k + snowdays*(j-1)
-			#Decide if a track is laid down on day k of event j, given the occupancy state
-			if(i==1) {
-				tracks[i,index] <- (runif(1)<(condl.track+condl.notrack)/2)*occupancy[i]
-			}else if(tracks[i-1,index]==1) 
-				{tracks[i,index] <- (runif(1)<condl.track)*occupancy[i]
-			}else 
-				{tracks[i,index] <- (runif(1)<condl.notrack)*occupancy[i]}
-
-			#Determine if any tracks have been laid down for event j.
+        for (k in 1:snowdays) {
+            index <- k + snowdays * (j - 1)
+            # Decide if a track is laid down on day k of event j, given the
+            # occupancy state.
+            if (i == 1) {
+                tracks[i, index] <- (runif(1) <
+                                     (condl.track + condl.notrack) / 2) *
+                                    occupancy[i]
+            } else if (tracks[i-1, index] == 1)
+                {tracks[i, index] <- (runif(1) < condl.track) * occupancy[i]
+            } else
+                {tracks[i, index] <- (runif(1) < condl.notrack) * occupancy[i]}
+
+            # Determine if any tracks have been laid down for event j.
             is.track <- max(is.track, tracks[i,index])
-			for(m in 1:obs) {
-				index <- m + obs*(k-1) + snowdays*obs*(j-1)
-				#Determine if the observer saw a track.
-				y[i,index] <- (runif(1)<p[m])*is.track + (runif(1)<E[m])*(1-is.track)
+            for (m in 1:obs) {
+                index <- m + obs * (k - 1) + snowdays * obs * (j - 1)
+                # Determine if the observer saw a track.
+                y[i, index] <- (runif(1) < p[m]) * is.track +
+                               (runif(1) < E[m]) * (1-is.track)
             }

### 2010-03-12 Paper.tex
Fixed code and all whitespace problems.

@@
     simplification results in estimates of \(\psi\) that do not seem accurate
-    and are not directly comparable to estimates of \(\psi\) on 2003's data.
+    and are not directly comparable to estimates of \(\psi\) on 2003 data.
 
@@
     likely for an observer to detect a track on later days -- to be thoroughly
-    tested. Most of the \(\psi\) and \(\theta\) estimates of 2003's data fall
+    tested. Most of the \(\psi\) and \(\theta\) estimates of 2003 data fall
     into ranges where our model is at its best, both with respect to convergence
@@
 samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi", "theta"))
-#samplesSet(c("a", "b", "p[]", "psi", "theta"))
 modelUpdate(10000)

### 2010-06-08 NewPaper.tex
First draft.

\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{natbib}
\usepackage{verbatim}
\usepackage{hyperref}

\renewcommand{\bibsection}{\section{References}}

\begin{document}
\addtolength{\textheight}{+0.18in}

\newcommand{\lf}{\left\lfloor}
\newcommand{\rf}{\right\rfloor}
\newcommand{\dis}{\displaystyle}
\newcommand{\sums}[2]{\sum_{\stackrel{{\scriptstyle {#1}}}{{#2}}}}
\newcommand{\prods}[2]{\prod_{\stackrel{{\scriptstyle {#1}}}{{#2}}}}

\long\def\symbolfootnote[#1]#2{\begingroup%
\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}

\def\prbox{\hfill\rule{1.2ex}{1.2ex}\vspace{.2in}}

\newenvironment{proof}{\noindent{\bf Proof }}{\prbox}

\def\Reals{\hbox{\rm I\kern-.18em R}}
\def\reals{\Reals}
\def\Complexes{\hbox{\rm C\kern-.43em
       \vrule depth 0ex height 1.4ex width .05em\kern.41em}}
\def\complexes{\Complexes}

\def\field{\hbox{\rm I\kern-.18em F}} %symbol for field
\def\Naturals{\hbox{\rm I\kern-.17em N}}
\def\naturals{\Naturals}
\def\integers{\hbox{\rm Z\kern-.3em Z}}
\def\hh{\hrule height0.9pt width1.1em}
\def\vv{\vrule width0.8pt depth0.12em height0.92em}
\def\square{\vbox{\kern0.15em\hh\kern0.9em\hh\kern-1.05em
            \hbox{\vv\kern0.93em\vv}}}
\def\Lv{L_{\alpha} }
\def\Lvp{L_{\alpha'} }
\def\sm{{\mbox{\boldmath $\sigma$}_m}}
\def\MMv{R_{\mbox{\boldmath $\sigma$}_v}}
\def\tMMv{\widetilde{R}_{\mbox{\boldmath $\sigma$}_v}}
\def\MMm{R_{\mbox{\boldmath $\sigma$}_m}}
\def\tMMm{\widetilde{R}_{\mbox{\boldmath $\sigma$}_m}}
\def\MMj{R_{\mbox{\boldmath $\sigma$}_j}}
\def\sn1{{\mbox{\boldmath $\sigma$}_{n-1}}}
\def\snv{{\mbox{\boldmath $\sigma$}_v}}

\hyphenation{eigen-value so-lu-tions heur-is-tics mod-eled de-note
prob-a-bil-ity start-ing}

\LARGE\bf
\noindent
\textsf{A Bayesian hierarchical occupancy model for track surveys conducted in
a series of linear, spatially correlated, sites}

\rm\large
\vspace{2ex}
\noindent
\textsf{[short title: Bayesian occupancy model for track surveys]}

\rm\large
\normalsize

\vspace{4ex}
\small\bf
\noindent
\textsf{Chrisna Aing\footnotemark[1]\footnotetext[1]{Department of Mathematics,
Carleton College, Northfield, MN 55057, USA}, Sarah Halls\footnotemark[1], Kiva
Oken\footnotemark[1], Robert Dobrow\footnotemark[1]\symbolfootnote[1]
{Correspondence author: E-mail: rdobrow@carleton.edu},
John Fieberg \footnotemark[2]\footnotetext[2]{Minnesota Department of Natural
Resources}}

\normalsize\rm
\vspace{2ex}
\noindent
\rule{5in}{.4mm}

\vspace{3ex}
\begin{quote}
\noindent
{\bf Summary}
\begin{enumerate}
    \item In creating an occupancy model for river otters ({\em Lontra
        canadensis}) in southern Minnesota we develop an analytical approach
        that can be used in conjunction with survey data collected in a set of
        linear aerial survey sites.
    \item Our approach incorporates several recent developments in occupancy
        modeling, including the prevalence of false positives, spatial
        correlation, and repeated sampling across distinct observers and across
        daily snow events.
    \item We test our method using simulated and real data. Simulations across
        a wide range of parameters show the model to be robust, both with
        regard to false positive and negatives, and with regard to the type of
        spatial correlation observable in our data.  Confidence intervals for
        both the false positive and false negative rates are remarkably small
        given the relative paucity of the data.
    \item Our approach lends itself to investigating the optimal allocation of
        sampling effort and we make several recommendations for future studies.
\end{enumerate}

\vspace{2ex}
\noindent
{\bf Key-words:} Hierarchical modeling, occupancy, otter, simulation, spatial
correlation, track survey
\end{quote}

\normalsize
\section{Introduction}

\textbf{[Should we add all of these sources to BibTeX?]}

Wildlife populations are notoriously difficult to monitor, owing to challenges
of detecting individuals and the costs of surveying difficult to reach habitat.
A variety of approaches have been developed for estimating detection
probabilities in wildlife surveys, include double sampling, removal methods,
multiple observers, mark-recapture, distance sampling, and regression models
\citep{Lancia2005}.

\textbf{[The first citation below needs Thompson et al. 2004.]}

These methods, however, typically require identifying and possibly marking
unique individuals, and are therefore difficult to reliably implement in large
scale monitoring efforts or with elusive species \citep{Stanley2005,
Johnson2008}.  As a result, survey designs that record presence or absence of
animal sign (e.g., tracks, scent marks, feces) remain popular among natural
resource agencies \citep{Stanley2005}.  A flurry of recent research has been
devoted to developing methods, most often based on a survey design in which
sites are repeatedly visited, to correct for uncertain detection in
presence-absence surveys; the collection of methods for monitoring populations
using repeated presence-absence surveys has come to be known as ``occupancy
modeling'' \citep{MacKenzie2006, Royle2008}.

The primary purpose of this paper is to illustrate the development of an
occupancy modeling approach to analyzing river otter (\textit{Lontra
canadensis}) track survey data collected in the winter following snowfall
events.  Our approach will incorporate several recent developments in occupancy
modeling, including the prevalence of false detections \citep{Royle2006},
spatial correlation \citep{Magoun2007}, and a model for how detection
probabilities depend on time since a snowfall event \citep{Stanley2005}.

The data we consider were collected by the Minnesota Department of Natural
Resources (MNDNR) in 2003 using helicopter surveys with multiple observers,
multiple sampling intervals (representing unique snowfall events), and multiple
days within each sampling interval (days since the last snowfall event).
Flight paths followed the main channels of river systems, which were later
divided into a series of spatially contiguous quadrats (hereafter, we will
refer to these quadrats as ``sites'').  Due to potential for clustering of
individual home ranges in space and movement across site boundaries, we
expected the response, presence of a track (Y/N), to be spatially correlated
among sites.  In addition, sites were more likely to contain tracks if they
were sampled several days after a snowfall event.  Lastly, observer-to-observer
variability was significant, likely owing to measurement error associated with
the post-hoc creation of site boundaries and the difficulty of correctly
identifying otter tracks \citep{Evans2009}.  Thus, these data serve as a
useful case study to illustrate methods for handling several complications that
may be present in repeated presence-absence survey applications.

In addition to fitting a series of models to the Minnesota otter data, we use
extensive simulations to evaluate the robustness and precision of model-based
estimates of occupancy probabilities under this sampling design.  To promote
learning, and to facilitate adaption to other applications, we implement our
approach using open source software, Program R \citep{R2009} and WinBUGS
\citep{Lunn2000}, with the R package BRugs \citep{Thomas2006} to communicate
between the two software platforms.  Lastly, we show how simulations can be
used to compare different sampling designs, and we make recommendations
regarding allocation of sampling effort in future surveys.

\section{Materials and Methods}

    \subsection{Otter survey data collection}

    \textbf{[Need to cite Danny's dissertation (see next line).]}

    Details of the survey are given in \citet{Martin2003} and Martin (cite
    Danny's dissertation), but generally, aerial snow-track surveys were
    conducted using a Bell OH-58A+ helicopter following snowfalls of greater
    than 2.5 cm in depth.  The data discussed in this paper were collected in
    2003 from the Mississippi River.  There were five unique snowfall events.
    Flights occurred one, two or three days after each unique snowfall event,
    although they did not occur on every day.  There were three observers, and
    in some cases multiple observers flew over the river in different flights
    on the same day.  Observers collected a Global Positioning System waypoint
    upon encountering a track, and continued to log waypoints every five
    seconds thereafter if a track remained present.  After these waypoints were
    logged, three different datasets were created.  The section of the
    Mississippi River that was flown was divided into 402 meter sites, 804
    meter sites, and 1608 meter sites.  The presence (1) or absence (0) of a
    waypoint in each site on a given flight was determined and recorded.

    \subsection{Model development and likelihood formulation}

    Fundamentally, these data consist of a set of correlated binary (0, 1)
    responses arising from a set of biological and observational processes.  To
    construct the likelihood, we need to think about the different ways of
    arriving at a 1 (recorded presence of a track) or a 0 (absence of an
    apparent track).

    Let:

    $\psi_s=P(\mbox{site } s \mbox{ is occupied})$

    $\theta_s=P(\mbox{a track is laid down in site } s \mbox{ during the course
    of a day} | \mbox{site } s \mbox{ is occupied})$

    $p_k=P(\mbox{Observer } k \mbox{ detects a track} | \mbox{a track is
    present}).$

    $e_k=P(\mbox{Observer } k \mbox{ falsely records a track}| \mbox{there is
    not a track})$

    \textbf{[This is ugly :(.]}

    We assume the $\theta$'s are independent across days and constant across
    sampling instances, and that $p_k$ and $e_k$ are constant across sites and
    sampling instances. We define a site as being occupied if an otter uses the
    site for habitat during the winter.

    These assumptions together imply that the probability site $i$ contains a
    track, and this track is correctly detected by observer $l$ $k$ days after
    a snowfall event is given by: \[\psi_s [1-(1- \theta_s)^j] p_k\text{.}\]
    Here, $1-(1- \theta_s)^j$ is the probability that at least one track is
    laid down between the snowfall event and the survey date. This implies that
    the probability of a track being laid down on a given day increases with
    time since snowfall events.  There are five other possible outcomes, three
    of which lead to a 0 (recorded absence) and two more of which lead to a 1
    (recorded presence); all 6 outcomes are depicted in Figure 1. [Note:  I
    will add a figure with the hierarchical process and the probabilities of
    each event at the bottom of the figure].

    The four parameters described above: (i) occupancy rate $\psi$, (ii) a
    track-laying parameter $\theta$, (iii) detection probability $p$, and the
    false positive rate $e$, are the central parameters of our model.

    \textbf{[Should we include the paragraph below?]}

    Our data indicate a strong potential for false positives, consistent with
    other otter studies \citep{Evans2009}. The false positive parameter $e$ is
    the probability that a track is ``observed'' given that no track exists in
    the site on the given day. Thus false positives can occur in unoccupied
    sites or in an occupied site without a track. Both the correct detection
    ($p$) and false detection ($e$) probabilities depend on the observer, so
    that $p=p_l$ and $e=e_l$ for some observer \(l\).

    We omit the details but the likelihood function constructed from the above
    processes is described as follows. Let $S$  be the number of snow events in
    the season. For each snow event $i$, let $N_i$ be the number of sites
    sampled and $D_i$ be the number of days that were sampled from that snow
    event, which is typically no greater than three. And for each snow event
    $i$ and sampled snow day $j$, let $O_{ij}$ be the number of observers who
    surveyed the river, which is also typically no greater than three.  We
    denote the data by the vector ${\bf X}$. If we assume that occupancy and
    track-laying probabilities are independent and constant across sites, then
    the model likelihood is:

    \textbf{[No more than 2 days were sampled for any snowfall event
    on MS in 2003, but the flights ranged from 1 to 3 days after the actual
    event, so sometimes only days 2 and 3. I think for both days and observers
    it might be a safer bet to use sets instead of these integers.]}

    \begin{eqnarray*}
        L(\psi, \theta, \mathbf{p}, \mathbf{e}|\mathbf{X})& =&
        \prod_{i=1}^S \prod_{j=1}^{D_i} \prod_{k=1}^{O_{ij}} \left[ \left( \psi
        [(1-(1-\theta)^j) p_k +(1-\theta)^j) e_k ] + (1-\psi)e_k
        \right)^{n_{ijk}} \right. \\
        && \times \left( \psi[(1-(1-\theta)^j)(1-p_k)+(1-\theta)^j(1-e_k)]
        \right. \\
        && \left. \left. +(1-\psi)(1-e_k) \right)^{N_i-n_{ijk}}\right],
    \end{eqnarray*}
    where $n_{ijk}$ is the number of sites that record a 1 for the survey that
    was conducted $j$ days after snow event $i$ by observer $k$.

    \textbf{[The first line of the equation above crosses the margin.]}

    The construction of the above likelihood, under a maximum-likelihood
    framework, assumes a small sample of sites from a theoretically infinite
    sample space, independence across sites and independence between observers.
    While the latter is reasonable, the first two assumptions are almost surely
    not. In order to make best use of flight time, the entire section of the
    river was sampled. In addition, otters tend to travel for long distances
    [?] in the snow and a track at site $i$ is positively correlated with the
    existence of a track at site $i+1$. We quantified this association with
    standard runs tests for independence in a sequence of Bernoulli trials and
    found that for the 402-meter designs the null hypothesis of independence
    was strongly rejected in most surveys.  Correlations decreased, as would be
    expected, with larger plot sizes, but the positive correlation was often
    still statistically significant.  We address the spatial correlation issue
    below.

    \textbf{[The paragraph below is also mentioned in the introduction.]}

    The hierarchical model is more easily implemented via Markov Chain Monte
    Carlo (MCMC) using R, WinBUGS, and BRugs (see Appendix). The MCMC approach
    generates a Markov Chain to approximate the otherwise intractable posterior
    distribution.  The base model converges within about ten minutes of
    computing time on a standard desktop PC and 5,000 iterations of the MCMC
    chain.  WinBUGS is popular free and open-source software for MCMC. BRugs is
    an open-source package for R that contains the OpenBUGS library.  This
    allows users most of WinBUGS' functionality from within R.  For background
    on Bayesian hierarchical modeling and MCMC see \citet{Royle2008} and
    \citet{Ntzoufras2009}.

    We first constructed a simpler model that assumes independence across sites
    (but not among observers. We used a Uniform$(0,1)$ prior distribution for
    \(\psi\) and \(\theta\).  The \(p_i\)'s are from a beta\((a, b)\)
    distribution and the \(e_i\)'s are from beta\((c, b)\) distribution, where
    \(a\), \(b\) and \(c\) each come from gamma\((1, 0.1)\) distributions.
    After running the iterations of the MCMC chain, the model returns
    approximations of the posterior distribution for each variable.

    \citet{Royle2006} point out a problem with the ``dual role'' of $p$ and
    $e$ that occurs when false positives are added to a likelihood-based model.
    This results from the lack of contextual information.  In particular, for
    our model if $\theta =1$ then the likelihood above has the symmetrical
    property that $L(\psi, \theta, p, e) = L(1-\psi, \theta, e, p)$. This gives
    rise to bimodal posterior distributions. The suggested solution is to
    restrict the parameter space so that $p >> e$.

    In our Bayesian implementation we restricted the support for $p$ and $e$ to
    $(.6, 1)$ and $(0, .3)$, respectively, and this worked nicely. Note that
    for general $\theta$ the duality in the likelihood is no longer as simple
    as described above. But the problem with unspecified parameters and
    bimodality persists.

    Finally, we needed to consider the spatial dependence the data displayed.
    To do this, we considered conditional autoregressive (CAR) models that
    explicitly take into account such correlation. We considered one such model
    with a CAR prior placed only on occupancy, $\psi$, and another model with
    CAR priors placed on both $\psi$ and the track-laying process, $\theta$.
    The conditional variance of the CAR distributions, $\tau^2$, had a
    gamma$(0.5,0.0005)$ prior.

    \subsection{Simulation work}

    To test the ability of the model to reliably and accurately converge,
    simulated data sets with and without spatial correlation were generated and
    analyzed by the model. Three types of data were generated: data
    uncorrelated in \(\psi\) and \(\theta\), data with spatial correlation in
    \(\psi\) and uncorrelated in \(\theta\), and data with spatial correlation
    in both \(\psi\) and \(\theta\).

    Simulated data without spatial correlation is generated with given values
    of $\psi$, $\theta$, and $p$ and $e$ for each observer.

    To determine the efficacy of the model over the range of possible \(\psi\)
    and \(\theta\) values, simulated datasets with different \(\psi\) and
    \(\theta\) values were run through the model. All other variables: numbers
    of plots (100), observers (3), snow events (5), and observations after each
    snow event (3), were fixed. The model's estimates were then compared to the
    parameters used to generate the simulated data.

    As discussed, it is reasonable to believe that if an otter occupies one
    site, the neighboring sites have a higher probability of occupancy because
    that same otter may occupy consecutive sites.  Such a correlation in the
    occupancy process is incorporated into the simulated data through an ARIMA
    process.  First, a latent autoregressive process, \(\epsilon_i,\)
    \(i=1,...,n,\) is generated with a given unconditional variance of
    \(\sigma^2\) and a given autocorrelation parameter \(a\), using the
    arima.sim function in R.  Then occupancy probabilities for each site are
    generated using [notation? bo?] \(\psi_i=\mbox{logit}(\epsilon_i+bo)\),
    \(bo\) is a parameter representing the mean of the logit-transformed
    \(\psi\)'s.  This ensures that the \(\psi\)'s are between \(0\) and \(1\)
    and further allows every site \(i\) to have a different \(\psi_i\) that is
    correlated with \(\psi_{i-1}\).  We can then approximate the overall
    occupancy rate using integration to calculate \(E[\psi]\).

    Simulated binary track presence data with correlation only in \(\psi\) were
    generated using the above ARIMA machinery. The various models' estimates
    were then compared to the paramaters used to generate the simulated data,
    and confidence intervals were examined.

    Because a single track can extend over multiple sites, we also simulated
    data that displayed correlation in both the track-laying and occupancy
    processes.  Correlation in the track-laying process is incorporated into
    the simulated data through the addition of the variable \(\alpha\) and the
    redefinition of \(\theta\).  In this simulation method, \(\theta\)
    represents the probability that an otter lays a track in plot \(i\), given
    that plot \(i\) is occupied and plot \(i-1\) does \textit{not} contain a
    track. If plot \(i\) is occupied and plot \(i-1\) \textit{does} contain a
    track, then the probability of an otter making a track in plot \(i\) is
    \(\theta+\alpha\). Note that the correlation in occupancy is still
    incorporated through the ARIMA process.

    \subsection{Analysis of data}

    \textbf{[Should we bother with this paragraph? It's not really correct
    anymore anyway.  Chrisna: For the actual data, it is correct.  I'm not sure
    we ever ran anything new on the actual data, and I'm not even sure if it's
    possible at the moment because I may not have finished that part of the
    programming.]}

    The model analyzed data for each plot size of each river. For all data, the
    model was run with a burn-in period of ten thousand iterations.  Then
    parameters were tracked for ten thousand additional iterations.  If the
    model converged on the first run of the data, no further analysis was done.
    If the model did not converge on the first run of the data, it was run a
    second time with a burn-in period of twenty thousand iterations and a
    tracking period of twenty thousand iterations.  No further analysis was
    done.  The parameters estimated were \(E\) and \(p\) for each observer,
    \(\psi\), and \(\theta\).

    \subsection{Simulation work to look at study design}

    \textbf{[Now aren't we more interested in confidence intervals and less
    interested in convergence? We will have to redo all of this work.  Chrisna:
    Yes, we are.  I think this is pretty obvious though -- the more data points
    you have, the smaller the confidence intervals will be.]}

    We used simulations of data to determine the most efficient sampling
    practices (fewest number of flights) that produce data that converge with
    accurate estimates in the model.  The model is run with a burn-in period of
    10,000 iterations followed by a tracking period of 10,000 iterations.  The
    values used for the parameters in the simulated data are similar to those
    found in the results of the actual data obtained by running the model.  All
    of the data that was simulated had spatial correlation in both \(\psi\) and
    \(\theta\).  In each simulation there were one hundred plots, \(\psi\) was
    .3 with an \(\alpha\) of .5, \(\theta\) was .4 with a \(\beta\) of .4, all
    \(p\)'s were .7, and all \(E\)'s were .05.

    The number of snow events, days flown after a snow event, and number of
    observers are varied in order to determine the best combination of these
    parameters that will result in the fewest flights and a successful run of
    the model.  Data collected using one, two, and three observers were
    simulated.  For each of these numbers of observers, data with one, two, and
    three snow events were simulated.  For each of these numbers of observers
    combined with each of these numbers of snow events, data with one, two, and
    three consecutive days since snow were simulated.  Most of these
    combinations returned accurate estimates and good convergence when run in
    the model.  For the combinations that did not, more data was simulated with
    the number of snow events increased to five and ten.  Simulating data with
    consecutive days since snow increased to five was also tried.

\section{Results}

    \subsection{Model efficacy}

    The efficacy of three models was tested with three types of simulated data.
    The three models are the basic hierarchical model, the basic model with a
    CAR prior distribution on \(\psi\), and the basic model with CAR prior
    distributions on both \(\psi\) and \(\theta\).  The three types of
    simulated data are data with fixed parameters \(\psi\) and \(\theta\), data
    with \(\psi\) generated by the previously described ARIMA process and
    \(\theta\) fixed, and data with both \(\psi\) and \(\theta\) generated by
    the ARIMA process.  The other parameters are held constant.  The number of
    sites is 100, the number of observers is three, the number of observations
    after each snow event is three, \(p = (0.7, 0.85, 0.9)\) and \(E = (0.05,
    0.10, 0.20)\).

    We state that the model converges well if the MCMC diagnostics show
    convergence of two simulated chains, the final posterior distributions are
    unimodal, and the simulation parameter values fall within two SE's of the
    model's estimates.  For the most part, \(SE[E] \approx 0.01\) and \(SE(p)
    \approx 0.02\).

    \textbf{[I have no idea how to put the simulation results into words.]}

    Regardless of the type of data given to each of the three models, they
    perform similarly.  The biggest variation between the three models concerns
    the amount of iterations required for the model to converge.  When given
    uncorrelated data, all three models take approximately the same amount of
    iterations to converge.  The number of iterations depends upon the values
    of \(\psi\) and \(\theta\).  As \(\psi\) approaches \(\theta\), the amount
    of time required for the model to converge tends to increase.  When given
    the other two types of data, with correlation in \(\psi\) and correlation
    in \(psi\) and \(\theta\) the two models with CAR prior distributions
    occasionally require additional burn-in to converge.

    \subsection{Analysis of data}

    % Mississippi River!
    \begin{table}
        \caption{Mississippi River. The data converged well at the 402m and
        804m plot sizes. At the 1608m plot size, \(E\), \(\psi\), and
        \(\theta\) did not converge well.}
        \label{Mississippi}
        \begin{center}
            \begin{tabular}{|l|l|l|l|}
                \hline
                \multicolumn{4}{|l|}{\textbf{Mississippi River}} \\
                \hline
                    & 402m & 804m & 1608m \\
                \hline
                \multirow{2}{*}{E[Danny]}
                    & 0.01954 & 0.04808 & 0.06885 \\
                    & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778,
                    0.1383) \\
                \hline
                \multirow{2}{*}{E[John]}
                    & 0.05216 & 0.07771 & 0.13650 \\
                    & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449,
                    0.2330) \\
                \hline
                \multirow{2}{*}{E[Tom]}
                    & 0.02668 & 0.06140 & 0.10230 \\
                    & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287,
                    0.1796) \\
                \hline
                \multirow{2}{*}{p[Danny]}
                    & 0.65240 & 0.67340 & 0.74960 \\
                    & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510,
                    0.8910) \\
                \hline
                \multirow{2}{*}{p[John]}
                    & 0.67770 & 0.77580 & 0.82540 \\
                    & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390,
                    0.9326) \\
                \hline
                \multirow{2}{*}{p[Tom]}
                    & 0.53670 & 0.54220 & 0.65330 \\
                    & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530,
                    0.8104) \\
                \hline
                \multirow{2}{*}{\(\psi\)}
                    & 0.59380 & 0.69370 & 0.75820 \\
                    & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110,
                    0.9300) \\
                \hline
                \multirow{2}{*}{\(\theta\)}
                    & 0.15280 & 0.20730 & 0.27690 \\
                    & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030,
                    0.3716) \\
                \hline
            \end{tabular}
        \end{center}
    \end{table}

    % Whitewater River!
    \begin{table}
        \caption{Whitewater River. While the 402m data converged well, \(E\),
        \(\psi\), and \(\theta\) did not converge well for the 804m data.}
        \label{Whitewater}
        \begin{center}
            \begin{tabular}{|l|l|l|}
                \hline
                \multicolumn{3}{|l|}{\textbf{Whitewater River, 2003}} \\
                \hline
                    & 402m & 804m \\
                \hline
                \multirow{2}{*}{E[Danny]}
                    & 0.05608 & 0.0555 \\
                    & (0.02474, 0.09554) & (0.002859, 0.1390) \\
                \hline
                \multirow{2}{*}{E[John]}
                    & 0.16570 & 0.1931 \\
                    & (0.10130, 0.23510) & (0.065910, 0.3384) \\
                \hline
                \multirow{2}{*}{E[Tom]}
                    & 0.15350 & 0.2241 \\
                    & (0.09758, 0.21660) & (0.110000, 0.3623) \\
                \hline
                \multirow{2}{*}{p[Danny]}
                    & 0.66990 & 0.6985 \\
                    & (0.56080, 0.79040) & (0.573100, 0.8366) \\
                \hline
                \multirow{2}{*}{p[John]}
                    & 0.92040 & 0.9877 \\
                    & (0.84270, 0.97590) & (0.945100, 1.0000) \\
                \hline
                \multirow{2}{*}{p[Tom]}
                    & 0.75910 & 0.8575 \\
                    & (0.65060, 0.85800) & (0.745400, 0.9600) \\
                \hline
                \multirow{2}{*}{\(\psi\)}
                    & 0.87590 & 0.9126 \\
                    & (0.72910, 0.98880) & (0.746900, 0.9955) \\
                \hline
                \multirow{2}{*}{\(\theta\)}
                    & 0.24100 & 0.3779 \\
                    & (0.17890, 0.30900) & (0.286500, 0.4721) \\
                \hline
            \end{tabular}
        \end{center}
    \end{table}

    % Zumbro River!
    \begin{table}
        \caption{Zumbro River. The 402m data converged well. While
        \(\psi\) and \(\theta\) converged well for the 804m data, none of the
        other parameters did.}
        \label{Zumbro}
        \begin{center}
            \begin{tabular}{|l|l|l|}
                \hline
                \multicolumn{3}{|l|}{\textbf{Zumbro River, 2003}} \\
                \hline
                    & 402m & 804m \\
                \hline
                \multirow{2}{*}{E[Danny]}
                    & 0.02922 & 0.007066 \\
                    & (0.006786, 0.05994) & (0.0000, 0.03403) \\
                \hline
                \multirow{2}{*}{E[John]}
                    & 0.03388 & 0.008312 \\
                    & (0.008584, 0.06880) & (0.0000, 0.04188) \\
                \hline
                \multirow{2}{*}{E[Tom]}
                    & 0.02228 & 0.012960 \\
                    & (0.002430, 0.05385) & (0.0000, 0.05642) \\
                \hline
                \multirow{2}{*}{p[Danny]}
                    & 0.81890 & 0.837100 \\
                    & (0.749500, 0.88540) & (0.7615, 0.90700) \\
                \hline
                \multirow{2}{*}{p[John]}
                    & 0.81770 & 0.852500 \\
                    & (0.740100, 0.88670) & (0.7597, 0.92230) \\
                \hline
                \multirow{2}{*}{p[Tom]}
                    & 0.73120 & 0.805400 \\
                    & (0.625400, 0.82440) & (0.6885, 0.89780) \\
                \hline
                \multirow{2}{*}{\(\psi\)}
                    & 0.66550 & 0.651600 \\
                    & (0.549100, 0.78150) & (0.5098, 0.78210) \\
                \hline
                \multirow{2}{*}{\(\theta\)}
                    & 0.29950 & 0.395800 \\
                    & (0.236400, 0.36550) & (0.3100, 0.48520) \\
                \hline
            \end{tabular}
        \end{center}
    \end{table}

    Tables \ref{Mississippi}, \ref{Whitewater}, and \ref{Zumbro} contain our
    estimates for all of the data.

    With respect to the model's ability to converge on the actual data, the
    most important factors are plot size, the number of flights recorded, and
    the number of estimated parameters. In general, both plot size and the
    number of flights recorded must be sufficiently small and large,
    respectively, for the standard model with \(E\) to converge.

    The presence of numerous observations that occur in the second and third
    days since the most recent snow event allows one of the more crucial
    assumptions of our hierarchical model -- that otter tracks laid on each day
    after a snow event are visible throughout the entire observation period and
    therefore, that it is more likely for an observer to detect a track on
    later days -- to be thoroughly tested. Most of the \(\psi\) and \(\theta\)
    estimates of the data fall into ranges where our model is at its best, both
    with respect to convergence and to accurately estimating \(\psi\).

    As plot size increases, every estimated parameter tends to increase as
    well.  Although the vast majority of these increases, even at the 1608m
    plot size, lie well within the 95\% confidence interval of the accompanying
    estimate at the 402m plot size, this trend is intuitive. Once 402m plots
    are aggregated into 1608m plots, only one of the four 402m plots needs to
    be assigned a 1 in order for the entire 1608m plot to receive a 1. The
    amount of correlation between 1s determines the increase in the proportion
    of 1s to the number of plots -- as the amount of correlation increases,
    there will be more aggregated strings of 1s, so the increase in the
    proportion decreases.

    \subsection{Study design}

    Data with higher numbers of snow events and consecutive days since snow was
    not simulated because the model ran successfully on many combinations of
    one, two, and three observers, snow events, and consecutive days since
    snow.  Increasing the number of snow events or consecutive days since snow
    would also increase the total number of flights and therefore would not be
    more efficient than the combinations that were found to run successfully on
    the model.  Similarly, data with more than three observers was not
    simulated because it would also increase the number of flights, and there
    was success in estimating the parameters with only one, two, and three
    observers.

    The number of observers appears to be the least important of the three
    sampling variables.  Accurate estimates were obtained for all number of
    observers, as long as there were sufficient numbers of snow events and
    observations after each snow event.  For one observer, at least three snow
    events and three observations on consecutive days after each snow event are
    required for the model to converge and give accurate estimates.  For two or
    three observers, two events with three observations on consecutive days
    after each snow event is sufficient.

    The number of snow events appears to be the second-most important of the
    three sampling variables.  Although two snow events are required, only with
    one observer does the addition of a third snow event noticeably improve the
    model's estimates.

    The number of observations after each snow event appears to be the most
    important of the three sampling variables.  At least two observations after
    each snow event on consecutive days are required.  Additional observations
    after each event help the model converge more quickly, but the estimates
    become only slightly more accurate.

    Therefore, the most efficient combination of these sampling practices (in
    number of flights)  was found to be one observer, three snow events, and
    three consecutive days after each snow event.

\section{Discussion}

Our model performed surprisingly well, particularly given the complex and
sparse nature of the data analyzed. We were successfully able to incorporate
false detection, spatial data correlation, and the dependence of detection
probabilities on time since snowfall events into a model and still obtain both
accurate and precise results. While others have had success incorporating one
of these processes into statistical models \citep{Magoun2007, Royle2006,
Stanley2005}, we are not aware of any studies that have incorporated all three.
[Talk more about specifics when we get them.] Having the ability to model such
complex processes will make data collection in similar presence/absence surveys
both easier and less costly.

Interestingly, in our simulations with spatially correlated data, neither of
our two spatially explicit models performed particularly better than the
simpler spatially inexplicit model. While confidence intervals were sometimes
slightly smaller for the conditional autoregressive (CAR) models that took into
account the spatial correlation, the differences were small, and due to the
significantly longer computing times necessitated by the added complexity of
the spatial models, the simpler model may be a better choice. This is contrary
to what has been found in other studies that attempting to use CAR models with
similar binary data \citep{Wintle2006}.  However, the level of correlation in
their data may have been higher.

In addition, because funds for aerial surveys such as the ones used to collect
data in this study are limited, we explored different survey designs to
determine which aspects are most important in yielding precise and accurate
results. In particular, we examined numbers of observers, snow events, days
surveyed following snow events, and sites. [Make some sort of grand sweeping
conclusions about which are most important?] No single parameter appears most
important. However, we recommend a minimum sampling effort of [what?] to ensure
that the standard error for psi falls below [what?]. It is notable that if
there is more data available, the model will generally take less time to
converge.

These statistical methods could be used on data for other species, as well.
[John said we should talk about this. Which species?]

\section{Acknowledgements}

We would like to thank...

% Bibliography!
\bibliographystyle{Style}
\bibliography{Bibliography}

\section{Missing References}
\noindent

\textbf{[These are the ones that didn't show up in the paper.]}

\leftskip 0.15in

\parindent -0.15in

Gorman, T., Erb, J., McMillan, B.\ \& Martin, D.\ (2006) Space Use and
Sociality of River Otters ({\em Lontra Canadensis}) in Minnesota. {\em Journal
of Mammalogy}, {\bf 87}, 740--747.

\parindent -0.15in
Mackenzie, D.\ \& Royle, J.\ (2005) Designing Occupancy Studies: General Advice
and Allocating Survey Effort, {\em Journal of Applied Ecology}, {\bf 42},
1105-1114.

\parindent -0.15in
Sargeant, G., Sovada, M., Slivinski, C.\ \& Johnson, D.\ (2005) Markov Chain
Monte Carlo Estimation of Species Distributions: A Case Study of the Swift Fox
in Western Kansas, {\em Journal of Wildlife Management}, {\bf 69}, 483--497.

\end{document}

### 2010-06-08 NewPaper.tex
Added the real discussion.

@@
 probabilities on time since snowfall events into a model and still obtain both
-accurate and precise results. While others have had success incorporating one
-of these processes into statistical models \citep{Magoun2007, Royle2006,
-Stanley2005}, we are not aware of any studies that have incorporated all three.
-[Talk more about specifics when we get them.] Having the ability to model such
-complex processes will make data collection in similar presence/absence surveys
-both easier and less costly.
+accurate and precise estimates for the occupancy rate. While others have had
+success incorporating one of these processes into occupancy models
+\citep{Magoun2007, Royle2006, Stanley2005}, we are not aware of any studies
+that have incorporated all three. [Talk more about specifics when we get them.]
+Having the ability to model such complex processes will make data collection in
+similar presence/absence surveys both easier and less costly.
 
@@
 simpler spatially inexplicit model. While confidence intervals were sometimes
-slightly smaller for the conditional autoregressive (CAR) models that took into
-account the spatial correlation, the differences were small, and due to the
+slightly smaller for the conditional autoregressive (CAR) models that took the
+spatial correlation into account, the differences were small, and due to the
 significantly longer computing times necessitated by the added complexity of
 the spatial models, the simpler model may be a better choice. This is contrary
-to what has been found in other studies that attempting to use CAR models with
+to what has been found in other studies that attempted to use CAR models with
 similar binary data \citep{Wintle2006}. However, the level of correlation in
@@
 
-In addition, because funds for aerial surveys such as the ones used to collect
-data in this study are limited, we explored different survey designs to
-determine which aspects are most important in yielding precise and accurate
-results. In particular, we examined numbers of observers, snow events, days
-surveyed following snow events, and sites. [Make some sort of grand sweeping
-conclusions about which are most important?] No single parameter appears most
-important. However, we recommend a minimum sampling effort of [what?] to ensure
-that the standard error for psi falls below [what?]. It is notable that if
-there is more data available, the model will generally take less time to
-converge.
+In addition, because funds for aerial surveys are limited, we explored
+different survey designs to determine which aspects are most important in
+yielding precise and accurate results. In particular, we examined numbers of
+observers, snow events, days surveyed following snow events, and sites. No
+single parameter appears most important. However, we recommend a minimum
+sampling effort of [what?] to ensure that the standard error for psi falls
+below [what?]. \citet{Mackenzie2005} found that when detection probabilities
+exceed 0.5, as they do in the case of our data, they recommended a minimum of
+three surveys. Because we separate the track-laying and detection processes
+into two separate parameters ($p$ and $\theta$, respectively), more than three
+surveys are necessary in studies like this one.
 
-These statistical methods could be used on data for other species, as well.
-[John said we should talk about this. Which species?]
+These statistical methods could be used on investigations of other species, as
+well. Similar aerial snow-track surveys have been used in population studies
+for other riparian species such as ptarmigan, snowshoe hares, and red foxes
+\citep{St-Georges1995}, and non-riparian species such as caribou
+\citep{Courtois2003} and wolverines \citep{Magoun2007}. Although there may not
+be a natural river corridor to fly along for non-riparian species, transects on
+which to fly could be drawn ad hoc. The would be worthwhile, as there is great
+potential for applications of complex occupancy models such as this one in
+studies of population dynamics for use in wildlife conservation and management.
 
@@
 \parindent -0.15in
-
 Gorman, T., Erb, J., McMillan, B.\ \& Martin, D.\ (2006) Space Use and
@@
 of Mammalogy}, {\bf 87}, 740--747.
-
-\parindent -0.15in
-Mackenzie, D.\ \& Royle, J.\ (2005) Designing Occupancy Studies: General Advice
-and Allocating Survey Effort, {\em Journal of Applied Ecology}, {\bf 42},
-1105-1114.
 
